# Investigating differences in OGU abundances across Baseline samples using Zero-inflated Regression Models

```{r setup}
#| echo: false
#| eval: true
#| label: setup

# This initial code block loads custom utility functions and environment variables
# These are stored in separate files to keep the main analysis clean and organized
suppressWarnings(suppressMessages({
  source("utils/r-utils.r")
  source("scripts/OGeeU.r")
  source("scripts/zeroGUS.r")

  source("_environment.required")
  load("data/seqdata.RData")

  library(glmmTMB)
  library(DHARMa)
  library(broom)
  library(patchwork)
  library(emmeans)
  library(ggtext)

  library(promicrobialTools)

  dir.create(
    file.path(RESULTS, "Baseline-OGU-ZIBR"),
    showWarnings = FALSE,
    recursive = TRUE
  )
}))
```

## Prepare Outcome and Predictor Variables

```{r data-prep}
#| eval: true
#| label: data prep
#| results: asis
baseline <- phyloseq::subset_samples(seqdata_bac, timepoint == "1Baseline")

data <- prepare_ogu_data(
  baseline,
  filter_prevalence = TRUE,
  prevalence_cutoff = 0.4,
  filter_abundance = FALSE,
  rownames_as_samples = TRUE
)

data$taxa_data <- data$taxa_data %>%
  mutate(Phylum = ifelse(str_detect(Phylum, "Bacillota"), "Bacillota", Phylum))

outcome <- data$ogu_data
rownames(outcome) <- str_replace(
  str_replace(str_remove(rownames(outcome), "_G"), "_B_B", "_B"),
  "_B_E",
  "_E"
)

predictors <- prepare_predictors(data$meta_data)
predictors <- predictors %>%
  select(subject, trt, age_B) %>%
  rownames_to_column(var = "sample") %>%
  mutate(
    baseline_sample = str_replace(sample, "_G_E|_B_E|_E|_G_B|_B_B", "_B"),
    sample = str_remove(sample, "_G"),
  ) %>%
  mutate(sample = str_replace(sample, "_B_B", "_B")) %>%
  mutate(sample = str_replace(sample, "_B_E", "_E"))
```

## Testing different distributions

The following distribution families were used to fit a zero-inflated model:

- Poisson
- Simple negative binomial (linear parameterization) [@Hardin2007]
- Complex negative binomial (quadratic parameterization) [@Hardin2007]
- Gaussian
- Log normal

```{r fit}
#| eval: false
nb2_results <- genZI(
  outcome,
  predictors,
  formula = "age_B",
  zi = "~ age_B",
  family = "nbinom2"
)

nb1_results <- genZI(
  outcome,
  predictors,
  formula = "age_B",
  zi = "~ age_B",
  family = "nbinom1"
)

poi_results <- genZI(
  outcome,
  predictors,
  formula = "age_B",
  zi = "~ age_B",
  family = "poisson"
)

gau_results <- genZI(
  outcome,
  predictors,
  formula = "age_B",
  zi = "~ age_B",
  family = "gaussian"
)

lognorm_results <- genZI(
  outcome,
  predictors,
  formula = "age_B",
  zi = "~ age_B",
  family = "lognormal"
)
```

```{r save}
#| eval: false
#| echo: false
saveRDS(
  nb2_results,
  file.path(RESULTS, "Baseline-OGU-ZIBR", "genZI-OGU-nb2_results.rds")
)

saveRDS(
  nb1_results,
  file.path(RESULTS, "Baseline-OGU-ZIBR", "genZI-OGU-nb1_results.rds")
)

saveRDS(
  poi_results,
  file.path(RESULTS, "Baseline-OGU-ZIBR", "genZI-OGU-poi_results.rds")
)

saveRDS(
  gau_results,
  file.path(RESULTS, "Baseline-OGU-ZIBR", "genZI-OGU-gau_results.rds")
)

saveRDS(
  lognorm_results,
  file.path(RESULTS, "Baseline-OGU-ZIBR", "genZI-OGU-lognorm_results.rds")
)
```

```{r load}
#| eval: true
#| echo: false
nb2_results <- readRDS(file.path(
  RESULTS,
  "Baseline-OGU-ZIBR",
  "genZI-OGU-nb2_results.rds"
))

nb1_results <- readRDS(file.path(
  RESULTS,
  "Baseline-OGU-ZIBR",
  "genZI-OGU-nb1_results.rds"
))

poi_results <- readRDS(file.path(
  RESULTS,
  "Baseline-OGU-ZIBR",
  "genZI-OGU-poi_results.rds"
))

gau_results <- readRDS(file.path(
  RESULTS,
  "Baseline-OGU-ZIBR",
  "genZI-OGU-gau_results.rds"
))

lognorm_results <- readRDS(file.path(
  RESULTS,
  "Baseline-OGU-ZIBR",
  "genZI-OGU-lognorm_results.rds"
))
```

### Model diagnostic failures

```{r}
#| eval: true
#| label: list

list <- list(
  gau = gau_results,
  lognorm = lognorm_results,
  poi = poi_results,
  nb1 = nb1_results,
  nb2 = nb2_results
)
```

```{r}
#| eval: true
#| label: diagnose-mods
#| message: false
#| warning: false

poor_fits <- lapply(1:length(list), function(x) {
  d <- diagnoseZI(list[[x]]$all_models)
  d$dist <- names(list)[x]

  return(d)
})

poor_fits <- data.table::rbindlist(poor_fits)
```

### Comparing Model AIC and BIC values

```{r}
#| eval: true
#| label: summaries

summaries <- list(
  gau = gau_results$model_summary,
  lognorm = lognorm_results$model_summary,
  poi = poi_results$model_summary,
  nb1 = nb1_results$model_summary,
  nb2 = nb2_results$model_summary
)

summaries <- data.table::rbindlist(summaries, idcol = "dist")

summaries <- left_join(summaries, poor_fits, by = c("outcome_id", "dist"))
```

```{r}
#| eval: true
#| label: any-errors

if (any(summaries$error_message != "")) {
  print("Some models returned an error message")
}

```

```{r}
#| eval: true
#| label: diag-compare
summaries <- summaries %>%
  filter(pass == TRUE) %>%
  select(!c(error_message, message)) %>%
  mutate(index = 1:nrow(.)) %>% #make it easier to distinguish between outcome ids
  pivot_longer(
    cols = !c(index, outcome_id, dist, pass),
    names_to = "criterion",
    values_to = "value"
  )
```

```{r}
#| eval: true
#| label: best-AIC

aic <- summaries %>%
  filter(criterion == "AIC") %>%
  group_by(criterion, outcome_id) %>%
  arrange(value) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, minAIC = value, distAIC = dist)
```

```{r}
#| eval: true
#| label: best-BIC

bic <- summaries %>%
  filter(criterion == "BIC") %>%
  group_by(criterion, outcome_id) %>%
  arrange(value) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, minBIC = value, distBIC = dist)

```

```{r}
#| eval: true
#| label: tbl-best-fits
#| tbl-cap: Best fit models based on AIC and BIC

bestDist <- left_join(aic, bic) %>%
  mutate(dist = ifelse(distAIC == distBIC, distAIC, "Undetermined"))
```

### Checking for deviations in model assumptions

::: {.callout-note}

The larger the dispersion value, $\theta$ in `nbinom2` models, the less variance there is in the model [@Brooks2017].
:::

`dharmaDisp` represents the dispersion ratio, which should ideally be close to 1

```{r}
#| eval: true
#| label: tbl-best-disp
#| tbl-cap: lowest disp value for models that passed diagnostics

dispRatio <- summaries %>%
  filter(criterion == "dharmaDisp") %>%
  group_by(criterion, outcome_id) %>%
  mutate(skew = abs(value - 1)) %>%
  arrange(skew) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, bestDisp = value, distDisp = dist)
```

```{r}
#| eval: true
#| label: tbl-pvals
#| tbl-cap: NA

pvals <- left_join(
  bestDist,
  summaries,
  by = c("index", "outcome_id", "dist")
) %>%
  filter(str_detect(criterion, "Pvalue"), value <= 0.05) %>%
  arrange(index)
```

`r n_distinct(pvals$outcome_id)` models have a significant test result making the model fit problematic

### Final Model Distributions

```{r}
#| eval: true
#| label: best-dist-filtered
#| fig-cap: NA
#| tbl-cap: NA

bestDist_filtered <- bestDist %>%
  filter(!outcome_id %in% pvals$outcome_id)
```

## Fitting models to best distribution

#### Zero-inflated Negative Binomial 1 Models

```{r}
#| eval: true
#| label: nbinom1 models

nb1_models <- keep_at(
  nb1_results$all_models,
  filter(bestDist_filtered, dist == "nb1")$outcome_id
)

```

```{r}
#| eval: true
#| label: nb1-summary

if (length(nb1_models) > 0) {
  nb1_unfit <- diagnoseZI(nb1_models)
  nb1_summary <- data.frame(
    outcome_id = names(nb1_models),
    continuous_distribution = "nbinom1",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
  nb1_ogu_models <- nb1_results$adj_results %>% filter(outcome_id %in% names(nb1_models))
}
```

Are there any models that fail diagnostics: `r any(nb1_unfit$pass == FALSE)`

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for nbinom1
if (length(nb1_models) > 0) {
  nb1_dharma_results <- process_dharma_results(nb1_models)

  kable(nb1_dharma_results$summary)
}

```

#### Zero-inflated Negative Binomial Polynomial Models

```{r}
#| eval: true
#| label: nbinom2 models

nb2_models <- keep_at(
  nb2_results$all_models,
  filter(bestDist_filtered, dist == "nb2")$outcome_id
)

```


```{r}
#| eval: true
#| label: nb2-summary
if (length(nb2_models) > 0) {
  nb2_unfit <- diagnoseZI(nb2_models)

  nb2_summary <- data.frame(
    outcome_id = names(nb2_models),
    continuous_distribution = "nbinom2",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
nb2_ogu_models <- nb2_results$adj_results %>% filter(outcome_id %in% names(nb2_models))
}
```

Are there any models that fail diagnostics: `r any(nb2_unfit$pass == FALSE)`

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for nbinom1
if (length(nb2_models) > 0) {
  nb2_dharma_results <- process_dharma_results(nb2_models)

  kable(nb2_dharma_results$summary)
}
```

#### Zero-inflated Poisson Models

```{r}
#| eval: true
#| label: poi models

poi_models <- keep_at(
  poi_results$all_models,
  filter(bestDist_filtered, dist == "poi")$outcome_id
)

```


```{r}
#| eval: true
#| label: poi-summary
if (length(poi_models) > 0) {
  poi_unfit <- diagnoseZI(poi_models)

  poi_summary <- data.frame(
    outcome_id = names(poi_models),
    continuous_distribution = "poisson",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
poi_ogu_models <- poi_results$adj_results %>% filter(outcome_id %in% names(poi_models))
}
```

Are there any models that fail diagnostics: `r any(poi_unfit$pass == FALSE)`

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for Poisson models
if (length(poi_models) > 0) {
  poi_dharma_results <- process_dharma_results(poi_models)

  kable(poi_dharma_results$summary)
}
```

#### Zero-inflated Gaussian Models

```{r}
#| eval: true
#| label: gau unfit models
gau_models <- keep_at(
  gau_results$all_models,
  filter(bestDist_filtered, dist == "gau")$outcome_id
)

```


```{r}
#| eval: true
#| label: gau-summary
if (length(gau_models) > 0) {
  gau_unfit <- diagnoseZI(gau_models)

  gau_summary <- data.frame(
    outcome_id = names(gau_models),
    continuous_distribution = "gaussian",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
gau_ogu_models <- gau_results$adj_results %>% filter(outcome_id %in% names(gau_models))
}
```

Are there any models that fail diagnostics: `r any(gau_unfit$pass == FALSE)`

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for Poisson models
if (length(gau_models) > 0) {
  gau_dharma_results <- process_dharma_results(gau_models)

  kable(gau_dharma_results$summary)
}
```

#### Zero-inflated Lognormal Models

```{r}
#| eval: true
#| label: lognorm unfit models

lognorm_models <- keep_at(
  lognorm_results$all_models,
  filter(bestDist_filtered, dist == "lognorm")$outcome_id
)

```

```{r}
#| eval: true
#| label: lognorm-summary
if (length(lognorm_models) > 0) {
  lognorm_unfit <- diagnoseZI(gau_models)

  lognorm_summary <- data.frame(
    outcome_id = names(lognorm_models),
    continuous_distribution = "Log normal",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
lognorm_ogu_models <- lognorm_results$adj_results %>% filter(outcome_id %in% names(lognorm_models))
}
```

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for Poisson models
if (length(lognorm_models) > 0) {
  lognorm_dharma_results <- process_dharma_results(lognorm_models)

  kable(lognorm_dharma_results$summary)
}
```

```{r}
#| eval: false
#| label: save
#| echo: false

save(
  nb1_models,
  nb2_models,
  poi_models,
  gau_models,
  lognorm_models,
  file = file.path(RESULTS, "Baseline-OGU-ZIBR", "finalised-fits.RData")
)
```

```{r}
#| eval: true
#| label: load
#| echo: false

load(file.path(RESULTS, "Baseline-OGU-ZIBR", "finalised-fits.RData"))

```

## Significant OGUs

```{r}
#| eval: true
#| label: tbl-summary
#| tbl-cap: Summary table of OGUs successfully fit by a model, and their corresponding coefficient p-values

nb1_summary <- left_join(nb1_summary, nb1_ogu_models)
nb2_summary <- left_join(nb2_summary, nb2_ogu_models)
lognorm_summary <- left_join(lognorm_summary, lognorm_ogu_models)

kable(
  summary <- rbind(
    nb1_summary,
    nb2_summary,
    lognorm_summary
  )
)
```

```{r}
#| eval: true
#| label: save-tbl-summary
#| echo: false

write.csv(
  summary,
  file.path(
    RESULTS,
    "Baseline-OGU-ZIBR",
    "zero-inflated-ogu-model-summaries.csv"
  ),
  row.names = FALSE
)
```

Of the `r ncol(outcome)` OGUs that passed filtering (`r ncol(outcome)/ncol(data$ogu_data)`), `r nrow(summary)` were sucessfully fit to a zero-inflated model (`r nrow(summary)/ncol(outcome)*100`%)

```{r}
#| eval: true
#| label: zi-sig

ziSig <- identify_significant_ogus(
  summary,
  data$taxa_data
)

colnames(ziSig)[colnames(ziSig) == "rowname"] <- "outcome_id"
```

## Individual abundance changes

```{r}
#| eval: true
#| label: tmp
#| collapse: true
#| tbl-cap: Individual abundance changes for each OGU

tmp <- joined_data %>%
  dplyr::filter(outcome_id %in% ziSig$outcome_id) %>%
  group_by(sample, outcome_id) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = value,
    values_fill = 0
  ) %>%
  mutate(EvsB = Endpoint - Baseline) %>%
  mutate(
    difference = case_when(
      EvsB < 0 ~ "Decrease",
      EvsB > 0 ~ "Increase",
      EvsB == 0 ~ "None",
      TRUE ~ NA
    )
  )

tmp %>%
  group_by(outcome_id, difference) %>%
  summarise(total = n()) %>%
  kable()
```

### Average abundance changes

```{r}
#| eval: true
#| label: tbl-average-sig-changes

aveChange <- joined_data %>%
  dplyr::filter(outcome_id %in% ziSig$outcome_id) %>%
  group_by(outcome_id, timepoint) %>%
  summarise(average_count = mean(value)) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = average_count,
    values_fill = NA
  ) %>%
  mutate(EvsB = Endpoint - Baseline) %>%
  mutate(
    difference = case_when(
      EvsB < 0 ~ "Decrease",
      EvsB > 0 ~ "Increase",
      EvsB == 0 ~ "None",
      TRUE ~ NA
    )
  )

aveChange %>%
  group_by(difference) %>%
  summarise(total = n()) %>%
  kable()
```

### Demonstrating the importance of random effects

OGU G012221805 has an average abundance change of zero, so why is the EvsB term significant? Random effects!

```{r}
#| eval: true
#| label: fig-indv-changes
#| fig-cap: Changes in count abundance for OGU G012221805, showing the clear impact of individual random effects and the importance of including such in a model. Note the black dotted line would be the standard regression line indicating no change over time.

joined_data %>%
  filter(outcome_id == "G012221805") %>%
  group_by(subject) %>%
  ggplot(aes(x = time, y = value, )) +
  geom_point() +
  geom_line(
    aes(
      group = subject,
      col = as.factor(arm)
    ),
    linewidth = 1
  ) +
  geom_smooth(
    method = "lm",
    se = F,
    color = "black",
    formula = y ~ x,
    linetype = 'dotted',
    linewidth = 3
  ) +
  xlab("Time Since Feed Start (days)") +
  ylab("Count") +
  scale_color_manual(values = armColours) +
  labs(color = "Intervention Arm")

```

## Save Significant Model Results

```{r}
#| eval: true
ziSig <- left_join(ziSig, aveChange, by = "outcome_id")

write.csv(
  ziSig,
  file.path(
    RESULTS,
    "Baseline-OGU-ZIBR",
    "significant-ogus-zero-inflated-model.csv"
  ),
  row.names = FALSE
)
```

## Zero-inflated component

```{r}
#| eval: true
#| label: zi-comp
#| fig-cap: NA
#| tbl-cap: NA

zi <- lapply(
  list(nb1_models, nb2_models, poi_models, gau_models),
  function(l) {
    lapply(l, function(x) {
      summary(x)$coefficients$zi[, 4]
    })
  }
)

#FIX make naming automatic using indexing
zip <- bind_rows(
  lapply(zi, function(model_type) {
    bind_rows(model_type, .id = "outcome_id")
  }),
  .id = "model_type"
) %>%
  # Add better names for model types
  mutate(
    model_type = case_when(
      model_type == "1" ~ "nb1",
      model_type == "2" ~ "nb2",
      model_type == "3" ~ "poi",
      model_type == "4" ~ "gau",
      TRUE ~ model_type
    )
  )

if (any(zip <= 0.05)) {
  zip_sig <- identify_significant_ogus(zip, data$taxa_data)
}
```

## Using EMMEANS to determine significant changes in taxa abundances

```{r}
#| eval: true
#| label: fit-mods

fits <- list(nb1_models, nb2_models, lognorm_models)

fits <- list_c(fits)
```

### Conditional Model EMMs

#### OGUs Estimated marginal means

The below estimates the mean for each `arm` at each `timepoint`, controlling for age at baseline

```{r}
#| eval: true
#| label: Estimated marginal means for OGU abundances within each intervention arm

emm <- lapply(fits, function(x) {
  emmeans(
    x,
    ~ age_B,
    adjust = "sidak",
    weights = "proportional",
    type = "scale"
  )
})

```

##### OGU EMM differences between intervention arm at baseline and endpoint

```{r}
#| eval: true
#| label: tbl-emm-pairs
#| tbl-cap: Comparison of abundance marginal mean estimates (EMMs) betweeen intervention arm at baseline and endpoint. Only those with p $\leq$ 0.05 are show for brevity.

emm_pairs <- lapply(emm, pairs, reverse = TRUE)

emm_pairs_df <- emm_bind(emm_pairs, kable = F)

sigArms <- emm_pairs_df %>%
  filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
  mutate(
    change = ifelse(estimate < 0, "Decrease", "Increase"))
```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint

plotPhylaEMMs(sigArms, highlight_common = FALSE)
```

##### Effect Size

The default value used to standardise the effect size is the model's residual standard deviation (sigma; within-group variation). However, this may underestimate effect sizes in mixed models where combined variance may be more accurate. Here, the random effect variance is used to standardise effect sizes.

```{r}
#| eval: true
#| label: sigArm-eff
#| fig-cap: NA
#| tbl-cap: Effect sizes calculated as a standardized mean difference similar to Cohen's d.

emmEff <- lapply(1:length(emm_pairs), function(i) {
  eff_size(
    emm_pairs[[i]],
    sigma = attr(VarCorr(fits[[i]])$cond$subject, "stddev") * sqrt(2),
    edf = df.residual(fits[[i]]),
    method = "identity"
  )
})

names(emmEff) <- names(emm_pairs)

emmEff_df <- emm_bind(emmEff, kable = F)

emmEff_df <- emmEff_df %>%
  filter(Model %in% sigArms$Model) %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    ),
    lower.CL = ifelse(is.na(lower.CL), asymp.LCL, lower.CL),
    upper.CL = ifelse(is.na(upper.CL), asymp.UCL, upper.CL)
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, time, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

sigArmsEff <- left_join(sigArms, emmEff_df, by = c("Model", "time"))

```

```{r}
#| eval: true
#| label: fig-phyla-emms-effsize
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint. Significant differences with an effect size > 0.5 (medium effect size) are shown in the plot.

plotPhylaEMMs(sigArmsEff) +
  facet_wrap(~timepoint, scales = "free")
```

#### OGU EMM changes over time within each intervention arm

```{r}
#| eval: true
#| label: tbl-emm-arm-trend
#| tbl-cap: Rate of change per unit increase of abundance marginal mean estimates over time. The time.trend column represents the slope of the relationship with time.

emm_bind(
  trend <- lapply(fits, function(x) {
    emtrends(x, ~ age_B, var = "time")
  })
)
```

```{r}
#| eval: true
#| label: tbl-emm-arm-trend-compare
#| tbl-cap: Comparison of the rates of change from baseline to endpoint in abundance EMMs between intervention arms.

emm_bind(slopes <- lapply(trend, pairs, reverse = TRUE))

slopes_df <- emm_bind(slopes, kable = F)

sigArmsTime <- slopes_df %>%
  filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
  mutate(timepoint = ifelse(time == 1, "Endpoint", "Basepoint"))
```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint

plotPhylaEMMs(sigArmsTime, text_size = 8)
```

##### Effect Size

The default value used to standardise the effect size is the model's residual standard deviation (sigma; within-group variation). However, this may underestimate effect sizes in mixed models where combined variance may be more accurate. Here, the random effect variance is used to standardise effect sizes.

```{r}
#| eval: true
#| label: sigArm-eff
#| fig-cap: NA
#| tbl-cap: Effect sizes calculated as a standardized mean difference similar to Cohen's d.

emmEff <- lapply(1:length(slopes), function(i) {
  eff_size(
    slopes[[i]],
    sigma = attr(VarCorr(fits[[i]])$cond$subject, "stddev") * sqrt(2),
    edf = df.residual(fits[[i]]),
    method = "identity"
  )
})

names(emmEff) <- names(slopes)

emmEff_df <- emm_bind(emmEff, kable = F)

emmEff_df <- emmEff_df %>%
  filter(Model %in% sigArmsTime$Model) %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    ),
    lower.CL = ifelse(is.na(lower.CL), asymp.LCL, lower.CL),
    upper.CL = ifelse(is.na(upper.CL), asymp.UCL, upper.CL)
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, time, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

sigArmsTimeEff <- left_join(sigArmsTime, emmEff_df, by = c("Model", "time"))

```

```{r}
#| eval: true
#| label: fig-phyla-emms-effsize
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint. Significant differences with an effect size > 0.5 (medium effect size) are shown in the plot.

plotPhylaEMMs(filter(sigArmsTimeEff, effect.size >= 0.5))
```

### Zero-inflated Componet EMMs

```{r}
#| eval: true
#| label: tbl-emms
#| tbl-cap: Estimated marginal means for OGU abundances within each intervention arm

emm_zi <- lapply(fits, function(x) {
  # Get EMMs for arm at each time point
  emmeans(
    x,
    ~ age_B,
    
    adjust = "sidak",
    weights = "proportional",
    type = "scale",
    component = "zi"
  )
})

# Convert to probability scale
zi_probs <- lapply(emm_zi, regrid, transform = "logit")
```

##### OGU EMM differences between intervention arm at baseline and endpoint

```{r}
#| eval: true
#| label: tbl-emm-pairs
#| tbl-cap: Comparison of abundance marginal mean estimates (EMMs) betweeen intervention arm at baseline and endpoint

emm_bind(emm_zi_pairs <- lapply(emm_zi, pairs, reverse = TRUE))

emm_zi_pairs_df <- emm_bind(emm_zi_pairs, kable = F)

ZIsigArms <- emm_zi_pairs_df %>%
  filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
  mutate(
    change = ifelse(estimate < 0, "Decrease", "Increase"),
    timepoint = ifelse(time == 1, "Endpoint", "Basepoint")
  )
```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint

plotPhylaEMMs(ZIsigArms) +
  facet_wrap(~timepoint, scales = "free")
```

##### Effect Size

> [!quote]
> When a logistic regression is calculated, the regression coefficient (b1) is the estimated increase in the log odds of the outcome per unit increase in the value of the exposure. In other words, the exponential function of the regression coefficient (eb1) is the odds ratio associated with a one-unit increase in the exposure. [@Szumilas2010]

i.e. if $\beta_x = log(odds)$, then the odds ratio is $OR = \exp(\beta_x) = e^{\beta_x}$

```{r}
#| eval: true
#| label: odds-ratio
#| tbl-cap: Effect size calculated as odds-ratio for zero-inflation component
ZIodds <- lapply(fits, function(model) {
  exp(fixef(model)$zi)
})

kable(
  ZIodds <- do.call(rbind, ZIodds) %>%
    as.data.frame() %>%
    rownames_to_column("Model")
)
```

```{r}
#| eval: true
#| label: cohens-d
#| fig-cap: NA
#| tbl-cap: NA
ZIemmEff <- lapply(1:length(emm_zi_pairs), function(i) {
  eff_size(
    emm_zi_pairs[[i]],
    sigma = attr(VarCorr(fits[[i]])$cond$subject, "stddev") * sqrt(2),
    edf = df.residual(fits[[i]]),
    method = "identity"
  )
})

names(ZIemmEff) <- names(emm_zi_pairs)

ZIemmEff <- emm_bind(ZIemmEff, kable = F)

ZIemmEff <- ZIemmEff %>%
  filter(Model %in% sigArms$Model) %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    ),
    lower.CL = ifelse(is.na(lower.CL), asymp.LCL, lower.CL),
    upper.CL = ifelse(is.na(upper.CL), asymp.UCL, upper.CL)
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, time, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

ZIsigArmsEff <- left_join(ZIsigArms, ZIemmEff, by = c("Model", "time"))
kable(ZIsigArmsEff <- left_join(ZIsigArmsEff, ZIodds, by = c("Model")))
```

```{r}
#| eval: true
#| label: fig-phyla-emms-effsize
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint. Significant differences with an effect size > 0.5 (medium effect size) are shown in the plot.

plotPhylaEMMs(ZIsigArmsEff) +
  facet_wrap(~timepoint, scales = "free")
```

#### OGU EMM changes over time within each intervention arm

```{r}
#| eval: true
#| label: tbl-emm-arm-trend
#| tbl-cap: Rate of change per unit increase of abundance marginal mean estimates over time. The time.trend column represents the slope of the relationship with time.

emm_bind(
  trend_zi <- lapply(fits, function(x) {
    emtrends(x, ~ age_B, var = "time")
  })
)
```

```{r}
#| eval: true
#| label: tbl-emm-arm-trend-compare
#| tbl-cap: Comparison of the rates of change from baseline to endpoint in abundance EMMs between intervention arms.

emm_bind(slopes_zi <- lapply(trend_zi, pairs, reverse = TRUE))

slopes_zi_df <- emm_bind(slopes_zi, kable = F)

ZIsigArmsTime <- slopes_zi_df %>%
  filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
  mutate(timepoint = ifelse(time == 1, "Endpoint", "Basepoint"))
```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint

plotPhylaEMMs(ZIsigArmsTime)
```

##### Effect Size

The default value used to standardise the effect size is the model's residual standard deviation (sigma; within-group variation). However, this may underestimate effect sizes in mixed models where combined variance may be more accurate. Here, the random effect variance is used to standardise effect sizes.

```{r}
#| eval: true
#| label: sigArm-eff
#| fig-cap: NA
#| tbl-cap: Effect sizes calculated as a standardized mean difference similar to Cohen's d.

emmEff_zi <- lapply(1:length(slopes_zi), function(i) {
  eff_size(
    slopes[[i]],
    sigma = attr(VarCorr(fits[[i]])$zi$subject, "stddev") * sqrt(2),
    edf = df.residual(fits[[i]]),
    method = "identity"
  )
}))

names(emmEff_zi) <- names(slopes_zi)

emmEff_zi_df <- emm_bind(emmEff_zi, kable = F)

emmEff_zi_df <- emmEff_df_zi %>%
  filter(Model %in% ZIsigArmsTime$Model) %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    ),
    lower.CL = ifelse(is.na(lower.CL), asymp.LCL, lower.CL),
    upper.CL = ifelse(is.na(upper.CL), asymp.UCL, upper.CL)
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, time, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

ZIsigArmsTimeEff <- left_join(
  ZIsigArmsTime,
  emmEff_zi_df,
  by = c("Model", "time")
)

```

```{r}
#| eval: true
#| label: fig-phyla-emms-effsize
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint.

plotPhylaEMMs(filter(ZIsigArmsTimeEff))
```

## Effect of Stunting on OGU Abundances

```{r}
#| eval: false
#| label: stunt-model

y <- outcome %>% select(names(poi_models))

poi_stunt <- genZI(
  y,
  predictors,
  formula = "age_B",
  zi = "~ age_B",
  family = "poisson"
)

compare_nested_models(list(
  poi_models = keep_at(poi_results, names(poi_models)),
  poi_stunt = poi_stunt
))

```


## Note on applying Structured Covariance Matricies to Zero-inflated data

Structured covariance matricies are typically avoided when the data is sparse, zero-inflated or have few repeated measures available. See notes on "Approaches to Modelling Covariance".