## Investigating changes in OGU abundance over time using Zero-inflated Beta Regression Models with Random Effects

```{r}
#| echo: false
#| eval: true
#| label: setup

# This initial code block loads custom utility functions and environment variables
# These are stored in separate files to keep the main analysis clean and organized
suppressMessages({
  source("utils/r-utils.r")
  source("scripts/OGeeU.r")
  source("scripts/zeroGUS.r")

  source("_environment.required")
  load("data/seqdata.RData")

  library(glmmTMB)
  library(DHARMa)
  library(broom)
  library(patchwork)

  dir.create(file.path(RESULTS, "EvsB-OGU-ZIBR"), showWarnings = FALSE, recursive = TRUE)

})
```

```{r}
#| eval: true
#| label: data prep

data <- prepare_ogu_data(absAbund, filter_prevalence = TRUE, prevalence_cutoff = 0.4, filter_abundance = FALSE)

outcome <- data$ogu_data
rownames(outcome) <- str_replace(
  str_replace(str_remove(rownames(outcome), "_G"), "_B_B", "_B"),
  "_B_E",
  "_E"
)

predictors <- prepare_predictors(data$meta_data)
predictors <- predictors %>%
  mutate(arm = case_when(arm == "CPM" ~ 0, arm == "FeZnPM" ~ 1)) %>%
  rownames_to_column(var = "sample") %>%
  mutate(
    baseline_sample = str_replace(sample, "_G_E|_B_E|_E|_G_B|_B_B", "_B"),
    sample = str_remove(sample, "_G"),
  ) %>%
  mutate(sample = str_replace(sample, "_B_B", "_B")) %>%
  mutate(sample = str_replace(sample, "_B_E", "_E")) 

results <- genZI(outcome, predictors)

saveRDS(results, file.path(RESULTS, "EvsB-OGU-ZIBR", "genZI-OGU-results.rds"))
```

```{r}
#| eval: false
#| label: results

results <- readRDS(file.path(RESULTS, "EvsB-OGU-ZIBR", "genZI-OGU-results.rds"))
```

```{r}
#| eval: true
#| label: fig-scatter-converged-models

zero_plot(results)
```

```{r}
#| eval: true
#| label: fig-summary-plots

plots <- create_summary_plots(results)
plots$combined
```

```{r}
#| eval: true
#| label: residuals

suppressWarnings({
res <- lapply(results$all_models, function(mod){
  r <- simulateResiduals(mod)
  t1 <- tidy(testUniformity(r, plot = FALSE))
  t2 <- tidy(testDispersion(r, plot = FALSE))
  t3 <- tidy(testOutliers(r, plot = FALSE))
  t4 <- tidy(testZeroInflation(r, plot = FALSE))

  return(list(Uniformity = t1,Dispersion = t2, Outliers  = t3, ZeroInflation = t4))
})}
)

dharma_results <- process_dharma_results(results$all_models)
```


```{r}
#| eval: false
#| label: refits
unfit <- results$adj_results %>%
  filter(if_any(everything(), is.na)) 

unfit <- unfit[,1]

outcome <- data$ogu_data 

rownames(outcome) <- str_replace(
  str_replace(str_remove(rownames(outcome), "_G"), "_B_B", "_B"),
  "_B_E",
  "_E"
)

outcome <- outcome %>%
  select(all_of(unfit))

ncol(outcome) == length(unfit)

tmp <- genZI(outcome, predictors, family = poisson)
```

A total of `r length(unfit)` models did not converge appropriately and were refit with a Poisson distribution.

```{r}
#| eval: true
#| label: unfit2

unfit2 <- tmp$adj_results %>%
  filter(if_any(everything(), is.na))

unfit2 <- unfit2[,1]

outcome <- data$ogu_data 

rownames(outcome) <- str_replace(
  str_replace(str_remove(rownames(outcome), "_G"), "_B_B", "_B"),
  "_B_E",
  "_E"
)

outcome <- outcome %>%
  select(all_of(unfit2))

ncol(outcome) == length(unfit2)

tmp2 <- genZI(outcome, predictors, family = gaussian)

```

```{r}
#| eval: true
#| label: unfit3

unfit3 <- tmp$adj_results %>%
  filter(if_any(everything(), is.na))

unfit3 <- unfit3[,1]

outcome <- outcome %>%
  select(all_of(unfit3))

ncol(outcome) == length(unfit3)

tmp3 <- genZI(outcome, predictors, family = lognormal)

lapply(tmp3$all_models, diagnose)
```

A total of `r length(unfit2)` models did not converge appropriately and were refit with a Log normal distribution. A single OGU (G001054195) resulted in large coefficients in the zero-inflated model component. A better fit could not be found.

```{r}
#| eval: false
#| label: auto

results <- run_zibr_pipeline (
  phyloseq_obj = bac_base_end,
  output_dir = file.path(RESULTS, "EvsB-OGU-GEE")
)

# Create summary plots
plots <- create_summary_plots(results)
print(plots$p_value_dist)
print(plots$volcano_plot)

# 5. Examine a specific significant OGU
# Replace "OGU_ID" with an actual OGU ID from your significant results
ogu_details <- examine_ogu(results, "OGU_ID")

# 6. Access detailed results for further analysis
significant_ogus <- results$significant_ogus
model_summaries <- results$model_results$convergence_summary

# 7. Export results to Excel for easier viewing
if (require(openxlsx)) {
  write.xlsx(
    list(
      Convergence_Summary = results$model_results$convergence_summary,
      Significant_OGUs = results$significant_ogus
    ),
    file = file.path(output_dir, "ZIBR_Results.xlsx")
  )
}
```
