# Baseline to Endpoint Functional Analysis

```{r setup}
#| echo: false
#| eval: true
#| label: setup
#| cache: false

# This initial code block loads custom utility functions and environment variables
suppressWarnings(suppressMessages({
  source("utils/r-utils.r")
  source("_environment.required")
  load(normalizePath("data/seqdata.RData"))

  metadata <- read_delim(normalizePath(METADATA))

  library(edgeR)
  library(limma)
  library(ComplexHeatmap)
  library(pairwiseAdonis)
  library(patchwork)

  library(promicrobialTools)

  annoBac <- read.csv(file.path(
    REF,
    "gene-ontology/goslim_prokaryote-formatted.csv"
  ))

  prefix <- "GO-EvsB-"

  filepath <- get_results_dir(directory_naming = "both")
}))
```

Results will be saved to ``r filepath``.

## Gene Ontology Enrichment

### Load data

```{r}
#| eval: true
#| label: load data

go_be <- read.csv(file.path(
  RESULTS,
  "09-go-term-analysis/GO-EvsB-annotated-go-terms.csv"
))
```

### GO term presence

#### EvsB calculate percentage of samples with particular GO terms

```{r}
#| eval: true
#| label: EvsB calculate percentage of samples with particular GO terms

go_be <- go_be %>%
  filter(go_id %in% annoBac$go_id & ontology != "Cellular Component") %>%
  arrange(dplyr::desc(count)) %>%
  group_by(go_id) %>%
  mutate(n = 1) %>%
  mutate(
    percent_samples = sum(n) / length(unique(.$sample)) * 100
  )

tmp2 <- go_be %>%
  group_by(Terms, arm, timepoint, ontology) %>%
  summarise(term_presence = mean(percent_samples))
```

There are `r length(unique(go_be$Terms))` unique generic GO terms in the baseline-endpoint dataset.

```{r}
#| eval: true
#| label: number-go-terms
#| fig-cap: NA
#| tbl-cap: NA
go_be %>%
  group_by(arm, timepoint) %>%
  summarise(term_presence = n_distinct(go_id))
```

```{r}
#| eval: true
#| label: number-go-terms
#| fig-cap: NA
#| tbl-cap: NA
go_be %>%
  group_by(arm, timepoint, ontology) %>%
  summarise(term_presence = n_distinct(go_id))
```

```{r}
#| eval: true
#| label: double checking the change in average counts of particular GO terms

go_be %>%
  group_by(go_id, timepoint) %>%
  summarise(average_count = mean(count)) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = average_count,
    values_fill = 0
  ) %>%
  mutate(EvsB = Endpoint - Baseline) %>%
  mutate(
    difference = case_when(
      EvsB < 0 ~ "Decrease",
      EvsB > 0 ~ "Increase",
      EvsB == 0 ~ "None",
      TRUE ~ NA
    )
  ) %>%
  group_by(difference) %>%
  summarise(total = n())
```

###### EvsB plot of percent samples with particular generic GO terms

```{r}
#| eval: true
#| label: EvsB plot of percent samples with particular generic GO terms
onto_col <- assign_unique_colors(go_be$ontology)

go_be %>%
  dplyr::filter(ontology == "Biological process") %>%
  ggplot(aes(
    y = reorder(Terms, dplyr::desc(ontology)),
    x = percent_samples,
    fill = ontology
  )) +
  geom_col(position = "dodge") +
  xlab("Percentage of Samples") +
  ylab(NULL) +
  scale_fill_manual(values = onto_col) +
  theme(axis.text.y = element_text(size = 14, face = 'bold', color = 'black')) +
  facet_wrap(~ timepoint + arm)

go_be %>%
  dplyr::filter(ontology == "Molecular function") %>%
  ggplot(aes(
    y = reorder(Terms, dplyr::desc(ontology)),
    x = percent_samples,
    fill = ontology
  )) +
  geom_col(position = "dodge") +
  xlab("Percentage of Samples") +
  ylab(NULL) +
  scale_fill_manual(values = onto_col) +
  theme(axis.text.y = element_text(size = 14, face = 'bold', color = 'black')) +
  facet_wrap(~ timepoint + arm)

tmp2 %>%
  dplyr::filter(ontology != "Cellular component") %>%
  mutate(
    timepoint = case_when(
      str_detect(timepoint, "Baseline") ~ "Baseline",
      str_detect(timepoint, "Endpoint") ~ "Endpoint"
    )
  ) %>%
  ggplot(aes(
    y = reorder(Terms, dplyr::desc(ontology)),
    x = term_presence,
    fill = ontology
  )) +
  geom_col(position = "dodge") +
  xlab("Percentage of Samples") +
  ylab(NULL) +
  scale_fill_manual(values = onto_col) +
  theme(axis.text.y = element_text(size = 14, face = 'bold', color = 'black')) +
  ggh4x::facet_nested_wrap(~ timepoint + arm)
```

There are `r length(unique(go_be$go_id))` unique GO terms among baseline-endpoint samples

### Differential GO term analysis

$H_0$: GO functional terms are not differentially abundant at endpoint compared to baseline
$H_1$: There is a difference between GO functional terms pathway abundance at endpoint compared to baseline

$H_0$: GO functional terms are not differentially abundant between intervention arms at either timepoint
$H_1$: There is a difference between GO functional terms pathway abundance between intervention arm at either timepoints

#### Cross-sectional comparisons

```{r}
#| eval: true
base <- go_be %>% filter(timepoint == "Baseline")
end <- go_be %>% filter(timepoint == "Endpoint")
```

##### Baseline arm differences

###### Data preparation

```{r}
#| eval: true
#| label: testing for de abundant pathways
base_counts <- base %>%
  dplyr::select(go_id, sample, count) %>%
  pivot_wider(names_from = sample, values_from = count, values_fill = 0) %>%
  as.data.frame()

base_counts <- DGEList(as.data.frame(base_counts))

base_counts$samples$arm <- as.factor(go_be$arm[match(
  rownames(base_counts$samples),
  go_be$sample
)])

base_counts$samples$group <- NULL

base_counts$go_id <- go_be %>%
  select(go_id, ontology, Terms) %>%
  as.data.frame()

base_counts.keep <- rowSums(base_counts$counts == 0) == ncol(base_counts$counts)
base_counts <- base_counts[base_counts.keep == FALSE, ]

base_counts <- normLibSizes(base_counts)
```

###### Contrast design

:::h0

$H_0$: go_be pathways are not differentially abundant between arms at baseline

:::

::: h1

$H_1$: There is a difference between go_be pathway abundance between arms at baseline
    $H_1a$: go_be pathway abundance *increases* between arms at baseline
    $H_1b$: go_be pathway abundance *decreases* between arms at baseline

:::

```{r}
#| eval: true
#| label: go_be pathway differences between baseline and endpoint
#Linear Model
#Contrasts
design <- stats::model.matrix(~arm, data = base_counts$samples)

contrasts <- makeContrasts(
  CPMvsFeZnPM = armFeZnPM,
  levels = design
)
```

###### Empirical Bayes Statistics for Differential Expression 

```{r}
#| eval: true
#Transform raw count data to log2-base_counts per million (logCPM), estimate the mean-variance relationship and use this to compute appropriate observation-level weights. The data are then ready for linear modelling.

#If the DGEList object has been scale-normalized in edgeR (or before), then leave normalize.method="none" in voom, i.e., the logCPM values should not usually be re-normalized in the voom call.
#go_be abundance values have not been normalised
# use updated edgeR voom procedure
v <- voomLmFit(base_counts, design, normalize.method = "none")
#v <- voom(base_counts, design, normalize.method="none") #Transforms count data to log2-base_counts per million
#vfit <- lmFit(v, design)
vfit <- contrasts.fit(v, contrasts = contrasts)
efit <- eBayes(vfit) #estimation of QL dispersions
```

```{r}
#| eval: true
#| label: tbl-fdr5
#| tbl-cap: OGUs that are significantly differentially abundant from zero for each contrast from a fit object (eBayes) containing p-values and test statistics. An FDR of 0.05 was used.

knitr::kable(summary(
  FDR5 <- decideTests(efit, adjust.method = "BH", p.value = 0.05, fc = 0)
))

write.fit(
  fit = efit,
  results = FDR5,
  file = file.path(
    filepath,
    paste0(prefix, "limma-write-fit-baseline-results-FDR5-eBayes.csv")
  ),
  row.names = F
)
```

###### Ordination Analysis and PERMANOVA of differences between arms at baseline

```{r}
#| eval: true
#| label: tbl-mds
#| tbl-cap: Results of PERMANOVA test investigating abundance differences between arms at baseline in paired samples.
#unsupervised learning: MDS plot
group_colours <- base_counts$samples$arm
levels(group_colours) <- c("CPM" = "#FC8D62", "FeZnPM" = "#8DA0CB")
group_colours <- as.character(group_colours)
df <- as.data.frame(t(base_counts$counts))
df <- cbind(df, group = base_counts$samples$arm)

#PERMANOVA testing differenital abundances between timepoints
mds_stat <- pairwiseAdonis::pairwise.adonis2(
  df[, 1:(ncol(df) - 1)] ~ group,
  data = df,
  method = 'eu'
) #returns non-adjusted p-values

#output table
knitr::kable(mds_stat[2])
```

```{r}
#| eval: true
#| label: tbl-mds-sum
#| tbl-cap: Results of PERMANOVA test investigating abundance differences between arms at baseline in paired samples.

pv <- lapply(2:length(mds_stat), function(x) {
  w <- mds_stat[[x]][1, 4]
  y <- mds_stat[[x]][1, 5]
  z <- names(mds_stat)[x]
  data.frame(contrast = z, p = y, F.stat = w)
})
pv <- bind_rows(pv)
knitr::kable(
  pv %>%
    mutate(p.adj = round(p.adjust(p, method = "BH"), 4)) %>%
    mutate(
      sig = case_when(
        p.adj <= 0.0001 ~ "****",
        p.adj <= 0.001 ~ "***",
        p.adj <= 0.01 ~ "**",
        p.adj <= 0.05 ~ "*",
        p.adj > 0.05 ~ "ns"
      )
    )
)
```

The p-value ($Pr(>F)$) for `r base_counts$samples$arm` is `r pv$p`, which is greater than the conventional significance level of 0.05. This suggests that we cannot reject the null hypothesis, implying that there is no significant overall difference in microbial pathway abundance between baseline and endpoint.

The `plotMDS()` function accepts "any data object that can be coerced to a matrix of log-expression values, for example an ExpressionSet or an EList" [@Smyth2025]. If the `edgeR::voomLmFit()` function was used the `EList` object can be obtained using `fit$EList`

```{r}
#| eval: true
#| label: MDS plot comparing arms at baseline in paired samples.

xmax = max(
  plotMDS(
    v$EList,
    col = group_colours,
    dim.plot = c(1, 2),
    pch = 16,
    gene.selection = "common"
  )$x
) -
  0.05
ymax = max(
  plotMDS(
    v$EList,
    col = group_colours,
    dim.plot = c(1, 2),
    pch = 16,
    gene.selection = "common"
  )$y
) -
  0.05

par(mar = c(7.1, 5.1, 5.1, 5.1), xpd = TRUE)
plotMDS(
  v$EList,
  col = group_colours,
  dim.plot = c(1, 2),
  pch = 16,
  gene.selection = "common"
)
legend(
  "bottom",
  pch = 16,
  col = unique(group_colours),
  legend = levels(v$targets$arm),
  inset = c(0, -0.165),
  bty = "n",
  ncol = 2
)
text(
  x = xmax - 0.05,
  y = ymax,
  labels = bquote(R^2 == .(round(as.data.frame(mds_stat[2])[1, 3], 3)))
)
text(
  x = xmax - 0.05,
  y = ymax - 0.12,
  labels = bquote(p == .(round(as.data.frame(mds_stat[2])[1, 5], 3)))
)
```

There are two obvious clusters that are not due to arm in the MDS graph.

##### Endpoint arm differences

###### Data preparation

```{r}
#| eval: true
#| label: testing for de abundant pathways
end_counts <- end %>%
  dplyr::select(go_id, sample, count) %>%
  pivot_wider(names_from = sample, values_from = count, values_fill = 0) %>%
  as.data.frame()

end_counts <- DGEList(as.data.frame(end_counts))

end_counts$samples$arm <- as.factor(go_be$arm[match(
  rownames(end_counts$samples),
  go_be$sample
)])

end_counts$samples$group <- NULL

end_counts$go_id <- go_be %>%
  select(go_id, ontology, Terms) %>%
  as.data.frame()

end_counts.keep <- rowSums(end_counts$counts == 0) == ncol(end_counts$counts)
end_counts <- end_counts[end_counts.keep == FALSE, ]

end_counts <- normLibSizes(end_counts)
```

####### Contrast design

:::h0

$H_0$: go_be pathways are not differentially abundant between arms at endline

:::

::: h1

$H_1$: There is a difference between go_be pathway abundance between arms at endline
    $H_1a$: go_be pathway abundance *increases* between arms at endline
    $H_1b$: go_be pathway abundance *decreases* between arms at endline

:::

```{r}
#| eval: true
#| label: go_be pathway differences between endline and endpoint
#Linear Model
#Contrasts
design <- stats::model.matrix(~arm, data = end_counts$samples)

contrasts <- makeContrasts(
  CPMvsFeZnPM = armFeZnPM,
  levels = design
)
```

####### Empirical Bayes Statistics for Differential Expression 

```{r}
#| eval: true
#Transform raw count data to log2-end_counts per million (logCPM), estimate the mean-variance relationship and use this to compute appropriate observation-level weights. The data are then ready for linear modelling.

#If the DGEList object has been scale-normalized in edgeR (or before), then leave normalize.method="none" in voom, i.e., the logCPM values should not usually be re-normalized in the voom call.
#go_be abundance values have not been normalised
# use updated edgeR voom procedure
v <- voomLmFit(end_counts, design, normalize.method = "none")
#v <- voom(end_counts, design, normalize.method="none") #Transforms count data to log2-end_counts per million
#vfit <- lmFit(v, design)
vfit <- contrasts.fit(v, contrasts = contrasts)
efit <- eBayes(vfit) #estimation of QL dispersions
```

```{r}
#| eval: true
#| label: tbl-fdr5
#| tbl-cap: OGUs that are significantly differentially abundant from zero for each contrast from a fit object (eBayes) containing p-values and test statistics. An FDR of 0.05 was used.

knitr::kable(summary(
  FDR5 <- decideTests(efit, adjust.method = "BH", p.value = 0.05, fc = 0)
))

write.fit(
  fit = efit,
  results = FDR5,
  file = file.path(
    filepath,
    paste0(prefix, "limma-write-fit-endpoint-results-FDR5-eBayes.csv")
  ),
  row.names = F
)
```

####### Ordination Analysis and PERMANOVA of differences between arms at endline

```{r}
#| eval: true
#| label: tbl-mds
#| tbl-cap: Results of PERMANOVA test investigating abundance differences between arms at endline in paired samples.
#unsupervised learning: MDS plot
group_colours <- v$targets$arm
levels(group_colours) <- c("CPM" = "#FC8D62", "FeZnPM" = "#8DA0CB")
group_colours <- as.character(group_colours)
df <- as.data.frame(t(v$EList$E))
df$group <- v$targets$arm[match(rownames(df), rownames(v$targets))]

#PERMANOVA testing differenital abundances between arms
mds_stat <- pairwiseAdonis::pairwise.adonis2(
  df[, 1:(ncol(df) - 1)] ~ group,
  data = df,
  method = 'eu'
) #returns non-adjusted p-values

#output table
knitr::kable(mds_stat[2])
```

```{r}
#| eval: true
#| label: tbl-mds-sum
#| tbl-cap: Results of PERMANOVA test investigating abundance differences between arms at endline in paired samples.

pv <- lapply(2:length(mds_stat), function(x) {
  w <- mds_stat[[x]][1, 4]
  y <- mds_stat[[x]][1, 5]
  z <- names(mds_stat)[x]
  data.frame(contrast = z, p = y, F.stat = w)
})
pv <- bind_rows(pv)
knitr::kable(
  pv %>%
    mutate(p.adj = round(p.adjust(p, method = "BH"), 4)) %>%
    mutate(
      sig = case_when(
        p.adj <= 0.0001 ~ "****",
        p.adj <= 0.001 ~ "***",
        p.adj <= 0.01 ~ "**",
        p.adj <= 0.05 ~ "*",
        p.adj > 0.05 ~ "ns"
      )
    )
)
```

The p-value ($Pr(>F)$) for `r end_counts$samples$arm` is `r pv$p`, which is greater than the conventional significance level of 0.05. This suggests that we cannot reject the null hypothesis, implying that there is no significant overall difference in microbial pathway abundance between endline and endpoint.

The `plotMDS()` function accepts "any data object that can be coerced to a matrix of log-expression values, for example an ExpressionSet or an EList" [@Smyth2025]. If the `edgeR::voomLmFit()` function was used the `EList` object can be obtained using `fit$EList`

```{r}
#| eval: true
#| label: MDS plot comparing arms at endline in paired samples.

xmax = max(
  plotMDS(
    v$EList,
    col = group_colours,
    dim.plot = c(1, 2),
    pch = 16,
    gene.selection = "common"
  )$x
) -
  0.05
ymax = max(
  plotMDS(
    v$EList,
    col = group_colours,
    dim.plot = c(1, 2),
    pch = 16,
    gene.selection = "common"
  )$y
) -
  0.05

par(mar = c(7.1, 5.1, 5.1, 5.1), xpd = TRUE)
plotMDS(
  v$EList,
  col = group_colours,
  dim.plot = c(1, 2),
  pch = 16,
  gene.selection = "common"
)
legend(
  "bottom",
  pch = 16,
  col = unique(group_colours),
  legend = levels(v$targets$arm),
  inset = c(0, -0.165),
  bty = "n",
  ncol = 2
)
text(
  x = xmax - 0.05,
  y = ymax,
  labels = bquote(R^2 == .(round(as.data.frame(mds_stat[2])[1, 3], 3)))
)
text(
  x = xmax - 0.05,
  y = ymax - 0.12,
  labels = bquote(p == .(round(as.data.frame(mds_stat[2])[1, 5], 3)))
)
```

#### Change in abundances overtime (ignoring arm)

```{r}
#| eval: true
counts <- go_be %>%
  dplyr::filter(ontology != "Cellular component") %>%
  dplyr::select(go_id, sample, count) %>%
  pivot_wider(names_from = sample, values_from = count, values_fill = 0) %>%
  as.data.frame()

counts <- DGEList(as.data.frame(counts))

counts$samples$arm <- as.factor(go_be$arm[match(
  rownames(counts$samples),
  go_be$sample
)])
counts$samples$timepoint <- go_be$timepoint[match(
  rownames(counts$samples),
  go_be$sample
)]
counts$samples$timepoint <- factor(
  counts$samples$timepoint,
  labels = c("Baseline", "Endpoint")
)

counts$genes <- go_be %>% select(go_id, Terms, ontology) %>% as.data.frame()

counts.keep <- rowSums(counts$counts == 0) == ncol(counts$counts)
counts <- counts[counts.keep == FALSE, ]

counts <- normLibSizes(counts)
counts <- estimateDisp(counts)
```

```{r}
#| eval: true
#| label: GO functional terms differences between treatment arms at baseline and endpoint
#Linear Model
#Contrasts
design <- stats::model.matrix(~timepoint, data = counts$samples)

colnames(design) <- c("Intercept", "Endpoint")

contrasts <- makeContrasts(
  EvsB = Endpoint,
  levels = design
)
```

```{r}
#| eval: true
#| label: glmQLfit
#| fig-cap: NA
#| tbl-cap: NA
fit <- glmQLFit(counts, design)

qlf <- glmQLFTest(fit, contrast = contrasts)
glm_results <- topTags(qlf, n = Inf) %>%
  as.data.frame()

glm_results %>%
  filter(abs(logFC) >= log2(2)) %>%
  ggplot(aes(y = reorder(Terms, logFC), x = logFC)) +
  geom_bar(
    position = position_dodge(),
    stat = "identity",
    aes(fill = ontology)
  ) +
  xlab(bquote(bold("Fold change (" ~ Log[2] ~ "FC )"))) +
  ylab(NULL) + #labels the y axis
  scale_fill_manual(values = onto_col)
```

####### Empirical Bayes Statistics for Differential Expression 

```{r}
#| eval: true
#| label: ebayes
#| fig-cap: NA
#| tbl-cap: NA

#Transform raw count data to log2-counts per million (logCPM), estimate the mean-variance relationship and use this to compute appropriate observation-level weights. The data are then ready for linear modelling.

#If the DGEList object has been scale-normalized in edgeR (or before), then leave normalize.method="none" in voom, i.e., the logCPM values should not usually be re-normalized in the voom call.
#go_be abundance values have not been normalised, however initial normalization duing raw sequence QC will have accounted for differences in sequencing depth/library size. go_be counts should reflect these adjustments and further normalization should not be necessary.

vfit <- voomLmFit(counts, design, normalize.method = "none") #Transforms count data to log2-counts per million
vfit <- contrasts.fit(vfit, contrasts = contrasts)
efit <- eBayes(vfit) #estimation of QL dispersions
fc1.2 <- treat(vfit, fc = 1.2)
fc2 <- treat(vfit, fc = 2)

```

```{r}
#| eval: true

top_efit <- as.data.frame(topTable(efit, n = Inf)) %>%
  select(
    go_id,
    EvsB = logFC,
    PValue = P.Value,
    FDR = adj.P.Val
  )

top_efit %>%
  arrange(go_id) %>%
  mutate(EvsB_FC = 2^abs(EvsB)) %>%
  tibble()
```

```{r}
#| eval: true
#| label: tbl-fdr5
#| tbl-cap: Number of GO terms that are significantly differentially expressed (p $\leq$ 0.05)
# input = a fit object (eBayes) containing p-values and test statistics.

knitr::kable(summary(
  FDR5 <- decideTests(efit, adjust.method = "BH", p.value = 0.05, fc = 0)
))

```

```{r}
#| eval: true
#| label: tbl-fdr5-fc1.2
#| tbl-cap: Number of GO terms that are significantly differentially expressed (p $\leq$ 0.05) and have a fold change of 1.2 or more

#FDR of 0.05 & FC > 1.2
knitr::kable(summary(
  FDR5.FC1.2 <- decideTests(fc1.2, adjust.method = "BH", p.value = 0.05)
))
```

```{r}
#| eval: true
#| label: tbl-fdr-fc2
#| tbl-cap: Number of GO terms that are significantly differentially expressed (p $\leq$ 0.05) and have a fold change of 2 or more

#FDR of 0.05 & FC > 2
knitr::kable(summary(
  FDR5.FC2 <- decideTests(fc2, adjust.method = "BH", p.value = 0.05)
))
```

###### Unsupervised ordination: MDS plot og GO terms between timepoints

```{r}
#| eval: true
#| label: MDS plot EvsB
#unsupervised learning: MDS plot
group_colours <- counts$samples$timepoint
levels(group_colours) <- c("Baseline" = "#66C2A5", "Endpoint" = "#E78AC3")
group_colours <- as.character(group_colours)
df <- as.data.frame(t(counts$counts))
df <- cbind(df, group = counts$samples$timepoint)

#PERMANOVA testing differenital abundances between timepoints
mds_stat <- pairwiseAdonis::pairwise.adonis2(
  df[, 1:(ncol(df) - 1)] ~ group,
  data = df,
  method = 'eu'
) #returns non-adjusted p-values

#output table
knitr::kable(mds_stat[2])
```

```{r}
#| eval: true
pv <- lapply(2:length(mds_stat), function(x) {
  w <- mds_stat[[x]][1, 4]
  y <- mds_stat[[x]][1, 5]
  z <- names(mds_stat)[x]
  data.frame(contrast = z, p = y, F.stat = w)
})

pv <- bind_rows(pv)
knitr::kable(
  pv %>%
    mutate(p.adj = round(p.adjust(p, method = "BH"), 4)) %>%
    mutate(
      sig = case_when(
        p.adj <= 0.0001 ~ "****",
        p.adj <= 0.001 ~ "***",
        p.adj <= 0.01 ~ "**",
        p.adj <= 0.05 ~ "*",
        p.adj > 0.05 ~ "ns"
      )
    )
)
```

The p-value ($Pr(>F)$) for `r counts$samples$timepoint` is `r pv$p`, which is greater than the conventional significance level of 0.05. This suggests that we cannot reject the null hypothesis, implying that there is no significant overall difference in microbial pathway abundance between baseline and endpoint.

```{r}
#| eval: true
#| label: fig-EvB-MDS
#| fig-cap: MDS plot comparing timepoints

xmax <- max(
  plotMDS(
    vfit$EList,
    col = group_colours,
    dim.plot = c(1, 2),
    pch = 16,
    gene.selection = "common"
  )$x
) -
  0.05
ymax = max(
  plotMDS(
    vfit$EList,
    col = group_colours,
    dim.plot = c(1, 2),
    pch = 16,
    gene.selection = "common"
  )$y
) -
  0.05

par(mar = c(7.1, 5.1, 5.1, 5.1), xpd = TRUE)
plotMDS(
  vfit$EList,
  col = group_colours,
  dim.plot = c(1, 2),
  pch = 16,
  gene.selection = "common"
)
legend(
  "bottom",
  pch = 16,
  col = unique(group_colours),
  legend = levels(counts$samples$arm),
  inset = c(0, -0.2),
  bty = "n",
  ncol = 2
)
text(
  x = xmax - 0.05,
  y = ymax,
  labels = bquote(R^2 == .(round(as.data.frame(mds_stat[2])[1, 3], 3)))
)
text(
  x = xmax - 0.05,
  y = ymax - 0.12,
  labels = bquote(p == .(round(as.data.frame(mds_stat[2])[1, 5], 3)))
)
```

## Results Summary

```{r}
#| eval: true
#| echo: false
#| results: asis

list_all_results()
```

```{r}
#| eval: true
#| echo: false
#| results: asis
generate_results_section(directory_naming = "both")
```

## Appendix

```{r}
#| eval: true
#| echo: false
#| label: info

print(sessionInfo(), locale = FALSE)
```