## Baseline-Endpoint Beta Diversity

This analysis examines how bacterial communities differ between samples, using multiple distance metrics to capture different aspects of community similarity.

Four different distance metrics were calculated:

1. *Bray-Curtis*: Considers abundance data, sensitive to dominant species
2. *Jaccard*: Based on presence/absence, treats all species equally
3. *Unweighted UniFrac*: Incorporates phylogenetic relationships, presence/absence
4. *Weighted UniFrac*: Combines phylogenetic relationships with abundance data

```{r}
#| echo: false
#| eval: true
#| label: setup

# This initial code block loads custom utility functions and environment variables
# These are stored in separate files to keep the main analysis clean and organized
source("utils/r-utils.r")
source("_environment.required")
load("data/seqdata.RData")

suppressPackageStartupMessages(library(vegan))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(pairwiseAdonis))

```

```{r}
#| eval: true
#| label: BE beta diversity distance matrix calculation

bray <- phyloseq::distance(bac_base_end, method = "bray", type = "samples") #calculated on abundance
jac <- phyloseq::distance(bac_base_end, method = "jaccard", type = "samples")
uunifrac <- phyloseq::distance(
  bac_base_end,
  method = "uunifrac",
  type = "samples"
)
wunifrac <- phyloseq::distance(
  bac_base_end,
  method = "wunifrac",
  type = "samples"
)
```

### Beta diversity statistial tests for differences

Three complementary approaches were used to test for community differences:

::: {.grid}

::: {.g-col-4}

**ANOSIM (Analysis of Similarity)**

Tests whether between-group differences are larger than within-group differences

- R statistic ranges from -1 to 1
- R ≈ 0: no difference between groups
- R > 0: groups are more different than expected by chance
- Stratification by timepoint accounts for temporal effects

:::

::: {.g-col-4}

**PERMDISP (Permutational Analysis of Multivariate Dispersions)**

Tests for homogeneity of dispersions between groups

- Important assumption check for PERMANOVA
- Analogous to Levene's test for univariate data
- Significant results could indicate different within-group variability

:::

::: {.g-col-4}

**PERMANOVA (Permutational Multivariate Analysis of Variance)**

Tests for differences in community centroids between groups

- Accounts for both treatment and time effects
- More robust than ANOSIM
- Effect sizes calculated using partial eta-squared and omega-squared
- Requires homogeneous dispersions (tested by PERMDISP)

:::

:::

#### ANOSIM: Analysis of Similarity

ANOSIM is a permutation test whcih calculates R statistic (-1,1) $\approx$ correlation coefficient $\approx$ effect size

$H_0$ = there is no difference between intervention arms in terms of beta diversity metrics ($R = 0$)

$H_1$ =  biofortified pearl millet differentlaly changes beta diversity metrics ($R \neq 0$)

```{r}
#| eval:  true
#| label: ANOSIM
#function requires that rows = sample and column = OGU
summary(vegan::anosim(
  bray,
  grouping = microbiome::meta(bac_base_end)$arm,
  strata = microbiome::meta(bac_base_end)$timepoint
)) #there is no difference in Bray Curtis distance between treatment groups
summary(vegan::anosim(
  jac,
  grouping = microbiome::meta(bac_base_end)$arm,
  strata = microbiome::meta(bac_base_end)$timepoint
)) #there is no difference in Jaccard distance between treatment groups
summary(vegan::anosim(
  uunifrac,
  grouping = microbiome::meta(bac_base_end)$arm,
  strata = microbiome::meta(bac_base_end)$timepoint
)) #there is no difference in unweighted unifrac distance between treatment groups
summary(vegan::anosim(
  wunifrac,
  grouping = microbiome::meta(bac_base_end)$arm,
  strata = microbiome::meta(bac_base_end)$timepoint
)) #there is no difference in weighted unifrac distance between treatment groups
```

#### PERMDISP

PERMANOVA assumes homogeneity of multivariate dispersions across groups therefore a significant PERMANOVA result could be due to dispersion, location or both many permutational techniqures can be sensitive to dispersion (differences in centroids and/or differences in how data points in clusters scatter around their centroid). In univariate analysis Levene's test can test for this. PERMDISP is an extension of this that can be used in multivariate analysis technique involves calculating the distance from each data point to its group centroid and then testing whether those distances differ among the groups however, certain metrics are non-metric or semi-metric e.g. Bray-Curtis and centroids cannot be directly calculated -> PCoA + cantroid calc on the coordinates ([cite](https://uw.pressbooks.pub/appliedmultivariatestatistics/chapter/permdisp/))

```{r}
#| eval:  true
#| label: PERMDISP
bray_disp <- betadisper(
  d = bray,
  group = microbiome::meta(bac_base_end)$arm,
  type = "centroid"
)
permutest(bray_disp, pairwise = F, group = timepoint)
report::report_table(anova(bray_disp))

jac_disp <- betadisper(
  d = jac,
  group = microbiome::meta(bac_base_end)$arm,
  type = "centroid"
)
permutest(jac_disp, pairwise = F, group = timepoint)
report::report_table(anova(jac_disp))

uunifrac_disp <- betadisper(
  d = uunifrac,
  group = microbiome::meta(bac_base_end)$arm,
  type = "centroid"
)
permutest(uunifrac_disp, pairwise = F, group = timepoint)
report::report_table(anova(uunifrac_disp))

wunifrac_disp <- betadisper(
  d = wunifrac,
  group = microbiome::meta(bac_base_end)$arm,
  type = "centroid"
)
permutest(wunifrac_disp, pairwise = F, group = timepoint)
report::report_table(anova(wunifrac_disp))

#variance (dispersion) is homogenous among treatment groups (P-val > 0.9)
#note that this does NOT account for time
```

#### PERMANOVA

$H_0$ =  there is no change in beta diversity metrics arosss time (baseline to endpoint)
$H_1$ =  beta diversity metrics change (increase) across time (baseline to endpoint)
assumption: variance of dispersions are homogenous across groups

```{r}
#| eval: true
#| label: PERMANOVA results output
(bray_adon <- adonis2(
  bray ~ timepoint * arm,
  data = microbiome::meta(bac_base_end)
))

(jac_adon <- adonis2(
  jac ~ timepoint * arm,
  data = microbiome::meta(bac_base_end)
))

(wunifrac_adon <- adonis2(
  wunifrac ~ timepoint * arm,
  data = microbiome::meta(bac_base_end)
))

(uunifrac_adon <- adonis2(
  uunifrac ~ timepoint * arm,
  data = microbiome::meta(bac_base_end)
))

```

#### Effect Size Calculations

##### Partial Eta-Squared Analysis

PERMANOVA effect size is typically measured using the partial eta-squared ($\eta^2_p$) statistic. The partial eta-squared is a measure of the proportion of total variance in the multivariate data explained by the grouping variable (i.e., the factor of interest). In other words, $\eta^2_p$ calculations assess the magnitude of community differences.

::: {.callout-warn}

This measure tends to overestimate effect sizes in small samples

:::

$$
(\eta^2_p = \frac{SS_{\text{between}}}{SS_{\text{total}}})
$$

- Partial eta-squared ranges from 0 to 1.
- Values < 0.6 indicate small effect sizes
- A value of 0 indicates that the grouping variable explains none of the variance in the dependent variable.
- A value of 1 indicates that the grouping variable explains all of the variance in the dependent variable.
- Can be interpreted similarly to $R^2$ in linear regression
- In the context of PERMANOVA, $\eta^2_p$ indicates the proportion of total variance in the multivariate data (e.g., beta diversity distances) that is explained by the grouping variable.

```{r}
#| eval: true
#| label: eta squared
#calculate effect size for permanova: eta-squared
etasq <- function(adonis) {
  SS_between <- adonis$SumOfSqs[1]
  SS_total <- adonis$SumOfSqs[length(adonis$SumOfSqs)]
  etasq <- SS_between / SS_total
  return(etasq)
}

etasq(adonis2(bray ~ timepoint * arm, data = microbiome::meta(bac_base_end)))
etasq(adonis2(jac ~ timepoint * arm, data = microbiome::meta(bac_base_end)))
etasq(adonis2(
  wunifrac ~ timepoint * arm,
  data = microbiome::meta(bac_base_end)
))
etasq(adonis2(
  uunifrac ~ timepoint * arm,
  data = microbiome::meta(bac_base_end)
))

#all <0.6 => small effect size
```

There are some concerns over the use of $\eta^2_p$ to determine effect size. As it describes the ratio of variance explained by a predictor variable ($x$), it is often deemed analogous to $r^2$. $\eta^2_p$ is thus a biased estimate of how much population variance is explained by the model (it measures the variance explained of the sample, not the population, meaning that it will always overestimate the effect size). Each additional variable added to the model will automatically increase $\eta^2_p$ (the more overparamaterised a model is the closer to a "perfect" fit it becomes i.e. the model accounts for total variance).

Omega-squared ($\omega^2$) provides a less biased measure of effect size for ANOVA-type analyses by accounting for the mean-squared error of the observed samples [@Olejnik2009]

$$
\omega^2 = \frac{SS_A - (a-1)\frac{SS_W}{(N-a)}}{SS_T +\frac{SS_W}{(N-a)}}
$$

Where:

- $SS_T$: total sum of squares
- $SS_W$: within-group sum of squares (sum of the squares of distances within groups divided by the number of subjects per group)
- $SS_A$: between-group sum of squares ($SS_A =SS_T − SS_W$)
- $(a − 1)$: degress of freedom of the grouping factor
- $(N - a)$: residual degrees of freedom

```{r}
#| eval: true
#| label: omega squared
omegasq <- function(aovm) {
  sum_stats <- aovm
  SSm <- sum_stats[["SumOfSq"]][1]
  SSr <- sum_stats[["SumOfSq"]][2]
  DFm <- sum_stats[["Df"]][1]
  MSr <- sum_stats[["Mean Sq"]][2]
  W2 <- (SSm - DFm * MSr) / (SSm + SSr + MSr)
  return(W2)
}

omegasq(adonis2(bray ~ timepoint * arm, data = microbiome::meta(bac_base_end)))
```

### Ordination Methods

Two different ordination approaches were used to visualize community differences:

#### Metric Multidimensional Scaling (MDS)

MDS creates a visualization where:

- Similar samples are placed closer together
- Distances between points approximate their actual dissimilarities
- `trymax = 100` ensures a stable solution by trying multiple starting configurations

```{r}
#| eval:  true
#| label: beta div MDS ordination

#perform metric multidimensional scaling
bray.ord.mds <- phyloseq::ordinate(
  bac_base_end,
  method = "MDS",
  distance = bray,
  trymax = 100
)
jac.ord.mds <- phyloseq::ordinate(
  bac_base_end,
  method = "MDS",
  distance = jac,
  trymax = 100
)
uunifrac.ord.mds <- phyloseq::ordinate(
  bac_base_end,
  method = "MDS",
  distance = uunifrac,
  trymax = 100
)
wunifrac.ord.mds <- phyloseq::ordinate(
  bac_base_end,
  method = "MDS",
  distance = wunifrac,
  trymax = 100
)
```

#### Non-metric Multidimensional Scaling (NMDS)

NMDS is an iterative rank based procedure that sets stress values rather than eigenvalues to determine the "success" of the ordination. It is typically to aim for stress values $\leq$ 0.1, which indicates a good representation of the relationship between values.

NMDS is often preferred for ecological data because:

- It works with rank-order distances rather than absolute distances
- More robust to non-linear relationships
- Stress value indicates how well the ordination represents the data

```{r}
#| eval:  true
#| label: beta div NMDS ordination
bray.ord.nmds <- phyloseq::ordinate(
  bac_base_end,
  method = "NMDS",
  distance = bray,
  trymax = 100
)
jac.ord.nmds <- phyloseq::ordinate(
  bac_base_end,
  method = "NMDS",
  distance = jac,
  trymax = 100
)
uunifrac.ord.nmds <- phyloseq::ordinate(
  bac_base_end,
  method = "NMDS",
  distance = uunifrac,
  trymax = 100
)
wunifrac.ord.nmds <- phyloseq::ordinate(
  bac_base_end,
  method = "NMDS",
  distance = wunifrac,
  trymax = 100
)
```

#### Ordination Plots with PERMANOVA $R^2$ and P-values

##### MDS

::: {#fig-box}

```{r}
#| eval:  true
#| label: fig-beta-mds
#| fig-height: 12
#| fig-width: 16
bray.plot.mds <- plot_ordination(
  bac_base_end,
  bray.ord.mds,
  color = "timepoint",
  shape = "arm"
) +
  labs(
    title = "Bray-Curtis Distance",
    color = "Timepoint",
    shape = "Intervention Arm"
  ) +
  geom_point(size = 4) +
  scale_color_discrete(labels = time_labels) +
  theme(
    legend.title = element_text(
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    plot.title = element_text(size = 12)
  ) +
  scale_shape_manual(values = c(16, 17), labels = trt_labels) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    hjust = 2,
    vjust = 2,
    label = bquote(atop(
      R^2 == .(round(bray_adon[1, 3], 3)),
      p == .(round(bray_adon[1, 5], 3))
    ))
  )

jac.plot.mds <- plot_ordination(
  bac_base_end,
  jac.ord.mds,
  color = "timepoint",
  shape = "arm"
) +
  labs(
    title = "Jaccard Disimilarity",
    color = "Timepoint",
    shape = "Intervention Arm"
  ) +
  geom_point(size = 4) +
  scale_color_discrete(labels = time_labels) +
  theme(
    legend.title = element_text(
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    plot.title = element_text(size = 12)
  ) +
  scale_shape_manual(values = c(16, 17), labels = trt_labels) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    hjust = 2,
    vjust = 2,
    label = bquote(atop(
      R^2 == .(round(jac_adon[1, 3], 3)),
      p == .(round(jac_adon[1, 5], 3))
    ))
  )

uunifrac.plot.mds <- plot_ordination(
  bac_base_end,
  uunifrac.ord.mds,
  color = "timepoint",
  shape = "arm"
) +
  labs(
    title = "Unweighted UniFrac",
    color = "Timepoint",
    shape = "Intervention Arm"
  ) +
  geom_point(size = 4) +
  scale_color_discrete(labels = time_labels) +
  theme(
    legend.title = element_text(
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    plot.title = element_text(size = 12)
  ) +
  scale_shape_manual(values = c(16, 17), labels = trt_labels) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    hjust = 2,
    vjust = 2,
    label = bquote(atop(
      R^2 == .(round(uunifrac_adon[1, 3], 3)),
      p == .(round(uunifrac_adon[1, 5], 3))
    ))
  )

wunifrac.plot.mds <- plot_ordination(
  bac_base_end,
  wunifrac.ord.mds,
  color = "timepoint",
  shape = "arm"
) +
  labs(
    title = "Weighted UniFrac",
    color = "Timepoint",
    shape = "Intervention Arm"
  ) +
  geom_point(size = 4) +
  scale_color_discrete(labels = time_labels) +
  theme(
    legend.title = element_text(
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    plot.title = element_text(size = 12)
  ) +
  scale_shape_manual(values = c(16, 17), labels = trt_labels) +
  annotate(
    "text",
    x = -Inf,
    y = Inf,
    hjust = -1,
    vjust = 2,
    label = bquote(atop(
      R^2 == .(round(wunifrac_adon[1, 3], 3)),
      p == .(round(wunifrac_adon[1, 5], 3))
    ))
  )

ggarrange(
  bray.plot.mds,
  jac.plot.mds,
  uunifrac.plot.mds,
  wunifrac.plot.mds,
  ncol = 2,
  nrow = 2,
  common.legend = T,
  legend = "bottom",
  labels = "AUTO"
)

```

**β-diversity ordination plots comparing intervention arms at baseline and endpoint:** There were no differences in  community structure, nor dispersion of sequences by treatment group for any β-diversity metric measured (PERMDISP; p > 0.05).  However, overall, there was a small shift in community structure at endpoint compared to baseline (determined by PERMANOVA) for (A) Bray Curtis distances (F statistic: `r bray_adon$F[1]`; p = `r bray_adon$'Pr(>F)'[1]`), (B) Jaccard similarity (F statistic: `r jac_adon$F[1]`; p = `r jac_adon$'Pr(>F)'[1]`), (C) Unweighted UniFrac (F  statistic: `r uunifrac_adon$F[1]`; p = `r uunifrac_adon$'Pr(>F)'[1]`), and (D) Weighted UniFrac (F statistic: `r wunifrac_adon$F[1]`; p = `r wunifrac_adon$'Pr(>F)'[1]`).

:::

#### NMDS

::: {fig-box}

```{r}
#| eval: true
#| label: fig-beta-nmds
#| fig-height: 12
#| fig-width: 16

bray.plot.nmds <- plot_ordination(
  bac_base_end,
  bray.ord.nmds,
  color = "timepoint",
  shape = "arm"
) +
  labs(
    title = "Bray-Curtis Distance",
    color = "Timepoint",
    shape = "Intervention Arm"
  ) +
  geom_point(size = 4) +
  #ggalt::geom_encircle(aes(group = arm, fill = arm), size = 1, alpha = 0.1) +
  scale_color_discrete(labels = time_labels) +
  theme(
    legend.title = element_text(
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    plot.title = element_text(size = 12)
  ) +
  scale_shape_manual(values = c(16, 17), labels = trt_labels) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    hjust = 2,
    vjust = 2,
    label = bquote(atop(
      R^2 == .(round(bray_adon[1, 3], 3)),
      p == .(round(bray_adon[1, 5], 3))
    ))
  )

jac.plot.nmds <- plot_ordination(
  bac_base_end,
  jac.ord.nmds,
  color = "timepoint",
  shape = "arm"
) +
  labs(
    title = "Jaccard Disimilarity",
    color = "Timepoint",
    shape = "Intervention Arm"
  ) +
  geom_point(size = 4) +
  scale_color_discrete(labels = time_labels) +
  theme(
    legend.title = element_text(
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    plot.title = element_text(size = 12)
  ) +
  scale_shape_manual(values = c(16, 17), labels = trt_labels) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    hjust = 2,
    vjust = 2,
    label = bquote(atop(
      R^2 == .(round(jac_adon[1, 3], 3)),
      p == .(round(jac_adon[1, 5], 3))
    ))
  )

uunifrac.plot.nmds <- plot_ordination(
  bac_base_end,
  uunifrac.ord.nmds,
  color = "timepoint",
  shape = "arm"
) +
  labs(
    title = "Unweighted UniFrac",
    color = "Timepoint",
    shape = "Intervention Arm"
  ) +
  geom_point(size = 4) +
  scale_color_discrete(labels = time_labels) +
  theme(
    legend.title = element_text(
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    plot.title = element_text(size = 12)
  ) +
  scale_shape_manual(values = c(16, 17), labels = trt_labels) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    hjust = 2,
    vjust = 2,
    label = bquote(atop(
      R^2 == .(round(uunifrac_adon[1, 3], 3)),
      p == .(round(uunifrac_adon[1, 5], 3))
    ))
  )

wunifrac.plot.nmds <- plot_ordination(
  bac_base_end,
  wunifrac.ord.nmds,
  color = "timepoint",
  shape = "arm"
) +
  labs(
    title = "Weighted UniFrac",
    color = "Timepoint",
    shape = "Intervention Arm"
  ) +
  geom_point(size = 4) +
  scale_color_discrete(labels = time_labels) +
  theme(
    legend.title = element_text(
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    plot.title = element_text(size = 12)
  ) +
  scale_shape_manual(values = c(16, 17), labels = trt_labels) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    hjust = 2,
    vjust = 2,
    label = bquote(atop(
      R^2 == .(round(wunifrac_adon[1, 3], 3)),
      p == .(round(wunifrac_adon[1, 5], 3))
    ))
  )

ggarrange(
  bray.plot.nmds,
  jac.plot.nmds,
  uunifrac.plot.nmds,
  wunifrac.plot.nmds,
  ncol = 2,
  nrow = 2,
  common.legend = T,
  legend = "bottom",
  labels = "AUTO"
)

```

**β-diversity ordination plots comparing intervention arms at baseline and endpoint:** There were no differences in  community structure, nor dispersion of sequences by treatment group for any β-diversity metric measured (PERMDISP; p > 0.05).  However, overall, there was a small shift in community structure at endpoint compared to baseline (determined by PERMANOVA) for (A) Bray Curtis distances (F statistic: `r bray_adon$F[1]`; p = `r bray_adon$'Pr(>F)'[1]`), (B) Jaccard similarity (F statistic: `r jac_adon$F[1]`; p = `r jac_adon$'Pr(>F)'[1]`), (C) Unweighted UniFrac (F  statistic: `r uunifrac_adon$F[1]`; p = `r uunifrac_adon$'Pr(>F)'[1]`), and (D) Weighted UniFrac (F statistic: `r wunifrac_adon$F[1]`; p = `r wunifrac_adon$'Pr(>F)'[1]`).

:::

### Beta diversity summary tables

```{r}
#| eval: true
#| label: Base-End Bacteria Beta diversity stat summary table

subsets <- list(
  baseCPM = phyloseq::subset_samples(
    bac_base_end,
    timepoint == "Baseline" & arm == "CPM"
  ),
  endCPM = phyloseq::subset_samples(
    bac_base_end,
    timepoint == "Endpoint" & arm == "CPM"
  ),
  baseFe = phyloseq::subset_samples(
    bac_base_end,
    timepoint == "Baseline" & arm == "FeZnPM"
  ),
  endFe = phyloseq::subset_samples(
    bac_base_end,
    timepoint == "Endpoint" & arm == "FeZnPM"
  )
)

list <- list(
  bray_Baseline_CPM = phyloseq::distance(subsets$baseCPM, method = "bray"),
  jaccard_Baseline_CPM = phyloseq::distance(
    subsets$baseCPM,
    method = "jaccard"
  ),
  uunifrac_Baseline_CPM = phyloseq::distance(
    subsets$baseCPM,
    method = "uunifrac"
  ),
  wunifrac_Baseline_CPM = phyloseq::distance(
    subsets$baseCPM,
    method = "wunifrac"
  ),
  bray_Endpoint_CPM = phyloseq::distance(subsets$endCPM, method = "bray"),
  jaccard_Endpoint_CPM = phyloseq::distance(subsets$endCPM, method = "jaccard"),
  uunifrac_Endpoint_CPM = phyloseq::distance(
    subsets$endCPM,
    method = "uunifrac"
  ),
  wunifrac_Endpoint_CPM = phyloseq::distance(
    subsets$endCPM,
    method = "wunifrac"
  ),
  bray_Baseline_FeZnPM = phyloseq::distance(subsets$baseFe, method = "bray"),
  jaccard_Baseline_FeZnPM = phyloseq::distance(
    subsets$baseFe,
    method = "jaccard"
  ),
  uunifrac_Baseline_FeZnPM = phyloseq::distance(
    subsets$baseFe,
    method = "uunifrac"
  ),
  wunifrac_Baseline_FeZnPM = phyloseq::distance(
    subsets$baseFe,
    method = "wunifrac"
  ),
  bray_Endpoint_FeZnPM = phyloseq::distance(subsets$endFe, method = "bray"),
  jaccard_Endpoint_FeZnPM = phyloseq::distance(
    subsets$endFe,
    method = "jaccard"
  ),
  uunifrac_Endpoint_FeZnPM = phyloseq::distance(
    subsets$endFe,
    method = "uunifrac"
  ),
  wunifrac_Endpoint_FeZnPM = phyloseq::distance(
    subsets$endFe,
    method = "wunifrac"
  )
)

list <- lapply(list, as.numeric)

betadivCPM <- data.frame(
  list$bray_Baseline_CPM,
  list$jaccard_Baseline_CPM,
  list$uunifrac_Baseline_CPM,
  list$wunifrac_Baseline_CPM,
  list$bray_Endpoint_CPM,
  list$jaccard_Endpoint_CPM,
  list$uunifrac_Endpoint_CPM,
  list$wunifrac_Endpoint_CPM
)

betadivCPM$id <- seq_len(nrow(betadivCPM))
betadivCPM <- betadivCPM %>%
  pivot_longer(cols = -id, names_to = "variable", values_to = "value") %>%
  mutate(variable = str_remove_all(variable, "list|\\.")) %>%
  separate(variable, into = c("metric", "timepoint", "arm"), sep = "_") %>%
  pivot_wider(names_from = metric) %>%
  dplyr::select(!id)

betadivFeZnPM <- data.frame(
  list$bray_Baseline_FeZnPM,
  list$jaccard_Baseline_FeZnPM,
  list$uunifrac_Baseline_FeZnPM,
  list$wunifrac_Baseline_FeZnPM,
  list$bray_Endpoint_FeZnPM,
  list$jaccard_Endpoint_FeZnPM,
  list$uunifrac_Endpoint_FeZnPM,
  list$wunifrac_Endpoint_FeZnPM
)

betadivFeZnPM$id <- seq_len(nrow(betadivFeZnPM))
betadivFeZnPM <- betadivFeZnPM %>%
  pivot_longer(cols = -id, names_to = "variable", values_to = "value") %>%
  mutate(variable = str_remove_all(variable, "list|\\.")) %>%
  separate(variable, into = c("metric", "timepoint", "arm"), sep = "_") %>%
  pivot_wider(names_from = metric) %>%
  dplyr::select(!id)

bac_base_end_betadiv <- bind_rows(betadivCPM, betadivFeZnPM)
```

\blandscape

```{r}
#| eval:  true
#| label: tbl-beta-sum
#| tbl-cap:  Microbiome beta diversity at baseline and endpoint (n=82 repeated samples), overall and by arm. Beta diversity effect size calculations, as determined by the Vargha and Delaney A effect size measure
vda <- function(data, variable, by, ...) {
  # Compute VD.A
  A <- effsize::VD.A(as.formula(glue::glue("`{variable}` ~ {by}")), data = data)
  tibble(vda_estimate = A$estimate, vda_magnitude = A$magnitude)
}

(p <- bac_base_end_betadiv %>%
  tbl_strata(
    strata = timepoint,
    .tbl_fun = ~ .x %>%
      tbl_summary(
        by = arm,
        digits = everything() ~ 2,
        missing = "no",
        statistic = list(
          all_continuous() ~ "{median}  \n({IQR}; {min}-{max})", # stats and format for continuous columns
          all_categorical() ~ "{n} / {N}  \n({p}%)"
        )
      ) %>% # stats and format for categorical columns
      add_n() %>%
      add_overall() %>%
      add_p(everything() ~ "wilcox.test", pvalue_fun = function(x) {
        style_pvalue(x, digits = 3)
      }) %>%
      add_q(pvalue_fun = function(x) style_pvalue(x, digits = 3)) %>%
      bold_p(q = T) %>%
      add_stat(fns = all_continuous() ~ vda) %>%
      modify_fmt_fun(list(vda_magnitude ~ as.character)) %>%
      modify_header(list(
        vda_estimate ~ "**Effect Size Estimate**",
        vda_magnitude ~ "**Magnitude**"
      )),
    .header = "**{strata}**, N = {n}"
  ) %>%
  as_gt() %>%
  fmt_markdown(columns = where(~ is.character(.x))) %>%
  fmt_number(columns = everything(), drop_trailing_zeros = TRUE) %>%
  fmt_number(
    columns = where(~ is.numeric(.x) < 0.0001),
    pattern = "<0.0001"
  ) %>%
  fmt_number(columns = where(~ is.numeric(.x) < 0.001), pattern = "<0.001") %>%
  fmt_number(columns = where(~ is.numeric(.x) < 0.01), pattern = "<0.01") %>%
  fmt_integer(columns = where(~ is.numeric(.x) < 0.05), pattern = "<0.05"))

htmltools::save_html(
  gt::as_raw_html(p),
  file.path(RESULTS, "base-end-diversity-analysis/beta-div.tbl")
)

```

\elandscape