## Antibiotic use at Baseline

```{r}
#| echo: false
#| eval: true
#| label: setup

# This initial code block loads custom utility functions and environment variables
# These are stored in separate files to keep the main analysis clean and organized
suppressWarnings(suppressMessages({
  source(normalizePath("utils/r-utils.r"))
  source(normalizePath("_environment.required"))

  baseline_metadata <- readRDS(normalizePath(
    "data/metadata/baseline-metadata.csv"
  ))

  # dir.create(
  #   file.path(RESULTS, "baseline-kegg"),
  #   showWarnings = FALSE,
  #   recursive = TRUE
  # )
}))
```

**Antibiotic variables**

| Variable     | Description                                                        |
|:-------------|:-------------------------------------------------------------------|
| abx          | Child: frequency of Abx use                                        |
| abx_yn       | Mother knows child's antibiotics history or has MCH card available |
| abxage       | Child: Age when Abx taken                                          |
| abxc02       | Child: Abx use, current or within past 1 month                     |
| abxcy02      | If AbxC02=Yes: Which Abx                                           |
| abxkind1     | Child: Abx form = injection                                        |
| abxkind2     | Child: Abx form = Tablet/pill                                      |
| abxkind3     | Child: Abx form = liquid                                           |
| abxm02       | Mother: Abx during pregnancy                                       |
| abxmb02      | Mother: Abx during breastfeeding                                   |
| abxmbfy02    | If AbxM02=Yes: Which Abx                                           |
| abxmdel02    | Mother: Abx since delivery                                         |
| abxmdely02   | If AbxMdel02=Yes: Which Abx                                        |
| abxmweeks02  | Mother: Abx in past 4 wks                                          |
| abxmweeksy02 | If AbxMweeks02=Yes: Which Abx                                      |
| abxmy02      | If AbxMB02=Yes: Which Abx                                          |
| abxtype1     | Child: Abx name = penicillin (including amoxicillin)               |
| abxtype2     | Child: Abx name = cephalosporins                                   |
| abxtype3     | Child: Abx name = macrolides                                       |
| abxtype4     | Child: Abx name = penicillin                                       |
| abxtype5     | Child: Abx name = amoxicillin                                      |
| abxwhy       | Child: Abx use reason (illness)                                    |

### Types of antibiotics used among baseline cohort

```{r}
#| eval: true
#| label: antibiotic use
baseline_metadata %>%
  filter(timepoint == "Baseline") %>%
  select(
    `Any past antibiotic use` = abx,
    `Antibiotic history known` = abx_yn,
    `Age when antibiotics taken` = abxage,
    `penicillin (including amoxicillin)` = abxtype1,
    cephalosporins = abxtype2,
    macrolides = abxtype3,
    penicillin = abxtype4,
    amoxicillin = abxtype5,
    reason = abxwhy
  ) %>%
  mutate_at(vars(everything()), type.convert, as.is = T) %>%
  mutate(
    across(
      !c(reason),
      ~ case_when(
        . == 1 ~ "Yes",
        . == 0 ~ "No",
        . == 8 ~ "Don't know",
        . == 9 ~ "Not applicable",
        TRUE ~ "Unknown"
      )
    )
  ) %>%
  tbl_summary(
    missing = "ifany",
    statistic = list(
      all_continuous() ~ "{median} ({IQR}; {min}-{max})",
      all_categorical() ~ "{n} / {N} ({p}%)"
    )
  ) %>%
  as_kable_extra(format = "markdown", escape = TRUE)
```

### Differences in Alpha diveristy among Children with known Antibiotic History

A total of 25 mothers were able to provide information on their child's past antibiotic use:
Reported antibiotic use = 15
Reported no antibiotic use = 10

```{r}
#| eval: true
#| label: tbl-
#| tbl-cap: alpha diveristy between antibiotic reports

kable(
  tmp <- baseline_metadata %>%
    select(
      abx,
      abx_yn,
      observed,
      chao1,
      diversity_shannon,
      evenness_pielou,
      pd
    ) %>%
    mutate_at(vars(everything()), str_trim) %>%
    type.convert(., as.is = TRUE) %>%
    mutate(
      abx = case_when(
        abx == 1 ~ "Never",
        abx == 2 ~ "1-2 times",
        abx == 3 ~ "3-4 times",
        abx == 4 ~ "> 4 times",
        TRUE ~ "Unknown"
      )
    ) %>%
    filter(abx != "Unknown"),
  format = "markdown"
)
```

```{r}
#| eval: true
#| label: alpha diveristy between antibiotic reports

tmp %>%
  select(!abx_yn) %>%
  gt_summary(by = "abx")

```

```{r}
#| eval: true
#| label: alpha diveristy between antibiotic reports

stats <- tmp %>%
  pivot_longer(cols = !c(abx, abx_yn), names_to = "diversity_metric")

#print summary stats
kable(
  stats %>%
    group_by(abx, diversity_metric) %>%
    get_summary_stats(value, type = "common"),
  format = "markdown"
)

kable(
  stats %>%
    group_by(diversity_metric) %>%
    pairwise_wilcox_test(
      value ~ abx,
      paired = FALSE,
      p.adjust.method = "BH"
    ) %>%
    select(!c(.y., p.adj.signif)),
  format = "markdown"
)
```

### Antibiotic use among children with reported illness e.g. gastrointestinal symptoms

| Variable     | Description                                                        | Values                                                                                                                                                         |
|---------------------------|--------------------------------------------------------------------|--------------------------------------------------------------------------------------------------------------------------------------------------|
| abx          | Child: frequency of Abx use                                        | 1 Never; 2 1-2 times; 3 3-4 times; 4 \> 4 times                                                                                                                |
| abx_yn       | Mother knows child's antibiotics history or has MCH card available | 0=No; 1=Yes                                                                                                                                                    |
| diag         | Diagnosis                                                          | 1 Tuberculosis; 2 Malaria, Dengue Fever, HIV, or Tuberculosis; 3 Pneumonia; 4 Worms;  5 Tonsillitis;  6 Conjunctivitis;  7 Asthma;  8 Other;  9 Not Applicable |
| diagother    | Diagnosis if other                                                 | free text                                                                                                                                                      |
| gi_yn        | Diarrhea other gastrointestinal symptoms, today or in the past     | 0=No; 1=Yes                                                                                                                                                    |
| fever_yn     | Fever today or in the past                                         | 0=No; 1=Yes                                                                                                                                                    |
| resp_yn      | Respiratory symptoms, ever                                         | 0=No; 1=Yes                                                                                                                                                    |
| freqnoserun  | Frequency of runny nose                                            | 1 Today  2 Last 7 days 3 Last 4 weeks  4 None  8 Don't know                                                                                                    |
| freqtrblbrth | Frequency of trouble breathing                                     | 1 Today  2 Last 7 days  3 Last 4 weeks  4 None  8 Don't know                                                                                                   |
| freqvom      | Frequency of vomiting due to illness                               | 1 Today  2 Last 7 days  3 Last 4 weeks  4 None  8 Don't know                                                                                                   |
| health6m_bc  | Health status in last 6 months                                     | 1=healthy; 2=little problems with fever andor coldcough; 3=child always has some problem or other                                                              |
| mucuscough   | Frequency of coughing up mucus                                     | 1 Today  2 Last 7 days  3 Last 4 weeks  4 None  8 Don't know                                                                                                   |
| rash_bc      | Does child have a rash?                                            | 0=No; 1=Yes                                                                                                                                                    |
| respsigns1   | Respriatory signs: normal                                          | 0=No; 1=Yes                                                                                                                                                    |
| respsigns2   | Respriatory signs: cyanosis                                        | 0=No; 1=Yes                                                                                                                                                    |
| respsigns3   | Respriatory signs: Lower chest wall indrawing                      | 0=No; 1=Yes                                                                                                                                                    |
| respsigns4   | Respriatory signs: Nasal flaring                                   | 0=No; 1=Yes                                                                                                                                                    |
| respsigns5   | Respriatory signs: Rapid breathing                                 | 0=No; 1=Yes                                                                                                                                                    |
| respsigns6   | Respriatory signs: Pale skin                                       | 0=No; 1=Yes                                                                                                                                                    |
| respsigns7   | Respriatory signs: Cyanotic skin                                   | 0=No; 1=Yes                                                                                                                                                    |

```{r} 
#| eval: true
#| label: children reported to be sick in last 4 weeks
missing <- metadata %>%
  filter(any_other_relevant_information == "1Baseline") %>%
  select(
    freq3diarr,
    freqcough,
    freqdaysdiarr,
    freqdiarrblood,
    freqdiarrmucus,
    freqfever,
    freqfeverrigors,
    freqfeversweat,
    freqnoserun,
    freqtrblbrth,
    freqvom,
    mucuscough,
    rash_bc
  ) %>%
  mutate_at(vars(everything()), str_trim) %>%
  mutate_at(vars(everything()), type.convert, as.is = T) %>%
  summarise_all(~ sum(is.na(.)) / 168 * 100) %>%
  select(
    `Cough` = freqcough,
    `Diarrhea (> x3/day)` = freq3diarr,
    `Does the child have a rash` = rash_bc,
    `Fever` = freqfever,
    `Any blood-stained stools` = freqdiarrblood,
    `Any loose/watery stools` = freqdaysdiarr,
    `Any mucus cough` = mucuscough,
    `Any mucus- or pus-filled stools` = freqdiarrmucus,
    `Any rigors with fever` = freqfeverrigors,
    `Any runny nose` = freqnoserun,
    `Any sweating with fever` = freqfeversweat,
    `Any trouble breathing` = freqtrblbrth,
    `Any vomiting due to illness` = freqvom
  ) %>%
  t()

missing %>%
  kbl(format = "markdown", col.names = c("Symptom$^a$", "Unreported(%)"))
footnote(alphabet = "today or within past 4 weeks")

paste0(
  "Median percent missingness (IQR; min-max) = ",
  round(median(missing), 2),
  "% (",
  round(calculate_iqr(missing), 2),
  "; ",
  round(min(missing), 2),
  "-",
  round(max(missing), 2),
  ")."
)


tmp <- metadata %>%
  filter(any_other_relevant_information == "1Baseline") %>%
  select(
    abx,
    freq3diarr,
    freqcough,
    freqdaysdiarr,
    freqdiarrblood,
    freqdiarrmucus,
    freqfever,
    freqfeverrigors,
    freqfeversweat,
    freqnoserun,
    freqtrblbrth,
    freqvom,
    gi_yn,
    mucuscough,
    rash_bc
  ) %>%
  mutate_at(vars(everything()), str_trim) %>%
  mutate_at(vars(everything()), type.convert, as.is = T) %>%
  mutate(
    abx = case_when(
      abx == 1 ~ "Never",
      abx == 2 ~ "1-2 times",
      abx == 3 ~ "3-4 times",
      abx == 4 ~ "> 4 times",
      TRUE ~ "Unknown"
    )
  )

tmp %>%
  select(abx, gi_yn) %>%
  mutate(
    gi_yn = case_when(
      gi_yn == 1 ~ "Yes",
      gi_yn == 0 ~ "No",
      TRUE ~ "Unknown"
    )
  ) %>%
  group_by(abx, gi_yn) %>%
  summarise(n = n()) %>%
  ggplot(aes(x = abx, y = n)) +
  geom_col(aes(fill = gi_yn))
```

```{r}
#| eval: true
#| label: Illnesses and Abx
tmp <- metadata %>%
  filter(any_other_relevant_information == "1Baseline") %>%
  select(
    abx,
    abx_yn,
    diag,
    diagother,
    gi_yn,
    fever_yn,
    resp_yn,
    freqnoserun,
    freqtrblbrth,
    freqvom,
    health6m_bc,
    mucuscough,
    rash_bc,
    respsigns1,
    respsigns2,
    respsigns3,
    respsigns4,
    respsigns5,
    respsigns6,
    respsigns7,
    freq3diarr,
    freqfever,
    freqcough
  ) %>%
  mutate_at(vars(everything()), str_trim) %>%
  mutate_at(vars(everything()), type.convert, as.is = T) %>%
  mutate(
    across(
      c(
        abx_yn,
        gi_yn,
        fever_yn,
        resp_yn,
        rash_bc,
        respsigns1,
        respsigns2,
        respsigns3,
        respsigns4,
        respsigns5,
        respsigns6,
        respsigns7
      ),
      ~ case_when(
        . == 1 ~ "Yes",
        . == 0 ~ "No",
        TRUE ~ "Unknown"
      )
    )
  ) %>%
  mutate(
    across(
      c(
        freqnoserun,
        freqtrblbrth,
        freqvom,
        mucuscough,
        freq3diarr,
        freqfever,
        freqcough
      ),
      ~ case_when(
        . == 1 ~ "Today",
        . == 2 ~ "Last 7 days",
        . == 3 ~ "Last 4 weeks",
        . == 4 ~ "None",
        . == 8 ~ "Don't know",
        TRUE ~ "Unknown"
      )
    )
  ) %>%
  mutate(
    abx = case_when(
      abx == 1 ~ "Never",
      abx == 2 ~ "1-2 times",
      abx == 3 ~ "3-4 times",
      abx == 4 ~ "> 4 times",
      TRUE ~ "Unknown"
    ),
    diag = case_when(
      diag == 1 ~ "Tuberculosis",
      diag == 2 ~ "Malaria, Dengue Fever, HIV, or Tuberculosis",
      diag == 3 ~ "Pneumonia",
      diag == 4 ~ "Worms",
      diag == 5 ~ "Tonsillitis",
      diag == 6 ~ "Conjunctivitis",
      diag == 7 ~ "Asthma",
      diag == 8 ~ "Other",
      diag == 9 ~ "Not Applicable",
      TRUE ~ diagother
    ),
    health6m_bc = case_when(
      health6m_bc == 1 ~ "healthy",
      health6m_bc == 2 ~ "little problems with fever and/or cold/cough",
      health6m_bc == 3 ~ "child always has some problem or other",
      TRUE ~ "Unknown"
    )
  ) %>%
  select(
    `Any past antibiotic use` = abx,
    `Antibiotic history known` = abx_yn,
    `Health in the past 6 months` = health6m_bc,
    `Any current diagnosis of disease` = diag,
    `Any gastrointestinal symptoms, now or in past` = gi_yn,
    `Diarrhea today or within past 4 weeks ` = freq3diarr,
    `Frequency of vomiting` = freqvom,
    `Any respiratory symptoms, now or in past` = resp_yn,
    `Frequency of runny nose` = freqnoserun,
    `Frequency of mucus cough` = mucuscough,
    `Cough today or within past 4 weeks` = freqcough,
    `Frequency of trouble breathing` = freqtrblbrth,
    `Current respiratory signs: normal` = respsigns1, # 100% have normal signs
    `Does the child have a rash` = rash_bc,
    `Any fever, now or in past` = fever_yn,
    `Fever today or within past 4 weeks ` = freqfever,
    # `Respiratory signs: cyanosis` = respsigns2,
    # `Respiratory signs: Lower chest wall indrawing` = respsigns3,
    # `Respiratory signs: Nasal flaring` = respsigns4,
    # `Respiratory signs: Rapid breathing` = respsigns5,
    # `Respiratory signs: Pale skin` = respsigns6,
    # `Respiratory signs: Cyanotic skin` = respsigns7,
  )
```

```{r}
#| eval: true
#| label: those with reported illness
#| tab-cap: Reported antibiotic use among infants with reported illness
#| results: asis

tmp %>%
  filter(`Any past antibiotic use` != "Unknown") %>%
  kable(format = "markdown")
```

### Generate summary table of antibiotic and illness reports 

```{r}
#| eval: true
#| label: summary table
#| tab-cap: summary table of antibiotic and illness reports and global chisq association test

#generate the summary table
tmp %>%
  tbl_summary(
    by = `Any past antibiotic use`,
    missing = "ifany",
    statistic = list(
      all_continuous() ~ "{median} ({IQR}; {min}-{max})", # stats and format for continuous mns
      all_categorical() ~ "{n} / {N} ({p}%)"
    ), # stats and format for categorical columns
    type = list(
      all_continuous() ~ "continuous2",
      all_categorical() ~ "categorical"
    )
  ) %>%
  modify_header(label = "**Antibiotic Frequency**") %>%
  add_n() %>%
  add_p(
    pvalue_fun = function(x) style_pvalue(x, digits = 3),
    test = all_categorical() ~ "chisq.test"
  ) %>%
  as_kable_extra(format = "markdown", escape = TRUE)
```

### Global $\chi^2$ test of association between categorical variables

The critical value for the $\chi^2$ test statistic is based on the degrees of freedom and the chosen $\alpha$ level. This can be Bonferroni (or other) adjusted $\alpha$ value if there are several comparisons being made. 

If the test statistic is lower than the critical value, we **fail to reject the null hypothesis** (we do not have sufficient evidence to state that there is an association between variables). 

If the test statistic is greater than the critical value, we **reject the null hypothesis** (there is sufficient evidence to suggest an association between variables). 

To understand which levels in the two categorical variables being compared are contributing to the significant difference between variables (as suggested by the global $\chi^2$ test), we can look at the test residuals ($\varepsilon$), specifically the standarised ones. These (either raw or standardised)can be used to determine the magnitude of differences between variable levels. We can determine if a level is significantly different from the expected frequency value by comparing the absolute standardised residual ($\lvert \varepsilon \rvert$) to the normal distribution critical value for $\alpha = 0.05\Rightarrow 1.96$. 

In the case where there are numerous comparisons being made, an adjusted critical value can be used. Here, the Bonferronni method is used to calculate the new $\alpha$, given the total number of category level comparsions being made (total number of cells in the contingency table). Use this new $\alpha$ value to determine the adjusted critical value for a significant deviation from the expected value 

::: {.callout-tip}

Residuals represent the difference between the observed and expected value

$$
\varepsilon = y_i - \hat{y}_i
$$

:::

```{r}
#| eval: true
#| label: chisq association test

#any correlations between abx use and illness?
#**missing data is still data**

lapply(1:ncol(tmp), function(i) {
  tbl <- table(tmp[, 1], tmp[, i])
  a <- 0.05 / prod(dim(tbl))
  chi <- rstatix::chisq_test(tbl)
  Xcrit <- qchisq(p = a, df = chi$df, lower.tail = FALSE) # If TRUE, the probability to the left of p in the F distribution is returned. If FALSE, the probability to the right is returned. Default is TRUE.
  V <- rstatix::cramer_v(tbl)

  if (abs(chi$statistic) < Xcrit) {
    text <- c(
      paste0(
        "DF = ",
        chi$df,
        "; adjusted $\\\\alpha$ = ",
        round(a, 3),
        "; critical value = ",
        round(Xcrit, 3)
      ),
      paste0(
        "Since ",
        round(chi$statistic, 3),
        " > ",
        round(Xcrit, 3),
        " $\\\\Rightarrow \\\\textbf{fail reject the null hypothesis}$"
      ),
      paste0(
        "(insufficient evidence to support an association between variables)."
      ),
      paste0(" "),
      paste0("$\\\\textbf{No further investigation required}$")
    )
    resid_crit <- 100
  } else {
    chi_resid <- rstatix::std_residuals(chi)
    resid_crit <- qnorm(p = a, lower.tail = FALSE)
    text <- c(
      paste0(
        "DF = ",
        chi$df,
        "; Adjusted $\\\\alpha$ = ",
        round(a, 3),
        "; Critical Value = ",
        round(Xcrit, 3)
      ),
      paste0(
        "Since ",
        round(chi$statistic, 3),
        " > ",
        round(Xcrit, 3),
        " $\\\\Rightarrow \\\\textbf{reject the null hypothesis}$"
      ),
      paste0(
        "(there is sufficient evidence to suggest an association between variables)."
      ),
      paste0(" "),
      paste0(
        "$\\\\textbf{Compare the standardised residuals to determine which levels differ }$"
      ),
      paste0(
        "Adjusted $\\\\alpha$ = ",
        round(a, 3),
        "; Critical Value = ",
        round(Xcrit, 3)
      ),
      paste0(" Values in red represent significantly different residuals")
    )
  }

  df <- as.data.frame(rstatix::std_residuals(chi))
  colnames(df) <- c("Antibiotc Use", "Var2", "Freq")
  df %>%
    mutate(Freq = round(Freq, 3)) %>%
    mutate(
      Freq = cell_spec(
        Freq,
        color = ifelse(abs(Freq) > resid_crit, "red", "black"),
        escape = FALSE
      )
    ) %>%
    pivot_wider(names_from = "Var2", values_from = "Freq") %>%
    kbl(
      escape = FALSE,
      caption = paste(
        "Antibiotic useage Chi-squared association with:",
        colnames(tmp)[i],
        "(standardised residuals)"
      ),
      digits = 3,
      booktabs = TRUE,
      table.attr = "style='width:80%;'",
      align = "c",
      centering = FALSE
    ) %>%
    footnote(
      number = c(
        paste("Cramer's V:", round(V, 3)),
        paste(
          "Chi squared test-statistic:",
          round(chi$statistic, 3),
          " (p = ",
          round(chi$p, 3),
          ")"
        )
      ),
      general = text,
      footnote_as_chunk = FALSE,
      escape = FALSE,
      threeparttable = FALSE,
      general_title = "",
      footnote_order = c("number", "alphabet", "symbol", "general")
    )
})
```
