## Differential abundance between Intervention arms at Endpoint with ANCOM-BC

```{r}
#| echo: false
#| eval: true
#| label: setup
#| cache: false
suppressWarnings(suppressMessages({
  source(normalizePath("utils/r-utils.r"))
  source(normalizePath("_environment.required"))
  load(normalizePath("data/seqdata.RData"))

  library(ANCOMBC)
  library(promicrobialTools)

  prefix <- "endArm-DA-"
  filepath <- get_results_dir(directory_naming = "both")
}))
```

Results will be saved to ``r filepath``.

```{r}
#| eval: true
#| label: data-prep
#| fig-cap: NA
#| tbl-cap: NA
phyloseq_summary(
  endpoint <- phyloseq::subset_samples(seqdata_bac, timepoint == "Endpoint")
)

check <- data_sanity_check(endpoint, fix_formula = "arm")
```

::: {.callout-important}
The `ancombc2` function is needed for repeated measures (can be used for cross-sectional too and can accommodate continuous covariates). `ancombc` can only be used for cross-sectional data.
:::

::: {.callout-warning}
The ANOCOM-BC2 method uses a natural log transformation not a log2 transformation like other DAA methods [@Lin2024].
:::

### Differences in between intervention arms at endpoint

```{r}
#| eval: true
#| label: ancombc
#| fig-cap: NA
#| tbl-cap: NA

endArmANCOM <- ancombc2(
  endpoint,
  taxa_are_rows = TRUE,
  assay_name = "abundance",
  rank = tax_level,
  tax_level = NULL,
  fix_formula = "arm",
  rand_formula = NULL,
  p_adj_method = "BH",
  pseudo = 0,
  pseudo_sens = TRUE,
  prv_cut = 0.1,
  lib_cut = 0,
  s0_perc = 0.05,
  struc_zero = FALSE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 1,
  verbose = TRUE,
  global = FALSE,
  pairwise = FALSE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(tol = 0.01, max_iter = 20, verbose = FALSE),
  em_control = list(tol = 1e-05, max_iter = 100),
  lme_control = lme4::lmerControl(),
  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
  trend_control = list(contrast = NULL, node = NULL, solver = "ECOS", B = 100)
)

endArmANCOM_prim <- endArmANCOM$res
```

```{r}
#| eval: true
#| results: asis

save_and_link(
  data = endArmANCOM_prim,
  filename = paste0(prefix, "ancombc.csv"),
  description = "Endpoint ANCOM-BC results comparing DA OGUs between intervention arms",
  directory_naming = "both"
)
```

```{r}
#| eval: true
#| label: endArm-filtered
endArm <- endArmANCOM_prim %>%
  select(taxon, ends_with("_armFeZnPM")) %>%
  dplyr::filter(
    passed_ss_armFeZnPM == TRUE
  )
```

`r nrow(endArm)` passed ANCOM-BC sensitivity filtering.

```{r}
#| eval: true
#| label: tbl-summary
#| tbl-cap: ANCOM-BC significant (FDR) OGU fold changes. All OGUs passed sensitivity filtering.

endArm <- endArm %>%
  mutate(
    Result_FC1.2 = case_when(
      lfc_armFeZnPM < 0 &
        q_armFeZnPM <= 0.05 &
        abs(lfc_armFeZnPM) >= log(1.2) ~
        "Decrease",
      lfc_armFeZnPM > 0 &
        q_armFeZnPM <= 0.05 &
        abs(lfc_armFeZnPM) >= log(1.2) ~
        "Increase",
      TRUE ~ "No change"
    ),
    Result_FC2 = case_when(
      lfc_armFeZnPM < 0 &
        q_armFeZnPM <= 0.05 &
        abs(lfc_armFeZnPM) >= log(2) ~
        "Decrease",
      lfc_armFeZnPM > 0 &
        q_armFeZnPM <= 0.05 &
        abs(lfc_armFeZnPM) >= log(2) ~
        "Increase",
      TRUE ~ "No change"
    )
  ) %>%
  inner_join(rownames_to_column(tax_df(endpoint), "taxon"), by = "taxon")

colnames(endArm) <- str_remove(colnames(endArm), "_armFeZnPM")

kable(
  endArm %>%
    select(taxon, Result_FC1.2, Result_FC2) %>%
    pivot_longer(cols = c(Result_FC1.2, Result_FC2)) %>%
    group_by(name, value) %>%
    summarise(n = n())
)
```

### Significant OGUs with a Fold Change of 1.2 or more

```{r}
#| eval: true
#| label: fig-baseArm
#| fig-cap: Differentially abundant taxa in the intervention arm, FeZnPM, compared to control (CPM) at Endpoint. Significant taxa were determined to be so if they had a FDR adjusted p-value $\leq$ 0.05 and a fold change of 1.2 or more. They must also have passed ANCOM-BC sensitivity filtering.

plotDAvol(
  endArm,
  title = "Differentially abundant taxa in FeZnPM compared CPM",
  p.fdr = q,
  p.threshold = 0.05,
  lfc.threshold = log(1.2)
)
```

```{r}
#| eval: true
#| label: fig-significant-1.2
#| fig-cap: Significant OGUs with a Fold Change of 1.2 or more

phylum_anno <- assign_unique_colors(as.vector(endArm$Phylum), phyla_colours)

endArm %>%
  filter(Result_FC1.2 != "No change") %>%
  mutate(Species = str_remove_all(Species, "_AD|^Member of|family spp")) %>%

  arrange(lfc) %>%
  ggplot(aes(y = reorder(Species, lfc), x = lfc, fill = Phylum)) +
  geom_bar(position = position_dodge(), stat = "identity") +
  xlab(NULL) +
  ylab(NULL) + #labels the y axis
  scale_color_manual(values = phylum_anno, aesthetics = 'fill') +
  geom_vline(xintercept = 0) +
  theme(
    axis.text.x = element_text(
      angle = 0,
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    axis.text.y = element_text(size = 12, face = 'bold', color = 'black')
  )
```

### Significant OGUs with a Fold Change of 2 or more

```{r}
#| eval: true
#| label: fig-endArm
#| fig-cap: Differentially abundant taxa in the intervention arm, FeZnPM, compared to control (CPM) at Endpoint. Significant taxa were determined to be so if they had a FDR adjusted p-value $\leq$ 0.05 and a fold change of 2 or more. They must also have passed ANCOM-BC sensitivity filtering.

plotDAvol(
  endArm,
  title = "Differentially abundant taxa in FeZnPM compared CPM",
  p.fdr = q,
  p.threshold = 0.05,
  lfc.threshold = log(2)
)
```

```{r}
#| eval: true
#| label: fig-significant-2
#| fig-cap: Significant OGUs with a Fold Change of 2 or more

endArm %>%
  filter(Result_FC2 != "No change") %>%
  mutate(Species = str_remove_all(Species, "_AD|^Member of|family spp")) %>%
  arrange(lfc) %>%
  ggplot(aes(y = reorder(Species, lfc), x = lfc, fill = Phylum)) +
  geom_bar(position = position_dodge(), stat = "identity") +
  xlab(NULL) +
  ylab(NULL) + #labels the y axis
  scale_color_manual(values = phylum_anno, aesthetics = 'fill') +
  geom_vline(xintercept = 0) +
  theme(
    axis.text.x = element_text(
      angle = 0,
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    axis.text.y = element_text(size = 12, face = 'bold', color = 'black')
  )
```

## Results Summary

```{r}
#| eval: true
#| echo: false
#| results: asis

list_all_results()
```

```{r}
#| results: asis
generate_results_section(directory_naming = "both")
```

## Appendix

```{r}
#| eval: true
#| label: info

sessionInfo()
```
