## Rate of Change of Differentially Abundant OGUs in FeZnPM at Endpoint with ANCOM-BC

```{r}
#| echo: false
#| eval: true
#| label: setup
#| cache: false
suppressWarnings(suppressMessages({
  source(normalizePath("utils/r-utils.r"))
  source(normalizePath("_environment.required"))
  load(normalizePath("data/seqdata.RData"))

  metadata <- read_delim(normalizePath(METADATA))
  bac_base_end_meta <- readRDS(normalizePath(
    "data/metadata/base-end-metadata.rds"
  ))

  library(promicrobialTools)
  library(ANCOMBC)

  prefix <- "EvsBarm-DA-"
  filepath <- get_results_dir(directory_naming = "both")
}))
```

Results will be saved to ``r filepath``.

```{r}
#| eval: true
#| label: data-prep

phyloseq_summary(absAbund)

check <- data_sanity_check(absAbund, fix_formula = "arm * timepoint")
```

::: {.callout-important}
The `ancombc2` function is needed for repeated measures (can be used for cross-sectional too and can accommodate continuous covariates). `ancombc` can only be used for cross-sectional data.
:::

::: {.callout-warning}
The ANOCOM-BC2 method uses a natural log transformation not a log2 transformation like other DAA methods [@Lin2025].
:::

### Cross-sectional comparisons

```{r}
#| eval: true
base <- subset_samples(absAbund, timepoint == "Baseline")
end <- subset_samples(absAbund, timepoint == "Endpoint")
```

#### Baseline arm differences

```{r}
#| eval: true
#| label: ancombc
#| fig-cap: NA
#| tbl-cap: NA

base_results <- ancombc2(
  base,
  taxa_are_rows = TRUE,
  assay_name = "abundance",
  rank = tax_level,
  tax_level = NULL,
  fix_formula = "arm",
  p_adj_method = "BH",
  pseudo = 0,
  pseudo_sens = TRUE,
  prv_cut = 0.1,
  lib_cut = 0,
  s0_perc = 0.05,
  struc_zero = FALSE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 1,
  verbose = TRUE,
  global = FALSE,
  pairwise = FALSE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(tol = 0.01, max_iter = 20, verbose = FALSE),
  em_control = list(tol = 1e-05, max_iter = 100),
  lme_control = lme4::lmerControl(),
  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
  trend_control = list(contrast = NULL, node = NULL, solver = "ECOS", B = 100)
)

base_prim <- base_results$res
```

```{r}
#| eval: true
#| results: asis
save_and_link(
  data = base_prim,
  filename = paste0(prefix, "baseline-ancombc.csv"),
  description = "ANCOM-BC results comparing DA OGUS between intervention arms at Baseline in paired samples only",
  directory_naming = "both"
)
```

```{r}
#| eval: true
#| label: fig-base
#| fig-cap: Differentially abundant taxa in the FeZnPM intervention group at Baseline. Significant taxa were determined to be so if they had a FDR adjusted p-value $\leq$ 0.05 and a fold change of 2 or more

baseFiltered <- base_prim %>%
  select(taxon, ends_with("_armFeZnPM")) %>%
  dplyr::filter(
    diff_armFeZnPM == TRUE & passed_ss_armFeZnPM == TRUE
  )

colnames(baseFiltered) <- str_remove(colnames(baseFiltered), "_armFeZnPM")

plotDAvol(
  baseFiltered,
  title = "Differentially abundant OGUs in the FeZnPM intervention group at Baseline",
  p.fdr = q
)

```

#### Endpoint arm differences

```{r}
#| eval: true
#| label: ancombc
#| fig-cap: NA
#| tbl-cap: NA

end_results <- ancombc2(
  end,
  taxa_are_rows = TRUE,
  assay_name = "abundance",
  rank = tax_level,
  tax_level = NULL,
  fix_formula = "arm",
  p_adj_method = "BH",
  pseudo = 0,
  pseudo_sens = TRUE,
  prv_cut = 0.1,
  lib_cut = 0,
  s0_perc = 0.05,
  struc_zero = FALSE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 1,
  verbose = TRUE,
  global = FALSE,
  pairwise = FALSE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(tol = 0.01, max_iter = 20, verbose = FALSE),
  em_control = list(tol = 1e-05, max_iter = 100),
  lme_control = lme4::lmerControl(),
  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
  trend_control = list(contrast = NULL, node = NULL, solver = "ECOS", B = 100)
)

end_prim <- end_results$res
```

```{r}
#| eval: true
#| results: asis

save_and_link(
  data = end_prim,
  filename = paste0(prefix, "endpoint-ancombc.csv"),
  description = "ANCOM-BC results comparing DA OGUS between intervention arms at Endpoint in paired samples only",
  directory_naming = "both"
)
```

```{r}
#| eval: true
#| label: endFiltered

endFiltered <- end_prim %>%
  select(taxon, ends_with("_armFeZnPM")) %>%
  dplyr::filter(
    passed_ss_armFeZnPM == TRUE
  )
```

`r nrow(endFiltered)` passed ANCOM-BC sensitivity filtering.

```{r}
#| eval: true
#| label: tbl-end-summary
#| tbl-cap: ANCOM-BC significant (FDR) OGU fold changes. All OGUs passed sensitivity filtering.

endFiltered <- endFiltered %>%
  mutate(
    Result_FC1.2 = case_when(
      lfc_armFeZnPM < 0 &
        q_armFeZnPM <= 0.05 &
        abs(lfc_armFeZnPM) >= log(1.2) ~
        "Decrease",
      lfc_armFeZnPM > 0 &
        q_armFeZnPM <= 0.05 &
        abs(lfc_armFeZnPM) >= log(1.2) ~
        "Increase",
      TRUE ~ "No change"
    ),
    Result_FC2 = case_when(
      lfc_armFeZnPM < 0 &
        q_armFeZnPM <= 0.05 &
        abs(lfc_armFeZnPM) >= log(2) ~
        "Decrease",
      lfc_armFeZnPM > 0 &
        q_armFeZnPM <= 0.05 &
        abs(lfc_armFeZnPM) >= log(2) ~
        "Increase",
      TRUE ~ "No change"
    )
  ) %>%
  inner_join(rownames_to_column(tax_df(end), "taxon"), by = "taxon")

colnames(endFiltered) <- str_remove(colnames(endFiltered), "_armFeZnPM")

kable(
  endFiltered %>%
    select(taxon, Result_FC1.2, Result_FC2) %>%
    pivot_longer(cols = c(Result_FC1.2, Result_FC2)) %>%
    group_by(name, value) %>%
    summarise(n = n())
)
```

##### Significant OGUs with a Fold Change of 1.2 or more

```{r}
#| eval: true
#| label: fig-endArm
#| fig-cap: Differentially abundant taxa in the FeZnPM intervention group at Endpoint. Significant taxa were determined to be so if they had a FDR adjusted p-value $\leq$ 0.05 and a fold change of 1.2 or more. They must also have passed ANCOM-BC sensitivity filtering.

plotDAvol(
  endFiltered,
  title = "Differentially abundant OGUs in the FeZnPM intervention group at Endpoint compared to CPM",
  p.fdr = q,
  p.threshold = 0.05,
  lfc.threshold = log(1.2)
)
```

```{r}
#| eval: true
#| label: fig-sig-endOGUs-1.2
#| fig-cap: Significant OGUs with a Fold Change of 1.2 or more

phylum_anno <- assign_unique_colors(
  as.vector(endFiltered$Phylum),
  phyla_colours
)

endFiltered %>%
  filter(Result_FC1.2 != "No change") %>%
  arrange(lfc) %>%
  ggplot(aes(y = reorder(Species, lfc), x = lfc, fill = Phylum)) +
  geom_bar(position = position_dodge(), stat = "identity") +
  xlab(NULL) +
  ylab(NULL) + #labels the y axis
  scale_color_manual(values = phylum_anno, aesthetics = 'fill') +
  geom_vline(xintercept = 0) +
  theme(
    axis.text.x = element_text(
      angle = 0,
      size = fontsize,
      face = 'bold',
      color = 'black'
    )
  )
```

##### Significant OGUs with a Fold Change of 2 or more

```{r}
#| eval: true
#| label: fig-endArm-2
#| fig-cap: Differentially abundant taxa in the intervention arm, FeZnPM, compared to control (CPM) at Endpoint. Significant taxa were determined to be so if they had a FDR adjusted p-value $\leq$ 0.05 and a fold change of 2 or more. They must also have passed ANCOM-BC sensitivity filtering.

plotDAvol(
  endFiltered,
  title = "Differentially abundant taxa in FeZnPM compared CPM",
  p.fdr = q,
  p.threshold = 0.05,
  lfc.threshold = log(2)
)
```

```{r}
#| eval: true
#| label: fig-sig-endOGUs-2
#| fig-cap: Significant OGUs with a Fold Change of 2 or more

endFiltered %>%
  filter(Result_FC2 != "No change") %>%
  mutate(Species = str_remove_all(Species, "_AD|^Member of|family spp")) %>%
  arrange(lfc) %>%
  ggplot(aes(y = reorder(Species, lfc), x = lfc, fill = Phylum)) +
  geom_bar(position = position_dodge(), stat = "identity") +
  xlab(NULL) +
  ylab(NULL) + #labels the y axis
  scale_color_manual(values = phylum_anno, aesthetics = 'fill') +
  geom_vline(xintercept = 0) +
  theme(
    axis.text.x = element_text(
      angle = 0,
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    axis.text.y = element_text(size = 12, face = 'bold', color = 'black')
  )
```

### Differences in rates of change overtime between intervention arms

```{r}
#| eval: true
#| label: ancombc

EvsBarmANCOM <- ancombc2(
  absAbund,
  taxa_are_rows = TRUE,
  assay_name = "abundance",
  rank = tax_level,
  tax_level = NULL,
  fix_formula = "timepoint * arm",
  p_adj_method = "BH",
  pseudo = 0,
  pseudo_sens = TRUE,
  prv_cut = 0.1,
  lib_cut = 0,
  s0_perc = 0.05,
  group = "arm",
  struc_zero = TRUE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 1,
  verbose = TRUE,
  global = FALSE,
  pairwise = FALSE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(tol = 0.01, max_iter = 20, verbose = FALSE),
  em_control = list(tol = 1e-05, max_iter = 100),
  lme_control = lme4::lmerControl(),
  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
  trend_control = list(contrast = NULL, node = NULL, solver = "ECOS", B = 100)
)

EvsBarmANCOM_prim <- EvsBarmANCOM$res
```


```{r}
#| eval: true
#| results: asis
EvsBarmANCOM_prim <- EvsBarmANCOM_prim %>%
  mutate(
    FC = exp(abs(EvsBarmANCOM_prim$`lfc_timepointEndpoint:armFeZnPM`))
  ) %>%
  mutate(
    Result_FC1.2 = case_when(
      `lfc_timepointEndpoint:armFeZnPM` < 0 &
        `q_timepointEndpoint:armFeZnPM` <= 0.05 &
        abs(`lfc_timepointEndpoint:armFeZnPM`) >= log(1.2) ~
        "Decrease",
      `lfc_timepointEndpoint:armFeZnPM` > 0 &
        `q_timepointEndpoint:armFeZnPM` <= 0.05 &
        abs(`lfc_timepointEndpoint:armFeZnPM`) >= log(1.2) ~
        "Increase",
      TRUE ~ "No change"
    ),
    Result_FC2 = case_when(
      `lfc_timepointEndpoint:armFeZnPM` < 0 &
        `q_timepointEndpoint:armFeZnPM` <= 0.05 &
        abs(`lfc_timepointEndpoint:armFeZnPM`) >= log(2) ~
        "Decrease",
      `lfc_timepointEndpoint:armFeZnPM` > 0 &
        `q_timepointEndpoint:armFeZnPM` <= 0.05 &
        abs(`lfc_timepointEndpoint:armFeZnPM`) >= log(2) ~
        "Increase",
      TRUE ~ "No change"
    )
  ) %>%
  inner_join(rownames_to_column(tax_df(absAbund), "taxon"), by = "taxon")

save_and_link(
  data = EvsBarmANCOM_prim,
  filename = paste0(prefix, "ancombc.csv"),
  description = "ANCOM-BC results comparing DA OGUS between intervention arms over time (rate of change)",
  directory_naming = "both"
)

```

```{r}
#| eval: true
#| label: comm-formatted
tmp <- EvsBarmANCOM_prim %>%
  select(
    OGU = taxon,
    Phylum,
    Family,
    Genus,
    Species,
    `Loge Fold Change (ANCOM-BC)` = `lfc_timepointEndpoint:armFeZnPM`,
    `Fold Change (expFC)` = FC,
    `ANCOM-BC P-value` = `p_timepointEndpoint:armFeZnPM`,
    `ANCOM-BC FDR adjusted p-value` = `q_timepointEndpoint:armFeZnPM`,
    `Significant Result Fold Change ≥ 1.2` = Result_FC1.2,
    `Significant Result Fold Change ≥ 2` = Result_FC2
  )
```

```{r}
#| eval: true
#| label: save-formatted
#| results: asis

save_and_link(
  data = tmp,
  filename = paste0(prefix, "ancombc-formatted.csv"),
  description = "ANCOM-BC results comparing DA OGUS between intervention arms over time (rate of change; formatted for publication)",
  directory_naming = "both"
)
```

```{r}
#| eval: true
#| label: EvsBarm-filtered

EvsBarmANCOMFiltered <- EvsBarmANCOM_prim %>%
  dplyr::filter(
    `passed_ss_timepointEndpoint:armFeZnPM` == TRUE
  )

colnames(EvsBarmANCOMFiltered) <- str_remove(
  colnames(EvsBarmANCOMFiltered),
  "_armFeZnPM"
)

```

`r nrow(EvsBarmANCOMFiltered)` passed ANCOM-BC sensitivity filtering.

```{r}
#| eval: true
#| label: tbl-summary
#| tbl-cap: ANCOM-BC significant (FDR) OGU fold changes. All OGUs passed sensitivity filtering.
kable(
  EvsBarmANCOMFiltered %>%
    select(taxon, Result_FC1.2, Result_FC2) %>%
    pivot_longer(cols = c(Result_FC1.2, Result_FC2)) %>%
    group_by(name, value) %>%
    summarise(n = n())
)
```

::: {.callout-tip}

Coefficient Interpretations:

1. `lfc_(Intercept)`
   - Represents the "starting point" for all comparisons
   - Reference group at reference timepoint
   - This is the baseline abundance for the CPM arm at Baseline
2. `lfc_armFeZnPM`
   - Main effect of treatment arm
   - Difference between FeZnPM vs CPM at the reference timepoint only (Baseline)
   - This is the treatment effect at baseline (before the timepoint change occurs)
   - Interpretation: "How different is FeZnPM from CPM at the starting timepoint?"
3. `lfc_timepointEndpoint`
  - Main effect of time
  - Difference between endpoint vs reference timepoint in the reference arm only
  - This is the time effect in the control group
  - Interpretation: "How much does the reference arm change from start to endpoint?"
4. `lfc_armFeZnPM:timepointEndpoint`
  - Interaction effect (most important for intervention studies)
  - Additional change in FeZnPM arm at endpoint, beyond what happens in the reference arm
  - This is the differential treatment effect - how the treatment response differs over time
  - Interpretation: "Does the FeZnPM treatment have a different effect at endpoint compared to the reference arm's change over time?"

:::

### Significant OGUs with a Fold Change of 1.2 or more

```{r}
#| eval: true
#| label: fig-EvsBArm
#| fig-cap: Impact of FeZnPM over time compared to CPM. Differentially abundant taxa in the intervention arm, FeZnPM, compared to control (CPM) at Endpoint. This is the differential treatment effect i.e. the additional change in OGU abundances in the FeZnPM arm at endpoint, beyond what happens in the CPM reference arm. Significant taxa were determined to be so if they had a FDR adjusted p-value $\leq$ 0.05 and a fold change of 1.2 or more. They must also have passed ANCOM-BC sensitivity filtering.

plotDAvol(
  EvsBarmANCOMFiltered,
  title = "OGUs that change at a significantly different rate overtime in the FeZnPM intervention group compared to the rate of change in the CPM group",
  p.fdr = q,
  p.threshold = 0.05,
  lfc.threshold = log(1.2)
)
```

```{r}
#| eval: true
#| label: fig-significant-1.2
#| fig-cap: Significant OGUs with a Fold Change of 1.2 or more

phylum_anno <- assign_unique_colors(
  as.vector(EvsBarmANCOMFiltered$Phylum),
  phyla_colours
)

EvsBarmANCOMFiltered %>%
  filter(Result_FC1.2 != "No change") %>%
  mutate(Species = str_remove_all(Species, "_AD|^Member of|family spp")) %>%
  arrange(lfc) %>%
  ggplot(aes(y = reorder(Species, lfc), x = lfc, fill = Phylum)) +
  geom_bar(position = position_dodge(), stat = "identity") +
  xlab(NULL) +
  ylab(NULL) + #labels the y axis
  scale_color_manual(values = phylum_anno, aesthetics = 'fill') +
  geom_vline(xintercept = 0) +
  theme(
    axis.text.x = element_text(
      angle = 0,
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    axis.text.y = element_text(size = 12, face = 'bold', color = 'black')
  )
```

### Significant OGUs with a Fold Change of 2 or more

```{r}
#| eval: true
#| label: fig-endArm
#| fig-cap: Differentially abundant taxa in the intervention arm, FeZnPM, compared to control (CPM) at Endpoint. Significant taxa were determined to be so if they had a FDR adjusted p-value $\leq$ 0.05 and a fold change of 2 or more. They must also have passed ANCOM-BC sensitivity filtering.

plotDAvol(
  EvsBarmANCOMFiltered,
  title = "Differentially abundant taxa in FeZnPM compared CPM",
  p.fdr = q,
  p.threshold = 0.05,
  lfc.threshold = log(2)
)
```

```{r}
#| eval: true
#| label: fig-significant-2
#| fig-cap: Significant OGUs with a Fold Change of 2 or more

EvsBarmANCOMFiltered %>%
  filter(Result_FC2 != "No change") %>%
  mutate(Species = str_remove_all(Species, "_AD|^Member of|family spp")) %>%
  arrange(lfc) %>%
  ggplot(aes(y = reorder(Species, lfc), x = lfc, fill = Phylum)) +
  geom_bar(position = position_dodge(), stat = "identity") +
  xlab(NULL) +
  ylab(NULL) + #labels the y axis
  scale_color_manual(values = phylum_anno, aesthetics = 'fill') +
  geom_vline(xintercept = 0) +
  theme(
    axis.text.x = element_text(
      angle = 0,
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    axis.text.y = element_text(size = 12, face = 'bold', color = 'black')
  )
```

## Results Summary

```{r}
#| eval: true
#| echo: false
#| results: asis

list_all_results()
```

```{r}
#| eval: true
#| echo: false
#| results: asis
generate_results_section(directory_naming = "both")
```

## Appendix

```{r}
#| eval: true
#| echo: false
#| label: info

print(sessionInfo(), locale = FALSE)
```

