# Longitudional Impact of Intervention on Differential Abundance

## OGUs Common between GLMMs and differential abundance changes over time

```{r setup}
#| echo: false
#| eval: true
#| label: setup

# This initial code block loads custom utility functions and environment variables
# These are stored in separate files to keep the main analysis clean and organized
suppressWarnings(suppressMessages({
  source("utils/r-utils.r")

  source("_environment.required")
  load("data/seqdata.RData")

  library(promicrobialTools)

  prefix <- "EvsBarmarm-"

  filepath <- get_results_dir(directory_naming = "both")
}))
```

Results will be saved to ``r filepath``.

## Comparision of ANCOM-BC Differentially Abundant OGUs and ZI-GLMM model results

```{r}
#| eval: true
#| label: FC2 + GLMM result comparisons

EvsBarm_ZI <- read.csv(
  file.path(
    RESULTS,
    "06a-zero-inflated-OGU-models/EvsBarm-OGU-zero-inflated-models",
    "EvsBarm-zero-inflated-ogu-fitted-model-significant-pvalues.csv"
  ),
  header = TRUE
)

EvsBarm_ANCOM <- read.csv(file.path(
  RESULTS,
  "07-differential-abundance-analysis/EvsBarm-ANCOM-BC/EvsBarm-DA-ancombc.csv"
))
```

```{r}
#| eval: true
#| label: FC2
#| fig-cap: NA
#| tbl-cap: NA

EvsBarm_ANCOM <- EvsBarm_ANCOM %>%
  mutate(FC = exp(abs(EvsBarm_ANCOM$lfc_timepointEndpoint))) %>%
  filter(passed_ss_timepointEndpoint == TRUE, q_timepointEndpoint <= 0.05) %>%
  select(
    taxon,
    lfc_timepointEndpoint,
    FC,
    p_timepointEndpoint,
    q_timepointEndpoint,
    Result_FC1.2,
    Result_FC2
  )

common <- EvsBarm_ZI %>%
  inner_join(., EvsBarm_ANCOM, by = c("outcome_id" = "taxon"))
```

```{r}
#| eval: true
#| label: save-common
#| results: asis

save_and_link(
  data = common,
  filename = paste0(prefix, "ANCOM-DA-OGUs-ZI-GLMM-common.csv"),
  description = "OGUs common between ANCOM-BC differential abundance analysis and ZI-GLMMs",
  directory_naming = "both"
)
```

```{r}
#| eval: true
#| label:
tmp <- common %>%
  select(
    OGU = outcome_id,
    Phylum,
    Family,
    Genus,
    Species,
    `ZI-GLMM Time P-value` = time,
    `ZI-GLMM Age at Baseline P-value` = age_B,
    `ZI-GLMM Interaction P-value` = time.age_B,
    `Loge Fold Change (ANCOM-BC)` = lfc_timepointEndpoint,
    `Fold Change (expFC)` = FC,
    `ANCOM-BC P-value` = p_timepointEndpoint,
    `ANCOM-BC FDR adjusted p-value` = q_timepointEndpoint,
    `Significant Result Fold Change ≥ 1.2` = Result_FC1.2,
    `Significant Result Fold Change ≥ 2` = Result_FC2
  )
```

```{r}
#| eval: true
#| label: save-formatted
#| results: asis

save_and_link(
  data = common,
  filename = paste0(prefix, "ANCOM-DA-OGUs-ZI-GLMM-common-formatted.csv"),
  description = "OGUs common between ANCOM-BC differential abundance analysis and ZI-GLMMs (formatted for publication)",
  directory_naming = "both"
)
```

```{r}
#| eval: true
#| label: tbl-com
#| tbl-cap: Total number of OGUs in each phylum

common %>%
  group_by(Phylum) %>%
  mutate(count = 1) %>%
  summarise(sum = sum(count))
```

### Common OGUs with FC 1.2 or more

```{r}
#| eval: true
#| label: common1.2
#| fig-cap: NA
#| tbl-cap: NA
kable(head(
  common1.2 <- common %>%
    filter(Result_FC1.2 != "No change")
))
```

There are `r nrow(common1.2)` OGUs with a fold change of 1.2 or more (`r length(unique(common1.2$Species))` unique species and `r length(unique(common1.2$Genus))` unique genera) are significantly changing over time between zero-inflated GLMM and DA results i.e. are common between the two methods.

```{r}
#| eval: true
#| label: hm-prep
#| fig-cap: NA

common_mat <- common1.2 %>%
  mutate(Species = ifelse(Species == "", outcome_id, Species)) %>%
  select(
    Phylum,
    Species,
    logFC = lfc_timepointEndpoint,
    FC,
    q_timepointEndpoint
  )

phylum_anno <- assign_unique_colors(as.vector(common_mat$Phylum), phyla_colours)
```

```{r}
#| eval: true
#| label: fig-com
#| fig-cap: Log2 fold change of OGUs shared across LMM and DA results
common_mat %>%
  arrange(logFC) %>%
  ggplot(aes(y = reorder(Species, logFC), x = logFC, fill = Phylum)) +
  geom_bar(position = position_dodge(), stat = "identity") +
  xlab(NULL) +
  ylab(NULL) + #labels the y axis
  scale_color_manual(values = phylum_anno, aesthetics = 'fill') +
  geom_vline(xintercept = 0) +
  theme(
    axis.text.x = element_text(
      angle = 0,
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    axis.text.y = element_text(size = 12, face = 'bold', color = 'black')
  )
```

### Common OGUs with FC 2 or more

```{r}
#| eval: true
#| label: common-sig
kable(head(
  common2 <- common %>%
    filter(Result_FC2 != "No change")
))

```

There are `r nrow(common2)` OGUs with a fold change of 2 or more (`r length(unique(common2$Species))` unique species and `r length(unique(common2$Genus))` unique genera) are significantly changing over time between zero-inflated GLMM and DA results i.e. are common between the two methods.

```{r}
#| eval: true
#| label: hm-prep
#| fig-cap: NA

common_mat <- common2 %>%
  mutate(Species = ifelse(Species == "", outcome_id, Species)) %>%
  select(
    Phylum,
    Species,
    logFC = lfc_timepointEndpoint,
    FC,
    q_timepointEndpoint
  )

phylum_anno <- assign_unique_colors(as.vector(common_mat$Phylum), phyla_colours)
```

```{r}
#| eval: true
#| label: fig-com
#| fig-cap: Log2 fold change of OGUs shared across LMM and DA results
common_mat %>%
  arrange(logFC) %>%
  ggplot(aes(y = reorder(Species, logFC), x = logFC, fill = Phylum)) +
  geom_bar(position = position_dodge(), stat = "identity") +
  xlab(NULL) +
  ylab(NULL) + #labels the y axis
  scale_color_manual(values = phylum_anno, aesthetics = 'fill') +
  geom_vline(xintercept = 0) +
  theme(
    axis.text.x = element_text(
      angle = 0,
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    axis.text.y = element_text(size = 12, face = 'bold', color = 'black')
  )
```

## Comparision of Differentially Abundant Species and GLMM Estimated Marginal Mean results

```{r}
#| eval: true
#| label: FC2 + GLMM result comparisons

emms <- read.csv(
  file.path(
    RESULTS,
    "06a-zero-inflated-OGU-models/EvsBarm-OGU-zero-inflated-models",
    "EvsBarm-significant-ogu-emms-time-ZI.csv"
  ),
  header = TRUE
) %>%
  rename(outcome_id = Model)

```

```{r}
#| eval: true
#| label: FC2
#| fig-cap: NA
#| tbl-cap: NA

commonEMM <- emms %>%
  inner_join(., EvsBarm_ANCOM, by = c("outcome_id" = "taxon"))
```

```{r}
#| eval: true
#| label: save
#| results: asis
#|
save_and_link(
  data = commonEMM,
  filename = paste0(prefix, "ANCOM-DA-OGUs-GLMM-EMMs-common.csv"),
  description = "OGUs common between ANCOM-BC differential abundance analysis and significant OGUs determined from ZI-GLMMs estimated marginal means",
  directory_naming = "both"
)
```

```{r}
#| eval: true
tmp <- commonEMM %>%
  select(
    OGU = outcome_id,
    Phylum,
    Family,
    Genus,
    Species,
    `ZI-GLMM EMM Estimate` = estimate,
    `ZI-GLMM EMM P-value (FDR ≤)` = p.value,
    `ZI-GLMM EMM Effect Size` = effect.size,
    `ZI-GLMM EMM Magnitude` = magnitude,
    `Loge Fold Change (ANCOM-BC)` = lfc_timepointEndpoint,
    `Fold Change (expFC)` = FC,
    `ZicoSeq P-value` = p_timepointEndpoint,
    `ZicoSeq FDR adjusted p-value` = q_timepointEndpoint,
    `Significant Result Fold Change ≥ 1.2` = Result_FC1.2,
    `Significant Result Fold Change ≥ 2` = Result_FC2
  )
```

```{r}
#| eval: true
#| results: asis

save_and_link(
  data = tmp,
  filename = paste0(prefix, "ANCOM-DA-OGUs-GLMM-EMMs-common-formatted.csv"),
  description = "OGUs common between ANCOM-BC differential abundance analysis and significant OGUs determined from ZI-GLMMs estimated marginal means (formatted for publication)",
  directory_naming = "both"
)
```

```{r}
#| eval: true
#| label: tbl-com
#| fig-cap: NA
#| tbl-cap: total number of species in each phylum

commonEMM %>%
  group_by(Phylum) %>%
  mutate(count = 1) %>%
  summarise(sum = sum(count))
```

### Common OGUs with FC 1.2 or more

```{r}
#| eval: true
#| label: commonEMM1.2
#| fig-cap: NA
#| tbl-cap: NA
kable(head(
  commonEMM1.2 <- commonEMM %>%
    filter(Result_FC1.2 != "No change")
))
```

There are `r nrow(commonEMMEMM1.2)` OGUs with a fold change of 1.2 or more (`r length(unique(commonEMM1.2$Species))` unique species and `r length(unique(commonEMM1.2$Genus))` unique genera) are significantly changing over time between zero-inflated GLMM and DA results i.e. are commonEMM between the two methods.

```{r}
#| eval: true
#| label: hm-prep
#| fig-cap: NA

commonEMM_mat <- commonEMM1.2 %>%
  mutate(Species = ifelse(Species == "", outcome_id, Species)) %>%
  select(
    Phylum,
    Species,
    logFC = lfc_timepointEndpoint,
    FC,
    q_timepointEndpoint
  )

phylum_anno <- assign_unique_colors(
  as.vector(commonEMM_mat$Phylum),
  phyla_colours
)
```

```{r}
#| eval: true
#| label: fig-com
#| fig-cap: Log2 fold change of OGUs shared across LMM and DA results
commonEMM_mat %>%
  arrange(logFC) %>%
  ggplot(aes(y = reorder(Species, logFC), x = logFC, fill = Phylum)) +
  geom_bar(position = position_dodge(), stat = "identity") +
  xlab(NULL) +
  ylab(NULL) + #labels the y axis
  scale_color_manual(values = phylum_anno, aesthetics = 'fill') +
  geom_vline(xintercept = 0) +
  theme(
    axis.text.x = element_text(
      angle = 0,
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    axis.text.y = element_text(size = 12, face = 'bold', color = 'black')
  )
```

### Common OGUs with FC 2 or more

```{r}
#| eval: true
#| label: commonEMM-sig
kable(head(
  commonEMM2 <- commonEMM %>%
    filter(Result_FC2 != "No change")
))

```

There are `r nrow(commonEMM2)` OGUs with a fold change of 2 or more (`r length(unique(commonEMM2$Species))` unique species and `r length(unique(commonEMM2$Genus))` unique genera) are significantly changing over time between zero-inflated GLMM and DA results i.e. are commonEMM between the two methods.

```{r}
#| eval: true
#| label: hm-prep
#| fig-cap: NA

commonEMM_mat <- commonEMM2 %>%
  mutate(Species = ifelse(Species == "", outcome_id, Species)) %>%
  select(
    Phylum,
    Species,
    logFC = lfc_timepointEndpoint,
    FC,
    q_timepointEndpoint
  )

phylum_anno <- assign_unique_colors(
  as.vector(commonEMM_mat$Phylum),
  phyla_colours
)
```

```{r}
#| eval: true
#| label: fig-com
#| fig-cap: Log2 fold change of OGUs shared across LMM and DA results
commonEMM_mat %>%
  arrange(logFC) %>%
  ggplot(aes(y = reorder(Species, logFC), x = logFC, fill = Phylum)) +
  geom_bar(position = position_dodge(), stat = "identity") +
  xlab(NULL) +
  ylab(NULL) + #labels the y axis
  scale_color_manual(values = phylum_anno, aesthetics = 'fill') +
  geom_vline(xintercept = 0) +
  theme(
    axis.text.x = element_text(
      angle = 0,
      size = fontsize,
      face = 'bold',
      color = 'black'
    ),
    axis.text.y = element_text(size = 12, face = 'bold', color = 'black')
  )
```