## Differential abundance analysis with ANCOM-BC

```{r}
#| echo: false
#| eval: true
#| label: setup
#| cache : false
suppressWarnings(suppressMessages({
  source(normalizePath("utils/r-utils.r"))
  source(normalizePath("_environment.required"))
  load(normalizePath("data/seqdata.RData"))

  library(ANCOMBC)
  library(promicrobialTools)

  prefix <- "EvsB-DA-"
  filepath <- get_results_dir(directory_naming = "both")
}))
```

Results will be saved to ``r filepath``.

```{r}
#| eval: true
#| label: data-prep
#| message: false
#| results: asis
absAbund %>% phyloseq_summary()

check <- data_sanity_check(absAbund, fix_formula = "timepoint")
```

::: {.callout-important}
The `ancombc2` function is needed for repeated measures (can be used for cross-sectional too and can accommodate continuous covariates). `ancombc` can only be used for cross-sectional data.
:::

::: {.callout-warning}
The ANOCOM-BC2 method uses a natural log transformation not a log2 transformation like other DAA methods [@Lin2024].
:::

### Change in abundances overtime (ignoring arm)

```{r}
#| eval: true
#| label: ancombc
#| fig-cap: NA
#| tbl-cap: NA

time_results <- ancombc2(
  absAbund,
  taxa_are_rows = TRUE,
  assay_name = "abundance",
  rank = tax_level,
  tax_level = NULL,
  fix_formula = "timepoint",
  p_adj_method = "BH",
  pseudo = 0,
  pseudo_sens = TRUE,
  prv_cut = 0.1,
  lib_cut = 0,
  s0_perc = 0.05,
  group = "timepoint",
  struc_zero = TRUE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 1,
  verbose = TRUE,
  global = FALSE,
  pairwise = FALSE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(tol = 0.01, max_iter = 20, verbose = FALSE),
  em_control = list(tol = 1e-05, max_iter = 100),
  lme_control = lme4::lmerControl(),
  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
  trend_control = list(contrast = NULL, node = NULL, solver = "ECOS", B = 100)
)

time_prim <- time_results$res
```

```{r}
#| eval: true
#| label: tbl-summary
#| tbl-cap: ANCOM-BC significant (FDR) OGU fold changes

time_prim <- time_prim %>%
  mutate(FC = exp(abs(time_prim$lfc_timepointEndpoint))) %>%
  mutate(
    Result_FC1.2 = case_when(
      lfc_timepointEndpoint < 0 &
        q_timepointEndpoint <= 0.05 &
        abs(lfc_timepointEndpoint) >= log(1.2) ~
        "Decrease",
      lfc_timepointEndpoint > 0 &
        q_timepointEndpoint <= 0.05 &
        abs(lfc_timepointEndpoint) >= log(1.2) ~
        "Increase",
      TRUE ~ "No change"
    ),
    Result_FC2 = case_when(
      lfc_timepointEndpoint < 0 &
        q_timepointEndpoint <= 0.05 &
        abs(lfc_timepointEndpoint) >= log(2) ~
        "Decrease",
      lfc_timepointEndpoint > 0 &
        q_timepointEndpoint <= 0.05 &
        abs(lfc_timepointEndpoint) >= log(2) ~
        "Increase",
      TRUE ~ "No change"
    )
  ) %>%
  left_join(rownames_to_column(tax_df(absAbund), "taxon"), by = "taxon")
```

```{r}
#| eval: true
#| label: save-results
#| results: asis
save_and_link(
  data = time_prim,
  filename = paste0(prefix, "ancombc.csv"),
  description = "ANCOM-BC results comparing DA OGUs between Baseline and Endpoint (ignoring intervention arms)",
  directory_naming = "both"
)
```

```{r}
#| eval: true
#| label: comm-formatted
tmp <- time_prim %>%
  select(
    OGU = taxon,
    Phylum,
    Family,
    Genus,
    Species,
    `Loge Fold Change (ANCOM-BC)` = lfc_timepointEndpoint,
    `Fold Change (expFC)` = FC,
    `ANCOM-BC P-value` = p_timepointEndpoint,
    `ANCOM-BC FDR adjusted p-value` = q_timepointEndpoint,
    `Significant Result Fold Change ≥ 1.2` = Result_FC1.2,
    `Significant Result Fold Change ≥ 2` = Result_FC2
  )
```

```{r}
#| eval: true
#| label: save-formatted
#| results: asis

save_and_link(
  data = tmp,
  filename = paste0(prefix, "ancombc-formatted.csv"),
  description = "ANCOM-BC results comparing DA OGUs between Baseline and Endpoint (ignoring intervention arms; formatted for publication)",
  directory_naming = "both"
)
```

```{r}
#| eval: true
#| label: tbl-EvsB
#| tbl-cap: Differentially abundant taxa at Endpoint Compared to Baseline. Significant taxa were determined to be so if they had a FDR adjusted p-value $\leq$ 0.05 and a fold change of 2 or more

EvsB <- time_prim %>%
  dplyr::filter(
    passed_ss_timepointEndpoint == TRUE,
    q_timepointEndpoint <= 0.05
  )

colnames(EvsB) <- str_remove(colnames(EvsB), "_timepointEndpoint")

kable(
  EvsB %>%
    select(taxon, Result_FC1.2, Result_FC2) %>%
    pivot_longer(cols = c(Result_FC1.2, Result_FC2)) %>%
    group_by(name, value) %>%
    summarise(n = n())
)
```

```{r}
#| eval: true
#| label: ancom-subset

FC1.2 <- EvsB %>%
  filter(Result_FC1.2 != "No change")

FC2 <- EvsB %>%
  filter(Result_FC2 != "No change")
```

```{r}
#| eval: true
#| label: tbl-unq-fc1.2
#| tbl-cap: Number of unique taxa represented by significantly differentially abundant OGUs (FC 1.2)

kable(
  FC1.2 %>%
    select(OGU = taxon, Phylum, Genus, Species) %>%
    pivot_longer(
      cols = c(OGU, Phylum, Genus, Species),
      names_to = "Taxonomy"
    ) %>%
    group_by(Taxonomy) %>%
    summarise(`Number of Unique Taxa` = n_distinct(value))
)
```

```{r}
#| eval: true
#| label: tbl-unq-fc2
#| tbl-cap: Number of unique taxa represented by significantly differentially abundant OGUs (FC 2)

kable(
  FC2 %>%
    select(OGU = taxon, Phylum, Genus, Species) %>%
    pivot_longer(
      cols = c(OGU, Phylum, Genus, Species),
      names_to = "Taxonomy"
    ) %>%
    group_by(Taxonomy) %>%
    summarise(`Number of Unique Taxa` = n_distinct(value))
)
```

```{r}
#| eval: true
#| label: fig-baseArm
#| fig-cap: Differentially abundant taxa in the intervention arm, FeZnPM, compared to control (CPM). Significant taxa were determined to be so if they had a FDR adjusted p-value $\leq$ 0.05 and a fold change of 1.2 or more. They must also have passed ANCOM-BC sensitivity filtering.
plotDAvol(
  EvsB,
  title = "Differentially abundant taxa at Endpoint Compared to Baseline",
  p.fdr = q,
  p.threshold = 0.05,
  lfc.threshold = log(1.2)
)
```

```{r}
#| eval: true
#| label: fig-FC2
#| fig-cap: NA

plotDAvol(
  EvsB,
  title = "Differentially abundant taxa at Endpoint Compared to Baseline",
  p.fdr = q
)

```

## Results Summary

```{r}
#| eval: true
#| echo: false
#| results: asis

list_all_results()
```

```{r}
#| results: asis
generate_results_section(directory_naming = "both")
```

## Appendix

```{r}
#| eval: true
#| label: info

print(sessionInfo(), locale = FALSE)
```

