# Investigating differences in OGU abundances across Baseline samples due to Age using Zero-inflated Regression Models

```{r setup}
#| echo: false
#| eval: true
#| label: setup
#| cache: false

# This initial code block loads custom utility functions and environment variables
# These are stored in separate files to keep the main analysis clean and organized
suppressWarnings(suppressMessages({
  source("utils/r-utils.r")
  source("scripts/OGeeU.r")
  source("scripts/zeroGUS.r")

  source("_environment.required")
  load("data/seqdata.RData")

  library(glmmTMB)
  library(DHARMa)
  library(broom)
  library(patchwork)
  library(emmeans)
  library(ggtext)

  library(promicrobialTools)

  prefix <- "baseArm-ZI-"
  filepath <- get_results_dir(directory_naming = "both")
}))
```

Results will be saved to `r filepath`.

## Prepare Outcome and Predictor Variables

```{r data-prep}
#| eval: true
#| label: data prep
#| results: asis
baseline <- phyloseq::subset_samples(seqdata_bac, timepoint == "Baseline")

data <- prepare_ogu_data(
  baseline,
  filter_prevalence = TRUE,
  prevalence_cutoff = 0.4,
  filter_abundance = FALSE,
  rownames_as_samples = TRUE
)

outcome <- data$ogu_data
rownames(outcome) <- str_replace(
  str_replace(str_remove(rownames(outcome), "_G"), "_B_B", "_B"),
  "_B_E",
  "_E"
)

predictors <- prepare_predictors(data$meta_data)
predictors <- predictors %>%
  select(subject, trt, age_B) %>%
  rownames_to_column(var = "sample") %>%
  mutate(
    baseline_sample = str_replace(sample, "_G_E|_B_E|_E|_G_B|_B_B", "_B"),
    sample = str_remove(sample, "_G"),
  ) %>%
  mutate(sample = str_replace(sample, "_B_B", "_B")) %>%
  mutate(sample = str_replace(sample, "_B_E", "_E"))
```

## Testing different distributions

The following distribution families were used to fit a zero-inflated model:

-   Poisson
-   Simple negative binomial (linear parameterization) [@Hardin2018]
-   Complex negative binomial (quadratic parameterization) [@Hardin2018]
-   Gaussian (normal)
-   Log normal

```{r fit}
#| eval: false
#| label: fit-genZI
#| cache: true
nb2_baseAge <- genZI(
  outcome,
  predictors,
  formula = "age_B",
  zi = "~ age_B",
  family = "nbinom2"
)

nb1_baseAge <- genZI(
  outcome,
  predictors,
  formula = "age_B",
  zi = "~ age_B",
  family = "nbinom1"
)

poi_baseAge <- genZI(
  outcome,
  predictors,
  formula = "age_B",
  zi = "~ age_B",
  family = "poisson"
)

gau_baseAge <- genZI(
  outcome,
  predictors,
  formula = "age_B",
  zi = "~ age_B",
  family = "gaussian"
)

lognorm_baseAge <- genZI(
  outcome,
  predictors,
  formula = "age_B",
  zi = "~ age_B",
  family = "lognormal"
)
```

```{r save}
#| eval: false
#| echo: false
saveRDS(
  nb2_baseAge,
  file.path(filepath, "baseAge-genZI-OGU-nb2_results.rds")
)

saveRDS(
  nb1_baseAge,
  file.path(filepath, "baseAge-genZI-OGU-nb1_results.rds")
)

saveRDS(
  poi_baseAge,
  file.path(filepath, "baseAge-genZI-OGU-poi_results.rds")
)

saveRDS(
  gau_baseAge,
  file.path(filepath, "baseAge-genZI-OGU-gau_results.rds")
)

saveRDS(
  lognorm_baseAge,
  file.path(filepath, "baseAge-genZI-OGU-lognorm_results.rds")
)
```

```{r load}
#| eval: true
#| echo: false
#| cache: true
nb2_baseAge <- readRDS(file.path(
  filepath,
  "baseAge-genZI-OGU-nb2_results.rds"
))

nb1_baseAge <- readRDS(file.path(
  filepath,
  "baseAge-genZI-OGU-nb1_results.rds"
))

poi_baseAge <- readRDS(file.path(
  filepath,
  "baseAge-genZI-OGU-poi_results.rds"
))

gau_baseAge <- readRDS(file.path(
  filepath,
  "baseAge-genZI-OGU-gau_results.rds"
))

lognorm_baseAge <- readRDS(file.path(
  filepath,
  "baseAge-genZI-OGU-lognorm_results.rds"
))
```

### Model diagnostic failures

```{r}
#| eval: true
#| label: list

list <- list(
  nb1 = nb1_baseAge,
  nb2 = nb2_baseAge,
  poi = poi_baseAge,
  gau = gau_baseAge,
  lognorm = lognorm_baseAge
)
```

```{r}
#| eval: true
#| label: diagnose-mods
#| message: false
#| warning: false
#| cache: true

poor_fits <- lapply(1:length(list), function(x) {
  d <- diagnoseZI(list[[x]]$all_models)
  d$dist <- names(list)[x]

  return(d)
})

poor_fits <- data.table::rbindlist(poor_fits)
```

### Comparing Model AIC and BIC values

```{r}
#| eval: true
#| label: summaries

summaries <- list(
  gau = gau_baseAge$model_summary,
  lognorm = lognorm_baseAge$model_summary,
  poi = poi_baseAge$model_summary,
  nb1 = nb1_baseAge$model_summary,
  nb2 = nb2_baseAge$model_summary
)

summaries <- data.table::rbindlist(summaries, idcol = "dist")

summaries <- left_join(summaries, poor_fits, by = c("outcome_id", "dist"))
```

```{r}
#| eval: true
#| label: any-errors

if (any(summaries$error_message != "")) {
  print("Some models returned an error message")
  ind <- which(summaries$error_message != "")
  kable(summaries[ind, ])
}

```

```{r}
#| eval: true
#| label: diag-compare
summaries <- summaries %>%
  dplyr::filter(pass == TRUE) %>%
  select(!c(error_message, message)) %>%
  mutate(index = 1:nrow(.)) %>% #make it easier to distinguish between outcome ids
  pivot_longer(
    cols = !c(index, outcome_id, dist, pass),
    names_to = "criterion",
    values_to = "value"
  )
```

```{r}
#| eval: true
#| label: best-AIC

aic <- summaries %>%
  filter(criterion == "AIC") %>%
  group_by(criterion, outcome_id) %>%
  arrange(value) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, minAIC = value, distAIC = dist)
```

```{r}
#| eval: true
#| label: best-BIC

bic <- summaries %>%
  filter(criterion == "BIC") %>%
  group_by(criterion, outcome_id) %>%
  arrange(value) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, minBIC = value, distBIC = dist)

```

```{r}
#| eval: true
#| label: tbl-best-fits
#| tbl-cap: Best fit models based on AIC and BIC

bestDist <- left_join(aic, bic) %>%
  mutate(dist = ifelse(distAIC == distBIC, distAIC, "Undetermined"))
```

### Checking for deviations in model assumptions

::: callout-note
The larger the dispersion value, $\theta$ in `nbinom2` models, the less variance there is in the model [@Brooks2017].
:::

`dharmaDisp` represents the dispersion ratio, which should ideally be close to 1

```{r}
#| eval: true
#| label: tbl-best-disp
#| tbl-cap: lowest disp value for models that passed diagnostics

dispRatio <- summaries %>%
  filter(criterion == "dharmaDisp") %>%
  group_by(criterion, outcome_id) %>%
  mutate(skew = abs(value - 1)) %>%
  arrange(skew) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, bestDisp = value, distDisp = dist)
```

```{r}
#| eval: true
#| label: tbl-pvals
#| tbl-cap: NA

pvals <- left_join(
  bestDist,
  summaries,
  by = c("index", "outcome_id", "dist")
) %>%
  filter(str_detect(criterion, "Pvalue"), value <= 0.05) %>%
  arrange(index)
```

`r n_distinct(pvals$outcome_id)` models have a significant test result making the model fit problematic

### Final Model Distributions

```{r}
#| eval: true
#| label: best-dist-filtered
#| fig-cap: NA
#| tbl-cap: NA

bestDist_filtered <- bestDist %>%
  filter(!outcome_id %in% pvals$outcome_id)
```

## Fitting models to best distribution

### Zero-inflated Negative Binomial 1 Models

```{r}
#| eval: true
#| label: nbinom1 models

nb1_models <- keep_at(
  nb1_baseAge$all_models,
  filter(bestDist_filtered, dist == "nb1")$outcome_id
)

```

```{r}
#| eval: true
#| label: nb1-summary

if (exists("nb1_models") && length(nb1_models) > 0) {
  nb1_unfit <- diagnoseZI(nb1_models)
  nb1_summary <- data.frame(
    outcome_id = names(nb1_models),
    continuous_distribution = "nbinom1",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
  nb1_ogu_models <- nb1_baseAge$adj_results %>%
    filter(outcome_id %in% names(nb1_models))
}
```

Are there any models that fail diagnostics: `r if (exists("nb1_models") && length(nb1_models) > 0) {any(nb1_unfit$pass == FALSE)} else {paste("NA- no models generated by this fit")}`

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for nbinom1
if (exists("nb1_models") && length(nb1_models) > 0) {
  nb1_dharma_results <- process_dharma_results(nb1_models)

  kable(nb1_dharma_results$summary)
}

```

### Zero-inflated Negative Binomial Polynomial Models

```{r}
#| eval: true
#| label: nbinom2 models

nb2_models <- keep_at(
  nb2_baseAge$all_models,
  filter(bestDist_filtered, dist == "nb2")$outcome_id
)

```

```{r}
#| eval: true
#| label: nb2-summary
if (exists("nb2_models") && length(nb2_models) > 0) {
  nb2_unfit <- diagnoseZI(nb2_models)

  nb2_summary <- data.frame(
    outcome_id = names(nb2_models),
    continuous_distribution = "nbinom2",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
  nb2_ogu_models <- nb2_baseAge$adj_results %>%
    filter(outcome_id %in% names(nb2_models))
}
```

Are there any models that fail diagnostics: `r if (exists("nb2_models") && length(nb2_models) > 0) {any(nb2_unfit$pass == FALSE)} else {paste("NA- no models generated by this fit")}`

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for nbinom1
if (exists("nb2_models") && length(nb2_models) > 0) {
  nb2_dharma_results <- process_dharma_results(nb2_models)

  kable(nb2_dharma_results$summary)
}
```

### Zero-inflated Poisson Models

```{r}
#| eval: true
#| label: poi models

poi_models <- keep_at(
  poi_baseAge$all_models,
  filter(bestDist_filtered, dist == "poi")$outcome_id
)

```

```{r}
#| eval: true
#| label: poi-summary
if (exists("poi_models") && length(poi_models) > 0) {
  poi_unfit <- diagnoseZI(poi_models)

  poi_summary <- data.frame(
    outcome_id = names(poi_models),
    continuous_distribution = "poisson",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
  poi_ogu_models <- poi_baseAge$adj_results %>%
    filter(outcome_id %in% names(poi_models))
}
```

Are there any models that fail diagnostics: `r if (exists("poi_models") && length(poi_models) > 0) {any(poi_unfit$pass == FALSE)} else {paste("NA- no models generated by this fit")}`

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for Poisson models
if (exists("poi_models") && length(poi_models) > 0) {
  poi_dharma_results <- process_dharma_results(poi_models)

  kable(poi_dharma_results$summary)
}
```

### Zero-inflated Gaussian Models

```{r}
#| eval: true
#| label: gau unfit models
gau_models <- keep_at(
  gau_baseAge$all_models,
  filter(bestDist_filtered, dist == "gau")$outcome_id
)

```

```{r}
#| eval: true
#| label: gau-summary
if (exists("gau_models") && length(gau_models) > 0) {
  gau_unfit <- diagnoseZI(gau_models)

  gau_summary <- data.frame(
    outcome_id = names(gau_models),
    continuous_distribution = "gaussian",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
  gau_ogu_models <- gau_baseAge$adj_results %>%
    filter(outcome_id %in% names(gau_models))
}
```

Are there any models that fail diagnostics: `r if (exists("gau_models") && length(gau_models) > 0) {any(gau_unfit$pass == FALSE)} else {paste("NA- no models generated by this fit")}`

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for Poisson models
if (exists("gau_models") && length(gau_models) > 0) {
  gau_dharma_results <- process_dharma_results(gau_models)

  kable(gau_dharma_results$summary)
}
```

### Zero-inflated Lognormal Models

```{r}
#| eval: true
#| label: lognorm unfit models

lognorm_models <- keep_at(
  lognorm_baseAge$all_models,
  filter(bestDist_filtered, dist == "lognorm")$outcome_id
)

```

```{r}
#| eval: true
#| label: lognorm-summary
if (exists("lognorm_models") && length(lognorm_models) > 0) {
  lognorm_unfit <- diagnoseZI(gau_models)

  lognorm_summary <- data.frame(
    outcome_id = names(lognorm_models),
    continuous_distribution = "Log normal",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
  lognorm_ogu_models <- lognorm_baseAge$adj_results %>%
    filter(outcome_id %in% names(lognorm_models))
}
```

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for Poisson models
if (exists("lognorm_models") && length(lognorm_models) > 0) {
  lognorm_dharma_results <- process_dharma_results(lognorm_models)

  kable(lognorm_dharma_results$summary)
}
```

Are there any models that fail diagnostics: `r if (exists("lognorm_models") && length(lognorm_models) > 0) {any(lognorm_unfit$pass == FALSE)} else {paste("NA- no models generated by this fit")}`

### Save Final Model Fits

```{r}
#| eval: false
#| label: save
#| echo: false

save(
  nb1_models,
  nb2_models,
  poi_models,
  gau_models,
  lognorm_models,
  file = file.path(filepath, "baseAge-finalised-fits.RData")
)
```

```{r}
#| eval: false
#| label: load
#| echo: false

load(file.path(filepath, "baseAge-finalised-fits.RData"))

```

## Significant OGUs

```{r}
#| eval: true
#| label: tbl-summary
#| tbl-cap: Summary table of OGUs successfully fit by a model, and their corresponding coefficient p-values

nb1_summary <- left_join(nb1_summary, nb1_ogu_models)
nb2_summary <- left_join(nb2_summary, nb2_ogu_models)
poi_summary <- left_join(poi_summary, poi_ogu_models)

kable(
  summary <- rbind(
    nb1_summary,
    nb2_summary,
    poi_summary
  )
)
```

```{r}
#| eval: true
#| label: save-tbl-summary
#| echo: false
#| results: asis

save_and_link(
  data = summary,
  filename = "baseAge-zero-inflated-ogu-model-summaries.csv",
  description = "Baseline zero-inflated OGU model summaries",
  directory_naming = "both"
)
```

Of the `r ncol(outcome)` OGUs that passed filtering (`r ncol(outcome)/ncol(data$ogu_data)`), `r nrow(summary)` were sucessfully fit to a zero-inflated model (`r nrow(summary)/ncol(outcome)*100`%)

```{r}
#| eval: true
#| label: zi-sig

ziSig <- identify_significant_ogus(
  summary,
  data$taxa_data
)

colnames(ziSig)[colnames(ziSig) == "rowname"] <- "outcome_id"
```

## Save Significant Model Results

```{r}
#| eval: true
#| results: asis

save_and_link(
  data = summary,
  filename = "baseAge-significant-ogus-zero-inflated-model.csv",
  description = "Significant OGUs in FeZnPM based on ZI-GLMMs",
  directory_naming = "both"
)
```

## Zero-inflated component

```{r}
#| eval: true
#| label: zi-comp
#| fig-cap: NA
#| tbl-cap: NA

zi <- lapply(
  list(nb1_models, nb2_models, poi_models, gau_models),
  function(l) {
    lapply(l, function(x) {
      summary(x)$coefficients$zi[, 4]
    })
  }
)

#FIX make naming automatic using indexing
zip <- bind_rows(
  lapply(zi, function(model_type) {
    bind_rows(model_type, .id = "outcome_id")
  }),
  .id = "model_type"
) %>%
  # Add better names for model types
  mutate(
    model_type = case_when(
      model_type == "1" ~ "nb1",
      model_type == "2" ~ "nb2",
      model_type == "3" ~ "poi",
      model_type == "4" ~ "gau",
      TRUE ~ model_type
    )
  )

if (any(zip <= 0.05)) {
  zip_sig <- identify_significant_ogus(zip, data$taxa_data)
}
```

## Using EMMEANS to determine significant changes in taxa abundances

```{r}
#| eval: true
#| label: fit-mods

fits <- list(nb1_models, nb2_models, poi_models)

fits <- list_c(fits)
```

### Conditional Model EMMs

#### OGUs Estimated marginal means

| Scaled value | Age in days | Age              |
|--------------|-------------|------------------|
| 0.0529       | 365         | 1yr (12mths)     |
| 0.454        | 456         | 1.25yrs (15mths) |
| 0.859        | 548         | 1.5yrs (18mths)  |

```{r}
#| eval: true
#| label: Estimated marginal means for OGU abundances across different ages

emm <- lapply(fits, function(x) {
  emm <- emmeans(
    x,
    ~age_B,
    at = list(age_B = c(0.0529, 0.454, 0.859)),
    weights = "proportional",
    type = "scale"
  )

  levels(emm)$age_B <- c(
    "12mths",
    "15mths",
    "18mths"
  )

  return(emm)
})

```

##### OGU EMM differences between age groups in the baseline cohort (n = 168)

```{r}
#| eval: true
#| label: tbl-emm-pairs
#| tbl-cap: Comparison of abundance marginal mean estimates (EMMs) betweeen age groups in the baseline cohort (n = 168). Only those with p $\leq$ 0.05 are show for brevity.

emm_pairs <- lapply(emm, function(x) {
  contrast(
    x,
    method = "revpairwise",
    type = "scale",
    weights = "proportional",
    adjust = "fdr"
  )
})

emm_pairs_df <- emm_bind(emm_pairs, kable = F)

sigAge <- emm_pairs_df %>%
  filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
  mutate(
    change = ifelse(estimate < 0, "Decrease", "Increase")
  )
```

##### Effect Size

The default value used to standardise the effect size is the model's residual standard deviation (sigma; within-group variation). However, this may underestimate effect sizes in mixed models where combined variance may be more accurate. Here, the random effect variance is used to standardise effect sizes.

```{r}
#| eval: true
#| label: sigArm-eff
#| fig-cap: NA
#| tbl-cap: Effect sizes calculated as a standardized mean difference similar to Cohen's d.

emmEff <- lapply(1:length(emm_pairs), function(i) {
  eff_size(
    emm_pairs[[i]],
    sigma = sigma(fits[[i]]),
    edf = df.residual(fits[[i]]),
    method = "identity",
    parens = FALSE
  )
})

names(emmEff) <- names(emm_pairs)

emmEff_df <- emm_bind(emmEff, kable = F)

emmEff_df <- emmEff_df %>%
  filter(Model %in% sigAge$Model) %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    )
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, contrast, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

sigAgeEff <- left_join(sigAge, emmEff_df, by = c("Model", "contrast"))

mth12v15 <- sigAgeEff %>%
  filter(str_detect(contrast, "12mths")) %>%
  filter(str_detect(contrast, "15mths"))

mth12v18 <- sigAgeEff %>%
  filter(str_detect(contrast, "12mths")) %>%
  filter(str_detect(contrast, "18mths"))
```

```{r}
#| eval: true
#| label: fig-phyla-emms-effsize
#| fig-cap: Estimated marginal mean differences between different age groups in the baseline cohort (n = 168). Significant differences with an effect size > 0.5 (medium effect size) are shown in the plot.

plotPhylaEMMs(mth12v15, highlight_common = FALSE)
```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between different age groups in the baseline cohort (n = 168). Significant differences with an effect size > 0.5 (medium effect size) are shown in the plot.

plotPhylaEMMs(mth12v18, highlight_common = FALSE)
```

### Zero-inflated Componet EMMs

```{r}
#| eval: true
#| label: tbl-emms
#| tbl-cap: Estimated marginal means for OGU abundances between age groups in the baseline cohort (n = 168)

emm_zi <- lapply(fits, function(x) {
  # Get EMMs for arm at each time point
  emm <- emmeans(
    x,
    ~age_B,
    at = list(age_B = c(0.0529, 0.454, 0.859)),
    weights = "proportional",
    type = "scale",
    component = "zi"
  )

  levels(emm)$age_B <- c(
    "12mths",
    "15mths",
    "18mths"
  )

  return(emm)
})

# Convert to probability scale
zi_probs <- lapply(emm_zi, regrid, transform = "logit")
```

##### OGU EMM differences between age groups in the baseline cohort (n = 168)

```{r}
#| eval: true
#| label: tbl-emm-pairs
#| tbl-cap: Comparison of abundance marginal mean estimates (EMMs) betweeen age groups in the baseline cohort (n = 168)

emm_zi_pairs <- lapply(emm_zi, function(x) {
  contrast(
    x,
    method = "revpairwise",
    type = "scale",
    weights = "proportional",
    adjust = "fdr"
  )
})

emm_zi_pairs_df <- emm_bind(emm_zi_pairs, kable = F)

ZIsigAge <- emm_zi_pairs_df %>%
  filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
  mutate(
    change = ifelse(estimate < 0, "Decrease", "Increase")
  )
```

##### Effect Size

> \[!quote\] When a logistic regression is calculated, the regression coefficient (b1) is the estimated increase in the log odds of the outcome per unit increase in the value of the exposure. In other words, the exponential function of the regression coefficient (eb1) is the odds ratio associated with a one-unit increase in the exposure. [@Szumilas2010]

i.e. if $\beta_x = log(odds)$, then the odds ratio is $OR = \exp(\beta_x) = e^{\beta_x}$

```{r}
#| eval: true
#| label: odds-ratio
#| tbl-cap: Effect size calculated as odds-ratio for zero-inflation component
ZIodds <- lapply(fits, function(model) {
  exp(fixef(model)$zi)
})

kable(
  ZIodds <- do.call(rbind, ZIodds) %>%
    as.data.frame() %>%
    rownames_to_column("Model")
)
```

```{r}
#| eval: true
#| label: cohens-d
#| fig-cap: NA
#| tbl-cap: NA
ZIemmEff <- lapply(1:length(emm_zi_pairs), function(i) {
  eff_size(
    emm_zi_pairs[[i]],
    sigma = sigma(fits[[i]]),
    edf = df.residual(fits[[i]]),
    method = "identity",
    parens = FALSE
  )
})

names(ZIemmEff) <- names(emm_zi_pairs)

ZIemmEff <- emm_bind(ZIemmEff, kable = F)

ZIemmEff <- ZIemmEff %>%
  filter(Model %in% ZIsigAge$Model) %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    )
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, contrast, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

ZIsigAgeEff <- left_join(ZIsigAge, ZIemmEff, by = c("Model", "contrast"))
kable(ZIsigAgeEff <- left_join(ZIsigAgeEff, ZIodds, by = c("Model")))

mth12v15zi <- ZIsigAgeEff %>%
  filter(str_detect(contrast, "12mths")) %>%
  filter(str_detect(contrast, "15mths"))

mth12v18zi <- ZIsigAgeEff %>%
  filter(str_detect(contrast, "12mths")) %>%
  filter(str_detect(contrast, "18mths"))

```

```{r}
#| eval: true
#| label: fig-phyla-emms-effsize
#| fig-cap: Estimated marginal mean differences between different age groups in the baseline cohort (n = 168). Significant differences with an effect size > 0.5 (medium effect size) are shown in the plot.

plotPhylaEMMs(mth12v15zi, highlight_common = FALSE)
```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between different age groups in the baseline cohort (n = 168). Significant differences with an effect size > 0.5 (medium effect size) are shown in the plot.

plotPhylaEMMs(mth12v18zi, highlight_common = FALSE)
```

## Results Summary

```{r}
#| eval: true
#| echo: false
#| results: asis

list_all_results()
```

```{r}
#| eval: true
#| echo: false
#| results: asis
generate_results_section(directory_naming = "both")
```

## Appendix

```{r}
#| eval: true
#| echo: false
#| label: info

print(sessionInfo(), locale = FALSE)
```
