# Investigating Longitudional Impact of Intervention on OGU abundance using Zero-inflated Regression Models with Random Effects

```{r setup}
#| echo: false
#| eval: true
#| label: setup
#| cache: false

# This initial code block loads custom utility functions and environment variables
# These are stored in separate files to keep the main analysis clean and organized
suppressWarnings(suppressMessages({
  source("utils/r-utils.r")
  source("scripts/OGeeU.r")
  source("scripts/zeroGUS.r")

  source("_environment.required")
  load("data/seqdata.RData")

  library(glmmTMB)
  library(DHARMa)
  library(broom)
  library(patchwork)
  library(emmeans)
  library(ggtext)

  library(promicrobialTools)
  prefix <- "EvsBarm-"

  filepath <- get_results_dir(directory_naming = "both")
}))
```

Results will be saved to ``r filepath``.

## Prepare Outcome and Predictor Variables
```{r data-prep}
#| eval: true
#| label: data prep
#| results: asis

data <- prepare_ogu_data(
  absAbund,
  filter_prevalence = TRUE,
  prevalence_cutoff = 0.4,
  filter_abundance = FALSE
)

outcome <- data$ogu_data
rownames(outcome) <- str_replace(
  str_replace(str_remove(rownames(outcome), "_G"), "_B_B", "_B"),
  "_B_E",
  "_E"
)

predictors <- prepare_predictors(data$meta_data)
predictors <- predictors %>%
  select(subject, time, timepoint, arm, age_B) %>%
  mutate(arm = case_when(arm == "CPM" ~ 0, arm == "FeZnPM" ~ 1)) %>%
  rownames_to_column(var = "sample") %>%
  mutate(
    baseline_sample = str_replace(sample, "_G_E|_B_E|_E|_G_B|_B_B", "_B"),
    sample = str_remove(sample, "_G"),
  ) %>%
  mutate(sample = str_replace(sample, "_B_B", "_B")) %>%
  mutate(sample = str_replace(sample, "_B_E", "_E"))
```

## Variable Characteristics

### Predictor Variables

```{r}
#| eval: true
#| label: predictor-describe
#| results: asis

Hmisc::describe(predictors)

```

### OGU Outcome Variables

```{r zero-hist}
#| eval: true
#| label: fig-outcome-zeroProp
#| fig-cap: Distribution of zero proportions across variables

analyze_zeros(outcome, "density")
```

```{r disp}
#| eval: true
#| label: fig-outcome-dispersion
#| fig-cap: Distribution of dispearsion, defined as the variance-to-mean ratio. A log10 scale is used.

outdisp <- varDisp(outcome, threshold = 2, verbose = FALSE)

plot(outdisp, col = "#8CAAEE")

```

Generally, if the dispersion ratio is > 2 (0.3 on the log_10 scale), fitting a Poisson model will not accurately model the data (Poisson assumes that variance $\approx$ mean).

|      Model       |          Continuous component           | ZI component |                   Test                    |
|:----------------:|:---------------------------------------:|:------------:|:-----------------------------------------:|
|   noZi_models    |                  y ~ 1                  |     ~ 0      |  Is the zero-inflated component needed?   |
|   null_models    |                  y ~ 1                  |     ~ 1      | Tests if predictors affect zero-inflation |
| noZi_rand_models |         y ~ 1 + (1 \| subject)          |     ~ 0      |      Is the random intercept needed?      |
| null_rand_models |         y ~ 1 + (1 \| subject)          |     ~ 1      |                                           |
|        m1        | y ~ arm * time + age_B + (1 \| subject) | ~ arm * time |                                           |
|                  |                                         |              |                                           |

: List of model formulae tested

## Determining Direction of Change in Abundance over Time

```{r}
#| eval: true
#| label: tbl-abundance-changes
#| tbl-cap: Count Changes between Baseline and Endpoint

joined_data <- predictors %>%
  left_join(rownames_to_column(outcome, "sample"), by = "sample") %>%
  pivot_longer(
    cols = !c(
      "sample",
      "subject",
      "arm",
      "timepoint",
      "time",
      "age_B",
      "baseline_sample"
    ),
    names_to = "outcome_id"
  )

(joined_data_changes <- joined_data %>%
  group_by(outcome_id, timepoint) %>%
  summarise(average_count = mean(value)) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = average_count,
    values_fill = 0
  ) %>%
  mutate(EvsB = Endpoint - Baseline) %>%
  mutate(
    difference = case_when(
      EvsB < 0 ~ "Decrease",
      EvsB > 0 ~ "Increase",
      EvsB == 0 ~ "None",
      TRUE ~ NA
    )
  ) %>%
  group_by(difference) %>%
  summarise(total = n()) %>%
  kable())
```

The purpose of these models will be to determine if these changes are significant.

## Testing different distributions

The following distribution families were used to fit a zero-inflated model:

- Poisson
- Simple negative binomial (linear parameterization) [@Hardin2007]
- Complex negative binomial (quadratic parameterization) [@Hardin2007]
- Gaussian
- Log normal

```{r fit}
#| eval: false
#| include: false
nb2_EvsBarm <- genZI(
  outcome,
  predictors,
  formula = "time + arm + age_B + arm:time + (1 | subject)",
  zi = "~ arm * time",
  family = "nbinom2"
)

nb1_EvsBarm <- genZI(
  outcome,
  predictors,
  formula = "time + arm + age_B + arm:time + (1 | subject)",
  zi = "~ arm * time",
  family = "nbinom1"
)

poi_EvsBarm <- genZI(
  outcome,
  predictors,
  formula = "time + arm + age_B + arm:time + (1 | subject)",
  zi = "~ arm * time",
  family = "poisson"
)

gau_EvsBarm <- genZI(
  outcome,
  predictors,
  formula = "time + arm + age_B + arm:time + (1 | subject)",
  zi = "~ arm * time",
  family = "gaussian"
)

lognorm_EvsBarm <- genZI(
  outcome,
  predictors,
  formula = "time + arm + age_B + arm:time + (1 | subject)",
  zi = "~ arm * time",
  family = "lognormal"
)
```

```{r save}
#| eval: false
#| echo: false
saveRDS(
  nb2_EvsBarm,
  file.path(filepath, paste0(prefix, "genZI-OGU-nb2.rds"))
)

saveRDS(
  nb1_EvsBarm,
  file.path(filepath, paste0(prefix, "genZI-OGU-nb1.rds"))
)

saveRDS(
  poi_EvsBarm,
  file.path(filepath, paste0(prefix, "genZI-OGU-poi.rds"))
)

saveRDS(
  gau_EvsBarm,
  file.path(filepath, paste0(prefix, "genZI-OGU-gau.rds"))
)

saveRDS(
  lognorm_EvsBarm,
  file.path(filepath, paste0(prefix, "genZI-OGU-lognorm.rds"))
)
```

```{r load}
#| eval: true
#| echo: false
nb2_EvsBarm <- readRDS(file.path(
  filepath,
  "EvsBarm-genZI-OGU-nb2.rds"
))

nb1_EvsBarm <- readRDS(file.path(
  filepath,
  "EvsBarm-genZI-OGU-nb1.rds"
))

poi_EvsBarm <- readRDS(file.path(
  filepath,
  "EvsBarm-genZI-OGU-poi.rds"
))

gau_EvsBarm <- readRDS(file.path(
  filepath,
  "EvsBarm-genZI-OGU-gau.rds"
))

lognorm_EvsBarm <- readRDS(file.path(
  filepath,
  "EvsBarm-genZI-OGU-lognorm.rds"
))
```

### Model diagnostic failures

```{r}
#| eval: true
#| label: list

list <- list(
  gau = gau_EvsBarm,
  lognorm = lognorm_EvsBarm,
  poi = poi_EvsBarm,
  nb1 = nb1_EvsBarm,
  nb2 = nb2_EvsBarm
)
```

```{r}
#| eval: false
#| label: diagnose-mods

poor_fits <- lapply(1:length(list), function(x) {
  d <- diagnoseZI(list[[x]]$all_models)
  d$dist <- names(list)[x]

  return(d)
})

poor_fits <- data.table::rbindlist(poor_fits)
```

```{r}
#| eval: true
#| label: eval-diagnose-mods
#| include: false
#| cache: true
#invisible execution to avoid output
poor_fits <- lapply(1:length(list), function(x) {
  d <- diagnoseZI(list[[x]]$all_models)
  d$dist <- names(list)[x]

  return(d)
})

poor_fits <- data.table::rbindlist(poor_fits)
```

### Comparing Model AIC and BIC values

```{r}
#| eval: true
#| label: summaries

summaries <- list(
  gau = gau_EvsBarm$model_summary,
  lognorm = lognorm_EvsBarm$model_summary,
  poi = poi_EvsBarm$model_summary,
  nb1 = nb1_EvsBarm$model_summary,
  nb2 = nb2_EvsBarm$model_summary
)

summaries <- data.table::rbindlist(summaries, idcol = "dist")

summaries <- left_join(summaries, poor_fits, by = c("outcome_id", "dist"))
```

```{r}
#| eval: true
#| label: any-errors

if (any(summaries$error_message != "")) {
  print("Some models returned an error message")
  ind <- which(summaries$error_message != "")
  kable(summaries[ind, ])
}

```

```{r}
#| eval: true
#| label: diag-compare
summaries <- summaries %>%
  dplyr::filter(pass == TRUE) %>%
  select(!c(error_message, message)) %>%
  mutate(index = 1:nrow(.)) %>% #make it easier to distinguish between outcome ids
  pivot_longer(
    cols = !c(index, outcome_id, dist, pass),
    names_to = "criterion",
    values_to = "value"
  )
```

```{r}
#| eval: true
#| label: best-AIC

aic <- summaries %>%
  dplyr::filter(criterion == "AIC") %>%
  group_by(criterion, outcome_id) %>%
  arrange(value) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, minAIC = value, distAIC = dist)
```

```{r}
#| eval: true
#| label: best-BIC

bic <- summaries %>%
  dplyr::filter(criterion == "BIC") %>%
  group_by(criterion, outcome_id) %>%
  arrange(value) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, minBIC = value, distBIC = dist)

```

```{r}
#| eval: true
#| label: tbl-best-fits
#| tbl-cap: Best fit models based on AIC and BIC

(bestDist <- left_join(aic, bic) %>%
  mutate(dist = ifelse(distAIC == distBIC, distAIC, "Undetermined")))
```

### Checking for deviations in model assumptions

::: {.callout-note}

The larger the dispersion value, $\theta$ in `nbinom2` models, the less variance there is in the model [@Brooks2017].
:::

`dharmaDisp` represents the dispersion ratio, which should ideally be close to 1

```{r}
#| eval: true
#| label: tbl-best-disp
#| tbl-cap: lowest disp value for models that passed diagnostics

dispRatio <- summaries %>%
  dplyr::filter(criterion == "dharmaDisp") %>%
  group_by(criterion, outcome_id) %>%
  mutate(skew = abs(value - 1)) %>%
  arrange(skew) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, bestDisp = value, distDisp = dist)
```

```{r}
#| eval: true
#| label: tbl-pvals

pvals <- left_join(
  bestDist,
  summaries,
  by = c("index", "outcome_id", "dist")
) %>%
  dplyr::filter(str_detect(criterion, "Pvalue"), value <= 0.05) %>%
  arrange(index)
```

`r n_distinct(pvals$outcome_id)` models have a significant test result making the model fit problematic

### Final Model Distributions

```{r}
#| eval: true
#| label: best-dist-filtered
#| fig-cap: NA
#| tbl-cap: NA

bestDist_filtered <- bestDist %>%
  dplyr::filter(!outcome_id %in% pvals$outcome_id)
```

## Fitting models to best distribution

### Zero-inflated Negative Binomial 1 Models

```{r}
#| eval: true
#| label: nbinom1 models

nb1_models <- keep_at(
  nb1_EvsBarm$all_models,
  dplyr::filter(bestDist_filtered, dist == "nb1")$outcome_id
)

```

```{r}
#| eval: true
#| label: nb1-summary

if (exists("nb1_models") && length(nb1_models) > 0) {
  nb1_unfit <- diagnoseZI(nb1_models)
  nb1_summary <- data.frame(
    outcome_id = names(nb1_models),
    continuous_distribution = "nbinom1",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
  nb1_ogu_models <- nb1_EvsBarm$adj_results %>%
    dplyr::filter(outcome_id %in% names(nb1_models))
}
```

Are there any models that fail diagnostics: `r if (exists("nb1_models") && length(nb1_models) > 0) {any(nb1_unfit$pass == FALSE)} else {paste("NA- no models generated by this fit")}`

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for nbinom1
if (exists("nb1_models") && length(nb1_models) > 0) {
  nb1_dharma_results <- process_dharma_results(nb1_models)

  kable(nb1_dharma_results$summary)
}

```

### Zero-inflated Negative Binomial Polynomial Models

```{r}
#| eval: true
#| label: nbinom2 models

nb2_models <- keep_at(
  nb2_EvsBarm$all_models,
  dplyr::filter(bestDist_filtered, dist == "nb2")$outcome_id
)

```


```{r}
#| eval: true
#| label: nb2-summary
if (exists("nb2_models") && length(nb2_models) > 0) {
  nb2_unfit <- diagnoseZI(nb2_models)

  nb2_summary <- data.frame(
    outcome_id = names(nb2_models),
    continuous_distribution = "nbinom2",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
  nb2_ogu_models <- nb2_EvsBarm$adj_results %>%
    dplyr::filter(outcome_id %in% names(nb2_models))
}
```

Are there any models that fail diagnostics: `r if (exists("nb2_models") && length(nb2_models) > 0) {any(nb2_unfit$pass == FALSE)} else {paste("NA- no models generated by this fit")}`

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for nbinom1
if (exists("nb2_models") && length(nb2_models) > 0) {
  nb2_dharma_results <- process_dharma_results(nb2_models)

  kable(nb2_dharma_results$summary)
}
```

### Zero-inflated Poisson Models

```{r}
#| eval: true
#| label: poi models

poi_models <- keep_at(
  poi_EvsBarm$all_models,
  dplyr::filter(bestDist_filtered, dist == "poi")$outcome_id
)

```

```{r}
#| eval: true
#| label: poi-summary
if (exists("poi_models") && length(poi_models) > 0) {
  poi_unfit <- diagnoseZI(poi_models)

  poi_summary <- data.frame(
    outcome_id = names(poi_models),
    continuous_distribution = "poisson",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
  poi_ogu_models <- poi_EvsBarm$adj_results %>%
    dplyr::filter(outcome_id %in% names(poi_models))
}
```

Are there any models that fail diagnostics: `r if (exists("poi_models") && length(poi_models) > 0) {any(poi_unfit$pass == FALSE)} else {paste("NA- no models generated by this fit")}`

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for Poisson models
if (exists("poi_models") && length(poi_models) > 0) {
  poi_dharma_results <- process_dharma_results(poi_models)

  kable(poi_dharma_results$summary)
}
```

### Zero-inflated Gaussian Models

```{r}
#| eval: true
#| label: gau unfit models
gau_models <- keep_at(
  gau_EvsBarm$all_models,
  dplyr::filter(bestDist_filtered, dist == "gau")$outcome_id
)

```

```{r}
#| eval: true
#| label: gau-summary
if (exists("gau_models") && length(gau_models) > 0) {
  gau_unfit <- diagnoseZI(gau_models)

  gau_summary <- data.frame(
    outcome_id = names(gau_models),
    continuous_distribution = "gaussian",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
  gau_ogu_models <- gau_EvsBarm$adj_results %>%
    dplyr::filter(outcome_id %in% names(gau_models))
}
```

Are there any models that fail diagnostics: `r if (exists("gau_models") && length(gau_models) > 0) {any(gau_unfit$pass == FALSE)} else {paste("NA- no models generated by this fit")}`

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for Poisson models
if (exists("gau_models") && length(gau_models) > 0) {
  gau_dharma_results <- process_dharma_results(gau_models)

  kable(gau_dharma_results$summary)
}
```

### Zero-inflated Lognormal Models

```{r}
#| eval: true
#| label: lognorm unfit models

lognorm_models <- keep_at(
  lognorm_EvsBarm$all_models,
  dplyr::filter(bestDist_filtered, dist == "lognorm")$outcome_id
)

```

```{r}
#| eval: true
#| label: lognorm-summary
if (exists("lognorm_models") && length(lognorm_models) > 0) {
  lognorm_unfit <- diagnoseZI(gau_models)

  lognorm_summary <- data.frame(
    outcome_id = names(lognorm_models),
    continuous_distribution = "Log normal",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
  lognorm_ogu_models <- lognorm_EvsBarm$adj_results %>%
    dplyr::filter(outcome_id %in% names(lognorm_models))
}
```

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for Poisson models
if (exists("lognorm_models") && length(lognorm_models) > 0) {
  lognorm_dharma_results <- process_dharma_results(lognorm_models)

  kable(lognorm_dharma_results$summary)
}
```

Are there any models that fail diagnostics: `r if (exists("lognorm_models") && length(lognorm_models) > 0) {any(lognorm_unfit$pass == FALSE)} else {paste("NA- no models generated by this fit")}`

### Save Final Model Fits

```{r}
#| eval: true
#| label: save
#| echo: false

save(
  nb1_models,
  nb2_models,
  poi_models,
  gau_models,
  lognorm_models,
  file = file.path(filepath, paste0(prefix, "finalised-fits.RData"))
)
```

```{r}
#| eval: false
#| label: load
#| echo: false

load(file.path(filepath, paste0(prefix, "finalised-fits.RData")))

```

## Significant OGUs

```{r}
#| eval: true
#| label: model-ogus
nb1_ogu_models <- nb1_EvsBarm$adj_results %>%
  dplyr::filter(outcome_id %in% names(nb1_models))

nb2_ogu_models <- nb2_EvsBarm$adj_results %>%
  dplyr::filter(outcome_id %in% names(nb2_models))

poi_ogu_models <- poi_EvsBarm$adj_results %>%
  dplyr::filter(outcome_id %in% names(poi_models))

gau_ogu_models <- gau_EvsBarm$adj_results %>%
  dplyr::filter(outcome_id %in% names(gau_models))
```

```{r}
#| eval: true
#| label: tbl-summary
#| tbl-cap: Summary table of OGUs successfully fit by a model, and their corresponding coefficient p-values

poi_summary <- left_join(poi_summary, poi_ogu_models)
nb1_summary <- left_join(nb1_summary, nb1_ogu_models)
nb2_summary <- left_join(nb2_summary, nb2_ogu_models)
gau_summary <- left_join(gau_summary, gau_ogu_models)


summary <- rbind(nb1_summary, nb2_summary, poi_summary, gau_summary)
kable(head(summary))
```

```{r}
#| eval: true
#| label: save-tbl-summary
#| echo: false
#| results: asis
tmp <- summary %>%
  left_join(rownames_to_column(data$taxa_data, "outcome_id"), by = "outcome_id")

save_and_link(
  data = tmp,
  filename = paste0(prefix, "zero-inflated-ogu-fitted-model-pvalues.csv"),
  description = "Summary table of OGUs successfully fit by a model",
  directory_naming = "both"
)

```

Of the `r ncol(outcome)` OGUs that passed filtering (`r ncol(outcome)/ncol(data$ogu_data)`), `r nrow(summary)` were sucessfully fit to a zero-inflated model (`r nrow(summary)/ncol(outcome)*100`%)

```{r}
#| eval: true
#| label: zi-sig

ziSig <- identify_significant_ogus(
  summary,
  data$taxa_data
)

colnames(ziSig)[colnames(ziSig) == "rowname"] <- "outcome_id"
```

### Average abundance changes

```{r}
#| eval: true
#| label: tbl-average-sig-changes

aveChange <- joined_data %>%
  dplyr::filter(outcome_id %in% ziSig$outcome_id) %>%
  group_by(outcome_id, timepoint) %>%
  summarise(average_count = mean(value)) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = average_count,
    values_fill = NA
  ) %>%
  mutate(EvsB = Endpoint - Baseline) %>%
  mutate(
    difference = case_when(
      EvsB < 0 ~ "Decrease",
      EvsB > 0 ~ "Increase",
      EvsB == 0 ~ "None",
      TRUE ~ NA
    )
  )

aveChange %>%
  group_by(difference) %>%
  summarise(total = n()) %>%
  kable()
```

### Save Significant Model Results

```{r}
#| eval: true
#| results: asis

save_and_link(
  data = ziSig,
  filename = paste0(
    prefix,
    "zero-inflated-ogu-fitted-model-significant-pvalues.csv"
  ),
  description = "Significant OGUs based on ZI-GLMMs",
  directory_naming = "both"
)
```

## Zero-inflated component

```{r}
#| eval: true
#| label: zi-comp
zi <- lapply(
  list(nb1_models, nb2_models, poi_models, gau_models),
  function(l) {
    lapply(l, function(x) {
      summary(x)$coefficients$zi[, 4]
    })
  }
)

#FIX make naming automatic using indexing
zip <- bind_rows(
  lapply(zi, function(model_type) {
    bind_rows(model_type, .id = "outcome_id")
  }),
  .id = "model_type"
) %>%
  # Add better names for model types
  mutate(
    model_type = case_when(
      model_type == "1" ~ "nb1",
      model_type == "2" ~ "nb2",
      model_type == "3" ~ "poi",
      model_type == "4" ~ "gau",
      TRUE ~ model_type
    )
  )

if (any(zip <= 0.05)) {
  zip_sig <- identify_significant_ogus(zip, data$taxa_data)
}
```

```{r}
#| eval: true
#| label: save-zip-sig
#| results: asis
save_and_link(
  data = zip_sig,
  filename = paste0(
    prefix,
    "significant-ogus-zero-inflated-model-ZI-component.csv"
  ),
  description = "Significant OGUs based on ZI-GLMMs (zero-inflated compponent)",
  directory_naming = "both"
)

```

## Using EMMEANS to determine significant changes in taxa abundances

```{r fits}
#| eval: true
#| label: fit-mods

fits <- list(nb1_models, nb2_models, poi_models, gau_models)

fits <- list_c(fits)
```

```{r}
#| eval: true
#| label: save-fits
#| results: asis

save_and_link(
  data = fits,
  filename = paste0(
    prefix,
    "zero-inflated-ogu-fitted-model-fits-variable.rds"
  ),
  description = "RDS file for `fits` variable",
  directory_naming = "both"
)
```

### Conditional Model EMMs

#### Overall OGUs Estimated marginal means

The below estimates the mean for each `arm` at each `timepoint`, controlling for age at baseline

```{r emm}
#| eval: true
#| label: tbl-within-arm
#| tbl-cap: Estimated marginal means for OGU abundances within each intervention arm

emm <- lapply(fits, function(x) {
  emm <- emmeans(
    x,
    ~ arm | time,
    at = list(time = c(0, 1)),
    weights = "proportional",
    type = "scale"
  )

  levels(emm)$time <- c("Baseline", "Endpoint")
  levels(emm)$arm <- c("CPM", "FeZnPM")

  return(emm)
})
```

#### OGU EMM differences between intervention arm at baseline and endpoint

```{r}
#| eval: true
#| label: tbl-within-arms-cont
#| tbl-cap: Comparison of abundance marginal mean estimates (EMMs) betweeen intervention arm at baseline and endpoint. Only those with p $\leq$ 0.05 are show for brevity.

emm_pairs <- lapply(emm, function(x) {
  contrast(
    x,
    by = c("time"),
    method = "revpairwise",
    type = "scale",
    weights = "proportional",
    adjust = "fdr"
  )
})

emm_pairs_df <- emm_bind(emm_pairs, kable = F)

kable(head(
  sigArms <- emm_pairs_df %>%
    dplyr::filter(p.value <= 0.05) %>%
    left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
    mutate(
      change = ifelse(estimate < 0, "Decrease", "Increase"),
      timepoint = time
    )
))
```

##### Effect Size

The default value used to standardise the effect size is the model's residual standard deviation (sigma; within-group variation). However, this may underestimate effect sizes in mixed models where combined variance may be more accurate. Here, the random effect variance is used to standardise effect sizes.

```{r}
#| eval: true
#| label: sigArm-eff
#| tbl-cap: Effect sizes calculated as a standardized mean difference similar to Cohen's d.

emmEff <- lapply(1:length(emm_pairs), function(i) {
  eff_size(
    emm_pairs[[i]],
    sigma = attr(VarCorr(fits[[i]])$cond$subject, "stddev") * sqrt(2),
    edf = df.residual(fits[[i]]),
    method = "identity",
    parens = FALSE
  )
})

names(emmEff) <- names(emm_pairs)

emmEff_df <- emm_bind(emmEff, kable = F)

emmEff_df <- emmEff_df %>%
  dplyr::filter(Model %in% sigArms$Model) %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    )
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, time, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

sigArmsEff <- left_join(sigArms, emmEff_df, by = c("Model", "time"))
```

```{r}
#| eval: true
#| label: fig-phyla-emms-effsize
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint. Significant differences with an effect size > 0.5 (medium effect size) are shown in the plot.

plotPhylaEMMs(sigArmsEff) +
  facet_wrap(~timepoint, scales = "free")
```

```{r}
#| eval: true
#| label: tbl-common-arm-EB
#| tbl-cap: Common OGUs at baseline and endpoint that have significanlty different marginal means between intervetion arms

common <- sigArmsEff %>%
  group_by(Model) %>%
  summarize(n = n_distinct(time)) %>%
  dplyr::filter(n > 1) %>%
  pull(Model)

sigArmsEff %>%
  dplyr::filter(Model %in% common) %>%
  mutate(Species = ifelse(Species == "", Genus, Species)) %>%
  select(Model, timepoint, Phylum, Species, change, magnitude) %>%
  kable()
```

#### Rate of OGU EMM changes over time within each intervention arm

```{r}
#| eval: true
#| label: tbl-emm-arm-trend
#| tbl-cap: Rate of change per unit increase of abundance marginal mean estimates over time. The time.trend column represents the slope of the relationship with time.
trend <- lapply(fits, function(x) {
  emm <- emtrends(
    x,
    ~arm,
    var = "time",
    weights = "proportional",
    type = "scale",
    adjust = "fdr"
  )

  levels(emm)$arm <- c("CPM", "FeZnPM")

  return(emm)
})
```

```{r}
#| eval: true
#| label: tbl-emm-arm-trend-compare
#| tbl-cap: Comparison of the rates of change from baseline to endpoint in abundance EMMs between intervention arms.

slopes <- lapply(trend, function(x) {
  contrast(
    x,
    method = "revpairwise",
    type = "scale",
    weights = "proportional",
    adjust = "fdr"
  )
})

slopes_df <- emm_bind(slopes, kable = F) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
  mutate(Species = ifelse(Species == "", Genus, Species))

sigArmsTime <- slopes_df %>%
  dplyr::filter(p.value <= 0.05)
```

##### Effect Size

The default value used to standardise the effect size is the model's residual standard deviation (sigma; within-group variation). However, this may underestimate effect sizes in mixed models where combined variance may be more accurate. Here, the random effect variance is used to standardise effect sizes.

```{r}
#| eval: true
#| label: sigArm-eff
#| fig-cap: NA
#| tbl-cap: Effect sizes calculated as a standardized mean difference similar to Cohen's d.

emmEff <- lapply(1:length(slopes), function(i) {
  eff_size(
    slopes[[i]],
    sigma = attr(VarCorr(fits[[i]])$cond$subject, "stddev") * sqrt(2),
    edf = df.residual(fits[[i]]),
    method = "identity",
    parens = FALSE
  )
})

names(emmEff) <- names(slopes)

emmEff_df <- emm_bind(emmEff, kable = F)

emmEff_df <- emmEff_df %>%
  dplyr::filter(Model %in% sigArmsTime$Model) %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    )
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, contrast, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

sigArmsTimeEff <- left_join(sigArmsTime, emmEff_df, by = c("Model", "contrast"))

```

```{r}
#| eval: true
#| label: save-sigTime
#| results: asis

save_and_link(
  data = slopes_df,
  filename = paste0(prefix, "ogu-emms-time-ZI.csv"),
  description = "OGU EMMs differences between timepoint",
  directory_naming = "both"
)

save_and_link(
  data = sigArmsTimeEff,
  filename = paste0(prefix, "significant-ogu-emms-time-ZI.csv"),
  description = "OGU EMMs significantly different at endpoint and their effect sizes",
  directory_naming = "both"
)
```

```{r}
#| eval: true
#| label: fig-phyla-emms-effsize
#| fig-cap: Rate of change in estimated marginal means between intervention arms. Significant differences with an effect size > 0.5 (medium effect size) are shown in the plot.

plotPhylaEMMs(
  dplyr::filter(sigArmsTimeEff, abs(effect.size) >= 0.5),
  highlight_common = F
)
```

```{r}
#| eval: true
#| fig-cap: NA
#| tbl-cap: NA
# Create the direction-based counts
df_with_direction <- sigArmsTimeEff %>%
  filter(magnitude != "small") %>%
  mutate(Direction = ifelse(estimate > 0, "Increase", "Decrease"))

phylum_direction_counts <- df_with_direction %>%
  count(Phylum, Direction) %>%
  arrange(desc(n))

# Create the plot with labels
ggplot(
  phylum_direction_counts,
  aes(x = reorder(Phylum, n), y = n, fill = Direction)
) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(
    aes(label = n),
    position = position_stack(vjust = 0.5), # Center labels in each segment
    color = "white",
    size = 3.5
  ) +
  coord_flip() +
  scale_fill_manual(
    values = c("Decrease" = "#E64B35", "Increase" = "#4DBBD5")
  ) +
  theme_minimal() +
  labs(
    x = "Phylum",
    y = "Number of OGUs",
    title = "Distribution of OGUs by Phylum and Direction of Change"
  ) +
  theme(
    axis.text.y = element_text(size = 10),
    plot.title = element_text(size = 12, face = "bold"),
    panel.grid.major.y = element_blank(),
    legend.position = "right"
  )

```


### Zero-inflated Componet EMMs

```{r}
#| eval: true
#| label: tbl-emms
#| tbl-cap: Estimated marginal means for OGU abundances within each intervention arm

emm_zi <- lapply(fits, function(x) {
  # Get EMMs for arm at each time point

  emm <- emmeans(
    x,
    ~ arm | time,
    at = list(time = c(0, 1)),
    weights = "proportional",
    type = "scale",
    component = "zi"
  )

  levels(emm)$time <- c("Baseline", "Endpoint")
  levels(emm)$arm <- c("CPM", "FeZnPM")

  return(emm)
})

# Convert to probability scale
zi_probs <- lapply(emm_zi, regrid, transform = "logit")
```

#### OGU EMM differences between intervention arm at baseline and endpoint

```{r}
#| eval: true
#| label: tbl-emm-pairs
#| tbl-cap: Comparison of abundance marginal mean estimates (EMMs) betweeen intervention arm at baseline and endpoint

emm_zi_pairs <- lapply(
  emm_zi,
  pairs,
  reverse = TRUE,
  by = c("time"),
  type = "scale",
  weights = "proportional",
  adjust = "fdr"
)

emm_zi_pairs_df <- emm_bind(emm_zi_pairs, kable = F)

ZIsigArms <- emm_zi_pairs_df %>%
  dplyr::filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
  mutate(
    change = ifelse(estimate < 0, "Decrease", "Increase"),
    timepoint = time
  )
```

##### Effect Size

> [!quote]
> When a logistic regression is calculated, the regression coefficient (b1) is the estimated increase in the log odds of the outcome per unit increase in the value of the exposure. In other words, the exponential function of the regression coefficient (eb1) is the odds ratio associated with a one-unit increase in the exposure. [@Szumilas2010]

i.e. if $\beta_x = log(odds)$, then the odds ratio is $OR = \exp(\beta_x) = e^{\beta_x}$

The "nonexposure" group in this context would be the CPM group

```{r}
#| eval: true
#| label: odds-ratio
#| tbl-cap: Effect size calculated as odds-ratio and converted to Cohen's d with respect to the rate of the desired outcome for zero-inflation component

oddsEff <- Map(summarize_zi_contrasts, emm_zi, emm_zi_pairs, fits)

oddsEff <- data.table::rbindlist(oddsEff, idcol = "Model") %>%
  select(
    Model,
    contrast,
    time,
    OR,
    OR_lower,
    OR_upper,
    P0,
    cohens_d,
    magnitude
  )
```

```{r}
#| eval: true
#| label: cohens-d
#| fig-cap: NA
#| tbl-cap: NA
ZIemmEff <- lapply(1:length(emm_zi_pairs), function(i) {
  eff_size(
    emm_zi_pairs[[i]],
    sigma = attr(VarCorr(fits[[i]])$cond$subject, "stddev") * sqrt(2),
    edf = df.residual(fits[[i]]),
    method = "identity",
    parens = FALSE,
    ratios = TRUE
  )
})

names(ZIemmEff) <- names(emm_zi_pairs)

ZIemmEff <- emm_bind(ZIemmEff, kable = F)

ZIemmEff <- ZIemmEff %>%
  dplyr::filter(Model %in% ZIsigArms$Model) %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    )
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, time, contrast, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

ZIsigArmsEff <- left_join(
  ZIsigArms,
  ZIemmEff,
  by = c("Model", "time", "contrast")
)

kable(
  ZIsigArmsEff <- left_join(
    ZIsigArmsEff,
    oddsEff,
    by = c("Model", "time", "contrast")
  ) %>%
    select(!c(Kingdom, Class, Order, Family, time))
)

if (!all(ZIsigArmsEff$magnitude == ZIsigArmsEff$OR_magnitude)) {
  echo("Warning: Effect size and odds ratio magnitudes do not match")
}
```

::: {.callout-note}
In zero-inflated models, the logit function models the probability of false zeros vs non-zeros, the interpretation of these emmeans contrasts would be:

- Positive Estimates (e.g., Rothia dentocariosa, estimate = 2.20):
  - Higher probability of being a true count (non-false zero) in FeZnPM compared to CPM
  - OR > 1 (9.056) means ~9 times more likely to be a true count in FeZnPM
  - P0 = 0.451 suggests about 45% probability of false zeros overall
- Negative Estimates (e.g., Yersinia spp, estimate = -8.15):
  - Lower probability of being a true count in FeZnPM compared to CPM
  - OR < 1 (0.000) means much less likely to be a true count in FeZnPM
  - P0 = 0.341 suggests about 34% probability of false zeros overall
- Statistical Significance:
  - All contrasts are significant (p ≤ 0.05)
  - Confidence intervals for ORs don't cross 1 (except where very wide)

This suggests differential presence/absence patterns between treatment groups, accounting for potential false zeros in the data.
:::

```{r}
#| eval: true
#| label: fig-phyla-emms-effsize
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint. Significant differences with an effect size > 0.5 (medium effect size) are shown in the plot.

plotPhylaEMMs(ZIsigArmsEff, text_size = 12) +
  facet_wrap(~timepoint, scales = "free")
```

#### Rates of OGU abundace changes over time within each intervention arm

```{r}
#| eval: true
#| label: tbl-emm-arm-trend
#| tbl-cap: Rate of change per unit increase of abundance marginal mean estimates over time. The time.trend column represents the slope of the relationship with time.

trend_zi <- lapply(fits, function(x) {
  emm <- emtrends(
    x,
    ~arm,
    var = "time",
    weights = "proportional",
    type = "scale"
  )

  levels(emm)$arm <- c("CPM", "FeZnPM")

  return(emm)
})
```

```{r}
#| eval: true
#| label: tbl-emm-arm-trend-compare
#| tbl-cap: Comparison of the rates of change from baseline to endpoint in abundance EMMs between intervention arms.

slopes_zi <- lapply(trend_zi, function(x) {
  contrast(
    x,
    method = "revpairwise",
    type = "scale",
    weights = "proportional",
    adjust = "fdr"
  )
})

slopes_zi_df <- emm_bind(slopes_zi, kable = F)

ZIsigArmsTime <- slopes_zi_df %>%
  dplyr::filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model")
```

##### Effect Size

The default value used to standardise the effect size is the model's residual standard deviation (sigma; within-group variation). However, this may underestimate effect sizes in mixed models where combined variance may be more accurate. Here, the random effect variance is used to standardise effect sizes.

```{r}
#| eval: true
#| label: sigArm-eff
#| fig-cap: NA
#| tbl-cap: Effect sizes calculated as a standardized mean difference similar to Cohen's d.

emmEff_zi <- lapply(1:length(slopes_zi), function(i) {
  eff_size(
    slopes_zi[[i]],
    sigma = attr(VarCorr(fits[[i]])$cond$subject, "stddev") * sqrt(2),
    edf = df.residual(fits[[i]]),
    method = "identity",
    parens = FALSE
  )
})

names(emmEff_zi) <- names(slopes_zi)

emmEff_zi_df <- emm_bind(emmEff_zi, kable = F)

emmEff_zi_df <- emmEff_zi_df %>%
  dplyr::filter(Model %in% ZIsigArmsTime$Model) %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    )
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, contrast, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

ZIsigArmsTimeEff <- left_join(
  ZIsigArmsTime,
  emmEff_zi_df,
  by = c("Model", "contrast")
)

```

```{r}
#| eval: true
#| label: fig-phyla-emms-effsize
#| fig-cap: Estimated marginal mean differences between intervention arms at baseline and endpoint.

plotPhylaEMMs(
  dplyr::filter(ZIsigArmsTimeEff, abs(effect.size) >= 0.5),
  highlight_common = FALSE
)
```

## Results Summary

```{r}
#| eval: true
#| echo: false
#| results: asis

list_all_results()
```

```{r}
#| results: asis
generate_results_section(directory_naming = "both")
```

## Appendix

```{r}
#| eval: true
#| label: info

print(sessionInfo(), locale = FALSE)
```

### Demonstrating the importance of random effects

OGU G012221805 has an average abundance change of zero, so why is the EvsB term significant? Random effects!

```{r}
#| eval: true
#| label: fig-indv-changes
#| fig-cap: Changes in count abundance for OGU G012221805, showing the clear impact of individual random effects and the importance of including such in a model. Note the black dotted line would be the standard regression line indicating no change over time.

joined_data %>%
  dplyr::filter(outcome_id == "G012221805") %>%
  group_by(subject) %>%
  ggplot(aes(x = time, y = value, )) +
  geom_point() +
  geom_line(
    aes(
      group = subject,
      col = as.factor(arm)
    ),
    linewidth = 1
  ) +
  geom_smooth(
    method = "lm",
    se = F,
    color = "black",
    formula = y ~ x,
    linetype = 'dotted',
    linewidth = 3
  ) +
  xlab("Time Since Feed Start (days)") +
  ylab("Count") +
  scale_color_manual(values = armColours) +
  labs(color = "Intervention Arm")

```

### Individual abundance changes in Signficant OGUs

```{r}
#| eval: true
#| label: tmp
#| collapse: true
#| tbl-cap: Individual abundance changes for each OGU deemed significant by ZI models

tmp <- joined_data %>%
  dplyr::filter(outcome_id %in% ziSig$outcome_id) %>%
  group_by(sample, outcome_id) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = value,
    values_fill = 0
  ) %>%
  mutate(EvsB = Endpoint - Baseline) %>%
  mutate(
    difference = case_when(
      EvsB < 0 ~ "Decrease",
      EvsB > 0 ~ "Increase",
      EvsB == 0 ~ "None",
      TRUE ~ NA
    )
  )

tmp %>%
  group_by(outcome_id, difference) %>%
  summarise(total = n()) %>%
  kable()
```