# Investigating changes in OGU abundance during the second year of life using Zero-inflated Regression Models with Random Effects

The difference between this analysis and previous ZI model analysis is that here, *intervention arm is being completely ignored* so as to look at this cohort as a whole and try understand the effect of time on the microbiota in the context of a high burden of inflammation

**Paired data is not used here**

```{r setup}
#| echo: false
#| eval: true
#| label: setup
#| cache: false

# This initial code block loads custom utility functions and environment variables
# These are stored in separate files to keep the main analysis clean and organized
suppressWarnings(suppressMessages({
  source("utils/r-utils.r")
  source("scripts/OGeeU.r")
  source("scripts/zeroGUS.r")

  source("_environment.required")
  load("data/seqdata.RData")

  library(glmmTMB)
  library(DHARMa)
  library(broom)
  library(patchwork)
  library(emmeans)
  library(ggtext)

  library(promicrobialTools)

  prefix <- "EvsB-"

  filepath <- get_results_dir(directory_naming = "both")
}))
```

Results will be saved to ``r filepath``.

## Prepare Outcome and Predictor Variables

```{r data-prep}
#| eval: true
#| label: data prep
#| results: asis

data <- prepare_ogu_data(
  absAbund,
  filter_prevalence = TRUE,
  prevalence_cutoff = 0.4,
  filter_abundance = FALSE
)

outcome <- data$ogu_data
rownames(outcome) <- str_replace(
  str_replace(str_remove(rownames(outcome), "_G"), "_B_B", "_B"),
  "_B_E",
  "_E"
)

meta <- prepare_predictors(data$meta_data)
predictors <- meta %>%
  rownames_to_column(var = "sample") %>%
  mutate(
    baseline_sample = str_replace(sample, "_G_E|_B_E|_E|_G_B|_B_B", "_B"),
    sample = str_remove(sample, "_G"),
  ) %>%
  mutate(sample = str_replace(sample, "_B_B", "_B")) %>%
  mutate(sample = str_replace(sample, "_B_E", "_E")) %>%
  select(
    sample,
    age_B,
    timepoint,
    time,
    subject
  )
```

## Testing different distributions

The following distribution families were used to fit a zero-inflated model:

- Poisson
- Simple negative binomial (linear parameterization) [@Hardin2007]
- Complex negative binomial (quadratic parameterization) [@Hardin2007]
- Gaussian
- Log normal

```{r fit}
#| eval: true
#| include: false
nb2_EvsB <- genZI(
  outcome,
  predictors,
  formula = "time + age_B + age_B:time + (1 | subject)",
  zi = "~ age_B * time",
  family = "nbinom2"
)

nb1_EvsB <- genZI(
  outcome,
  predictors,
  formula = "time + age_B + age_B:time + (1 | subject)",
  zi = "~ age_B * time",
  family = "nbinom1"
)

poi_EvsB <- genZI(
  outcome,
  predictors,
  formula = "time + age_B + age_B:time + (1 | subject)",
  zi = "~ age_B * time",
  family = "poisson"
)

gau_EvsB <- genZI(
  outcome,
  predictors,
  formula = "time + age_B + age_B:time + (1 | subject)",
  zi = "~ age_B * time",
  family = "gaussian"
)

lognorm_EvsB <- genZI(
  outcome,
  predictors,
  formula = "time + age_B + age_B:time + (1 | subject)",
  zi = "~ age_B * time",
  family = "lognormal"
)
```

```{r save}
#| eval: true
#| echo: false
saveRDS(
  nb2_EvsB,
  file.path(filepath, paste0(prefix, "genZI-OGU-nb2_results.rds"))
)

saveRDS(
  nb1_EvsB,
  file.path(filepath, paste0(prefix, "genZI-OGU-nb1_results.rds"))
)

saveRDS(
  poi_EvsB,
  file.path(filepath, paste0(prefix, "genZI-OGU-poi_results.rds"))
)

saveRDS(
  gau_EvsB,
  file.path(filepath, paste0(prefix, "genZI-OGU-gau_results.rds"))
)

saveRDS(
  lognorm_EvsB,
  file.path(filepath, paste0(prefix, "genZI-OGU-lognorm_results.rds"))
)
```

```{r load}
#| eval: false
#| echo: false
nb2_EvsB <- readRDS(file.path(
  filepath,
  "EvsB-genZI-OGU-nb2_results.rds"
))

nb1_EvsB <- readRDS(file.path(
  filepath,
  "EvsB-genZI-OGU-nb1_results.rds"
))

poi_EvsB <- readRDS(file.path(
  filepath,
  "EvsB-genZI-OGU-poi_results.rds"
))

gau_EvsB <- readRDS(file.path(
  filepath,
  "EvsB-genZI-OGU-gau_results.rds"
))

lognorm_EvsB <- readRDS(file.path(
  filepath,
  "EvsB-genZI-OGU-lognorm_results.rds"
))
```

### Model diagnostic failures

```{r}
#| eval: true
#| label: list

list <- list(
  gau = gau_EvsB,
  lognorm = lognorm_EvsB,
  poi = poi_EvsB,
  nb1 = nb1_EvsB,
  nb2 = nb2_EvsB
)
```
```{r}
#| eval: false
#| label: diagnose-mods

poor_fits <- lapply(1:length(list), function(x) {
  d <- diagnoseZI(list[[x]]$all_models)
  d$dist <- names(list)[x]

  return(d)
})

poor_fits <- data.table::rbindlist(poor_fits)
```

```{r}
#| eval: true
#| label: eval-diagnose-mods
#| include: false
#| cache: true
#invisible execution to avoid output
poor_fits <- lapply(1:length(list), function(x) {
  d <- diagnoseZI(list[[x]]$all_models)
  d$dist <- names(list)[x]

  return(d)
})

poor_fits <- data.table::rbindlist(poor_fits)
```

### Comparing Model AIC and BIC values

```{r}
#| eval: true
#| label: summaries

summaries <- list(
  gau = gau_EvsB$model_summary,
  lognorm = lognorm_EvsB$model_summary,
  poi = poi_EvsB$model_summary,
  nb1 = nb1_EvsB$model_summary,
  nb2 = nb2_EvsB$model_summary
)

summaries <- data.table::rbindlist(summaries, idcol = "dist")

summaries <- left_join(summaries, poor_fits, by = c("outcome_id", "dist"))
```

```{r}
#| eval: true
#| label: any-errors

if (any(summaries$error_message != "")) {
  print("Some models returned an error message")
  ind <- which(summaries$error_message != "")
  kable(summaries[ind, ])
}

```

```{r}
#| eval: true
#| label: diag-compare
summaries <- summaries %>%
  filter(pass == TRUE) %>%
  select(!c(error_message, message)) %>%
  mutate(index = 1:nrow(.)) %>% #make it easier to distinguish between outcome ids
  pivot_longer(
    cols = !c(index, outcome_id, dist, pass),
    names_to = "criterion",
    values_to = "value"
  )
```

```{r}
#| eval: true
#| label: best-AIC

aic <- summaries %>%
  filter(criterion == "AIC") %>%
  group_by(criterion, outcome_id) %>%
  arrange(value) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, minAIC = value, distAIC = dist)
```

```{r}
#| eval: true
#| label: best-BIC

bic <- summaries %>%
  filter(criterion == "BIC") %>%
  group_by(criterion, outcome_id) %>%
  arrange(value) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, minBIC = value, distBIC = dist)

```

```{r}
#| eval: true
#| label: tbl-best-fits
#| tbl-cap: Best fit models based on AIC and BIC

(bestDist <- left_join(aic, bic) %>%
  mutate(dist = ifelse(distAIC == distBIC, distAIC, "Undetermined")))
```

### Checking for deviations in model assumptions

::: {.callout-note}

The larger the dispersion value, $\theta$ in `nbinom2` models, the less variance there is in the model [@Brooks2017].
:::

`dharmaDisp` represents the dispersion ratio, which should ideally be close to 1

```{r}
#| eval: true
#| label: tbl-best-disp
#| tbl-cap: lowest disp value for models that passed diagnostics

dispRatio <- summaries %>%
  filter(criterion == "dharmaDisp") %>%
  group_by(criterion, outcome_id) %>%
  mutate(skew = abs(value - 1)) %>%
  arrange(skew) %>%
  slice(1) %>%
  ungroup() %>%
  select(index, outcome_id, bestDisp = value, distDisp = dist)
```

```{r}
#| eval: true
#| label: tbl-pvals

pvals <- left_join(
  bestDist,
  summaries,
  by = c("index", "outcome_id", "dist")
) %>%
  filter(str_detect(criterion, "Pvalue"), value <= 0.05) %>%
  arrange(index)
```

`r n_distinct(pvals$outcome_id)` models have a significant test result making the model fit problematic

### Final Model Distributions

```{r}
#| eval: true
#| label: best-dist-filtered
#| fig-cap: NA
#| tbl-cap: NA

bestDist_filtered <- bestDist %>%
  filter(!outcome_id %in% pvals$outcome_id)
```

## Fitting models to best distribution

### Zero-inflated Negative Binomial 1 Models

```{r}
#| eval: true
#| label: nbinom1 models

nb1_models <- keep_at(
  nb1_EvsB$all_models,
  filter(bestDist_filtered, dist == "nb1")$outcome_id
)

```

```{r}
#| eval: true
#| label: nb1-summary

if (exists("nb1_models") && length(nb1_models) > 0) {
  nb1_unfit <- diagnoseZI(nb1_models)
  nb1_summary <- data.frame(
    outcome_id = names(nb1_models),
    continuous_distribution = "nbinom1",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
  nb1_ogu_models <- nb1_EvsB$adj_results %>%
    filter(outcome_id %in% names(nb1_models))
}
```

Are there any models that fail diagnostics: `r if (exists("nb1_models") && length(nb1_models) > 0) {any(nb1_unfit$pass == FALSE)} else {paste("NA- no models generated by this fit")}`

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for nbinom1
if (exists("nb1_models") && length(nb1_models) > 0) {
  nb1_dharma_results <- process_dharma_results(nb1_models)

  kable(nb1_dharma_results$summary)
}

```

### Zero-inflated Negative Binomial Polynomial Models

```{r}
#| eval: true
#| label: nbinom2 models

nb2_models <- keep_at(
  nb2_EvsB$all_models,
  filter(bestDist_filtered, dist == "nb2")$outcome_id
)

```


```{r}
#| eval: true
#| label: nb2-summary
if (exists("nb2_models") && length(nb2_models) > 0) {
  nb2_unfit <- diagnoseZI(nb2_models)

  nb2_summary <- data.frame(
    outcome_id = names(nb2_models),
    continuous_distribution = "nbinom2",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
  nb2_ogu_models <- nb2_EvsB$adj_results %>%
    filter(outcome_id %in% names(nb2_models))
}
```

Are there any models that fail diagnostics: `r if (exists("nb2_models") && length(nb2_models) > 0) {any(nb2_unfit$pass == FALSE)} else {paste("NA- no models generated by this fit")}`

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for nbinom1
if (exists("nb2_models") && length(nb2_models) > 0) {
  nb2_dharma_results <- process_dharma_results(nb2_models)

  kable(nb2_dharma_results$summary)
}
```

### Zero-inflated Poisson Models

```{r}
#| eval: true
#| label: poi models

poi_models <- keep_at(
  poi_EvsB$all_models,
  filter(bestDist_filtered, dist == "poi")$outcome_id
)

```


```{r}
#| eval: true
#| label: poi-summary
if (exists("poi_models") && length(poi_models) > 0) {
  poi_unfit <- diagnoseZI(poi_models)

  poi_summary <- data.frame(
    outcome_id = names(poi_models),
    continuous_distribution = "poisson",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
  poi_ogu_models <- poi_EvsB$adj_results %>%
    filter(outcome_id %in% names(poi_models))
}
```

Are there any models that fail diagnostics: `r if (exists("poi_models") && length(poi_models) > 0) {any(poi_unfit$pass == FALSE)} else {paste("NA- no models generated by this fit")}`

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for Poisson models
if (exists("poi_models") && length(poi_models) > 0) {
  poi_dharma_results <- process_dharma_results(poi_models)

  kable(poi_dharma_results$summary)
}
```

### Zero-inflated Gaussian Models

```{r}
#| eval: true
#| label: gau unfit models
gau_models <- keep_at(
  gau_EvsB$all_models,
  filter(bestDist_filtered, dist == "gau")$outcome_id
)

```


```{r}
#| eval: true
#| label: gau-summary
if (exists("gau_models") && length(gau_models) > 0) {
  gau_unfit <- diagnoseZI(gau_models)

  gau_summary <- data.frame(
    outcome_id = names(gau_models),
    continuous_distribution = "gaussian",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
  gau_ogu_models <- gau_EvsB$adj_results %>%
    filter(outcome_id %in% names(gau_models))
}
```

Are there any models that fail diagnostics: `r if (exists("gau_models") && length(gau_models) > 0) {any(gau_unfit$pass == FALSE)} else {paste("NA- no models generated by this fit")}`

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for Poisson models
if (exists("gau_models") && length(gau_models) > 0) {
  gau_dharma_results <- process_dharma_results(gau_models)

  kable(gau_dharma_results$summary)
}
```

### Zero-inflated Lognormal Models

```{r}
#| eval: true
#| label: lognorm unfit models

lognorm_models <- keep_at(
  lognorm_EvsB$all_models,
  filter(bestDist_filtered, dist == "lognorm")$outcome_id
)

```

```{r}
#| eval: true
#| label: lognorm-summary
if (exists("lognorm_models") && length(lognorm_models) > 0) {
  lognorm_unfit <- diagnoseZI(gau_models)

  lognorm_summary <- data.frame(
    outcome_id = names(lognorm_models),
    continuous_distribution = "Log normal",
    continuous_link = "log",
    covaraince_structure = "unstructured",
    zero_distribution = "binary",
    zero_link = "logit",
    optimizer = "nlminb"
  )
  lognorm_ogu_models <- lognorm_EvsB$adj_results %>%
    filter(outcome_id %in% names(lognorm_models))
}
```

```{r}
#| eval: true
#| label: dharma-results
#| tbl-cap: Summary of Dharma residual tests for Poisson models
if (exists("lognorm_models") && length(lognorm_models) > 0) {
  lognorm_dharma_results <- process_dharma_results(lognorm_models)

  kable(lognorm_dharma_results$summary)
}
```

Are there any models that fail diagnostics: `r if (exists("lognorm_models") && length(lognorm_models) > 0) {any(lognorm_unfit$pass == FALSE)} else {paste("NA- no models generated by this fit")}`

### Save Final Model Fits

```{r}
#| eval: true
#| label: save
#| echo: false

save(
  nb1_models,
  nb2_models,
  poi_models,
  gau_models,
  lognorm_models,
  file = file.path(filepath, paste0(prefix, "finalised-fits.RData"))
)
```

```{r}
#| eval: false
#| label: load
#| echo: false

load(file.path(filepath, paste0(prefix, "finalised-fits.RData")))

```

## Significant OGUs

```{r}
#| eval: true
#| label: model-ogus
nb1_ogu_models <- nb1_EvsB$adj_results %>%
  filter(outcome_id %in% names(nb1_models))

nb2_ogu_models <- nb2_EvsB$adj_results %>%
  filter(outcome_id %in% names(nb2_models))

poi_ogu_models <- poi_EvsB$adj_results %>%
  filter(outcome_id %in% names(poi_models))
```

```{r}
#| eval: true
#| label: tbl-summary
#| tbl-cap: Summary table of OGUs successfully fit by a model, and their corresponding coefficient p-values

poi_summary <- left_join(poi_summary, poi_ogu_models)
nb1_summary <- left_join(nb1_summary, nb1_ogu_models)
nb2_summary <- left_join(nb2_summary, nb2_ogu_models)

kable(head(
  summary <- rbind(
    nb1_summary,
    nb2_summary,
    poi_summary
  )
))
```

```{r}
#| eval: true
#| label: save-tbl-summary
#| echo: false
#| results: asis
#| cache: false
tmp <- summary %>%
  left_join(rownames_to_column(data$taxa_data, "outcome_id"), by = "outcome_id")

save_and_link(
  data = tmp,
  filename = paste0(prefix, "zero-inflated-ogu-fitted-model-pvalues.csv"),
  description = "Summary table of OGUs successfully fit by a model",
  directory_naming = "both"
)

```

Of the `r ncol(outcome)` OGUs that passed filtering (`r ncol(outcome)/ncol(data$ogu_data)`), `r nrow(summary)` were sucessfully fit to a zero-inflated model (`r nrow(summary)/ncol(outcome)*100`%)

```{r}
#| eval: true
#| label: zi-sig

ziSig <- identify_significant_ogus(
  summary,
  data$taxa_data
)

colnames(ziSig)[colnames(ziSig) == "rowname"] <- "outcome_id"
```

```{r}
#| eval: true
#| results: asis
#| cache: false

save_and_link(
  data = ziSig,
  filename = paste0(
    prefix,
    "zero-inflated-ogu-fitted-model-significant-pvalues.csv"
  ),
  description = "Significant OGUs based on ZI-GLMMs",
  directory_naming = "both"
)
```

## Zero-inflated component

```{r}
#| eval: true
#| label: zi-comp
zi <- lapply(
  list(nb1_models, nb2_models, poi_models, gau_models),
  function(l) {
    lapply(l, function(x) {
      summary(x)$coefficients$zi[, 4]
    })
  }
)

#FIX make naming automatic using indexing
zip <- bind_rows(
  lapply(zi, function(model_type) {
    bind_rows(model_type, .id = "outcome_id")
  }),
  .id = "model_type"
) %>%
  # Add better names for model types
  mutate(
    model_type = case_when(
      model_type == "1" ~ "nb1",
      model_type == "2" ~ "nb2",
      model_type == "3" ~ "poi",
      model_type == "4" ~ "gau",
      TRUE ~ model_type
    )
  )

if (any(zip <= 0.05)) {
  zip_sig <- identify_significant_ogus(zip, data$taxa_data)
}
```

```{r}
#| eval: true
#| label: save-zip-sig
#| results: asis
#| cache: false
save_and_link(
  data = zip_sig,
  filename = paste0(
    prefix,
    "significant-ogus-zero-inflated-model-ZI-component.csv"
  ),
  description = "Significant OGUs based on ZI-GLMMs (zero-inflated compponent)",
  directory_naming = "both"
)

```

## Using EMMEANS to determine significant changes in taxa abundances

```{r fits}
#| eval: true
#| label: fit-mods

fits <- list(nb1_models, nb2_models, poi_models)

fits <- list_c(fits)
```

```{r}
#| eval: true
#| label: save-fits
#| results: asis

save_and_link(
  data = fits,
  filename = paste0(
    prefix,
    "zero-inflated-ogu-fitted-model-fits-variable.rds"
  ),
  description = "RDS file for `fits` variable",
  directory_naming = "both"
)
```

### Conditional Model EMMs

#### Overall OGUs Estimated marginal means

The below estimates the mean for each `timepoint`, age at baseline at a constant

| Scaled value | Age in days | Age    |
|--------------|-------------|--------|
| 0.0539  | 365    | 1yr (12mths)   |
| 0.454   | 456    | 1.25yrs (15mths) |
| 0.859   | 548    | 1.5yrs (18mths) |

```{r emm}
#| eval: true
#| label: tbl-within-time
#| tbl-cap: Estimated marginal means for OGU abundances within each timepoint

emm <- lapply(fits, function(x) {
  emm <- emmeans(
    x,
    ~ age_B | time,
    at = list(time = c(0, 1), age_B = c(0.0529, 0.454, 0.859)),
    weights = "proportional",
    type = "scale"
  )

  levels(emm)$time <- c("Baseline", "Endpoint")
  levels(emm)$age_B <- c(
    "12mths at Baseline",
    "15mths at Baseline",
    "18mths at Baseline"
  )

  return(emm)
})
```

#### OGU EMM differences between baseline ages at baseline and endpoint

```{r}
#| eval: true
#| label: tbl-within-time-cont
#| tbl-cap: Comparison of abundance marginal mean estimates (EMMs) betweeen intervention arm at baseline and endpoint. Only those with p $\leq$ 0.05 are show for brevity.

emm_pairs <- lapply(emm, function(x) {
  contrast(
    x,
    by = c("time"),
    method = "revpairwise",
    type = "scale",
    weights = "proportional",
    adjust = "fdr"
  )
})

emm_pairs_df <- emm_bind(emm_pairs, kable = F)

kable(head(sigAgeB <- emm_pairs_df %>%
  filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
  mutate(
    change = ifelse(estimate < 0, "Decrease", "Increase"),
    timepoint = time
  )))
```

##### Effect Size

The default value used to standardise the effect size is the model's residual standard deviation (sigma; within-group variation). However, this may underestimate effect sizes in mixed models where combined variance may be more accurate. Here, the random effect variance is used to standardise effect sizes.

```{r}
#| eval: true
#| label: sigArm-eff
#| tbl-cap: Effect sizes calculated as a standardized mean difference similar to Cohen's d.

emmEff <- lapply(1:length(emm_pairs), function(i) {
  eff_size(
    emm_pairs[[i]],
    sigma = attr(VarCorr(fits[[i]])$cond$subject, "stddev") * sqrt(2),
    edf = df.residual(fits[[i]]),
    method = "identity",
    parens = FALSE
  )
})

names(emmEff) <- names(emm_pairs)

emmEff_df <- emm_bind(emmEff, kable = F)

emmEff_df <- emmEff_df %>%
  filter(Model %in% sigAgeB$Model) %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    )
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, contrast, time, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

sigAgeBeff <- left_join(sigAgeB, emmEff_df, by = c("Model", "contrast", "time"))

mth12v15 <- sigAgeBeff %>%
  filter(str_detect(contrast, "12mths")) %>%
  filter(str_detect(contrast, "15mths"))

mth12v18 <- sigAgeBeff %>%
  filter(str_detect(contrast, "12mths")) %>%
  filter(str_detect(contrast, "18mths"))

```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between infants who started intervention at 12 months vs those who were 15 months old at baseline and endpoint

plotPhylaEMMs(mth12v15) +
  facet_wrap(~timepoint, scales = "free")
```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between infants who started intervention at 12 months vs those who were 18 months old at baseline and endpoint

plotPhylaEMMs(mth12v18) +
  facet_wrap(~timepoint, scales = "free")
```

#### Rates of OGU abundace changes over time with respect to age at baseline

```{r}
#| eval: true
#| label: tbl-emm-arm-trend
#| tbl-cap: Rate of change per unit increase of abundance marginal mean estimates over time. The time.trend column represents the slope of the relationship with time, while age at baseline is held constant
trend <- lapply(fits, function(x) {
  emm <- emtrends(
    x,
    ~age_B,
    var = "time",
    at = list(age_B = c(0.0529, 0.454, 0.859)),
    adjust = "fdr"
  )

  levels(emm)$age_B <- c(
    "12mths at Baseline",
    "15mths at Baseline",
    "18mths at Baseline"
  )

  return(emm)
})
```

```{r}
#| eval: true
#| label: tbl-emm-arm-trend-compare
#| tbl-cap: Comparison of the rates of change from baseline to endpoint in abundance EMMs between intervention arms.

slopes <- lapply(trend, function(x) {
  contrast(
    x,
    method = "revpairwise",
    type = "scale",
    weights = "proportional",
    adjust = "fdr"
  )
})

slopes_df <- emm_bind(slopes, kable = F)

sigArmsTime <- slopes_df %>%
  filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model")
```

##### Effect Size

The default value used to standardise the effect size is the model's residual standard deviation (sigma; within-group variation). However, this may underestimate effect sizes in mixed models where combined variance may be more accurate. Here, the random effect variance is used to standardise effect sizes.

```{r}
#| eval: true
#| label: sigArm-eff
#| fig-cap: NA
#| tbl-cap: Effect sizes calculated as a standardized mean difference similar to Cohen's d.

emmEff <- lapply(1:length(slopes), function(i) {
  eff_size(
    slopes[[i]],
    sigma = attr(VarCorr(fits[[i]])$cond$subject, "stddev") * sqrt(2),
    edf = df.residual(fits[[i]]),
    method = "identity",
    parens = FALSE
  )
})

names(emmEff) <- names(slopes)

emmEff_df <- emm_bind(emmEff, kable = F)

emmEff_df <- emmEff_df %>%
  dplyr::filter(Model %in% sigArmsTime$Model) %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    )
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, contrast, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

sigArmsTimeEff <- left_join(sigArmsTime, emmEff_df, by = c("Model", "contrast"))

mth12v15 <- sigArmsTimeEff %>%
  dplyr::filter(str_detect(contrast, "12mths")) %>%
  dplyr::filter(str_detect(contrast, "15mths"))

mth12v18 <- sigArmsTimeEff %>%
  dplyr::filter(str_detect(contrast, "12mths")) %>%
  dplyr::filter(str_detect(contrast, "18mths"))

```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between infants who started intervention at 12 months vs those who were 15 months old at baseline and endpoint

plotPhylaEMMs(dplyr::filter(mth12v15, abs(effect.size) >= 0.5), highlight_common = FALSE)
```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between infants who started intervention at 12 months vs those who were 18 months old at baseline and endpoint

plotPhylaEMMs(dplyr::filter(mth12v18, abs(effect.size) >= 0.5), highlight_common = FALSE)
```

#### OGU EMM changes over time (ignoring baseline age)

```{r emm}
#| eval: true
#| label: Changes in estimated marginal means for OGU abundances
#| warning: false
#| message: false

emm_time <- lapply(fits, function(x) {
  # Get EMMs for arm at each time point
  emm <- emmeans(
    x,
    ~time,
    at = list(time = c(0, 1)),
    weights = "proportional",
    type = "scale"
  )
  levels(emm)$time <- c("Baseline", "Endpoint")
  return(emm)
})

```

```{r}
#| eval: true
#| label: tbl-emm-time
#| tbl-cap: Changes in abundance marginal mean estimates over time.

time_cont <- lapply(emm_time, pairs, reverse = TRUE)

time_cont_df <- emm_bind(time_cont, kable = F) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
  mutate(Species = ifelse(Species == "", Genus, Species))

kable(head(sigTime <- time_cont_df %>%
  filter(p.value <= 0.05)
))
```

Of `r nrow(time_cont_df)` OGUs sucessfully fitted with ZI-GLMMs, `r nrow(sigTime)` OGUs had FDR \u2264 0.05

##### Effect Size

The default value used to standardise the effect size is the model's residual standard deviation (sigma; within-group variation). However, this may underestimate effect sizes in mixed models where combined variance may be more accurate. Here, the random effect variance is used to standardise effect sizes.

```{r}
#| eval: true
#| label: sigArm-eff
#| fig-cap: NA
#| tbl-cap: Effect sizes calculated as a standardized mean difference similar to Cohen's d.

emmEff <- lapply(1:length(time_cont), function(i) {
  eff_size(
    time_cont[[i]],
    sigma = attr(VarCorr(fits[[i]])$cond$subject, "stddev") * sqrt(2),
    edf = df.residual(fits[[i]]),
    paren = FALSE,
    method = "identity"
  )
})

names(emmEff) <- names(time_cont)

emmEff_df <- emm_bind(emmEff, kable = F)

emmEff_df <- emmEff_df %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    )
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

sigTimeEff <- left_join(
  sigTime,
  filter(emmEff_df, Model %in% sigTime$Model),
  by = c("Model")
)

kable(head(time_cont_df <- time_cont_df %>%
  left_join(emmEff_df, by = "Model")))
```

Of `r nrow(time_cont_df)` OGUs sucessfully fitted with ZI-GLMMs: 

- `r nrow(time_cont_df %>% filter(magnitude == "medium" | magnitude == "large"))` had an effect size of medium or more
- `r nrow(time_cont_df %>% filter(p.value <= 0.05)` OGUs had FDR \u2264 0.05
- `r nrow(time_cont_df %>% filter(p.value <= 0.05, magnitude == "medium" | magnitude == "large"))` OGUs had FDR \u2264 0.05 and an effect size of medium or more

```{r}
#| eval: true
#| label: tbl-eff-nums
#| tbl-cap: Number of significant OGUs (FDR \u2264 0.05) with particular effect sizes

kable(time_cont_df %>%
  filter(p.value <= 0.05) %>%
  mutate(direction = ifelse(effect.size < 0, "Decrease", "Increase")) %>%
  group_by(magnitude, direction) %>%
  mutate(count = 1) %>%
  summarise(sum = sum(count)))
```

```{r}
#| eval: true
#| label: save-sigTime
#| results: asis
#| cache: false

save_and_link(
  data = time_cont_df,
  filename = paste0(prefix, "ogu-emms-time-ZI.csv"),
  description = "OGU EMMs differences between timepoint",
  directory_naming = "both"
)

save_and_link(
  data = sigTimeEff,
  filename = paste0(prefix, "significant-ogu-emms-time-ZI.csv"),
  description = "OGU EMMs significantly different at endpoint and their effect sizes",
  directory_naming = "both"
)
```

```{r}
#| eval: true
#| label: fig-phyla-emms-effsize
#| fig-cap: Estimated marginal mean differences between baseline and endpoint. Significant differences with an effect size > 0.5 (medium effect size) are shown in the plot.

plotPhylaEMMs(
  filter(sigTimeEff, abs(effect.size) >= 0.5),
  text_size = 8,
  highlight_common = FALSE
)
```

### Zero-inflated Componet EMMs

#### Overall OGUs Estimated marginal means for Zero Inflated Model Component

```{r}
#| eval: true
#| label: tbl-emms
#| tbl-cap: Zero-inflated estimated marginal means for OGU abundances within each timepoint

emm_zi <- lapply(fits, function(x) {
  # Get EMMs for arm at each time point
  emm <- emmeans(
    x,
    ~ age_B | time,
    at = list(time = c(0, 1), age_B = c(0.0529, 0.454, 0.859)),
    weights = "proportional",
    type = "scale",
    component = "zi",
    data = predictors
  )
  levels(emm)$time <- c("Baseline", "Endpoint")
  levels(emm)$age_B <- c(
    "12mths at Baseline",
    "15mths at Baseline",
    "18mths at Baseline"
  )

  return(emm)
})

# Convert to probability scale
zi_probs <- lapply(emm_zi, regrid, transform = "logit")
```

#### OGU EMM differences between baseline ages at baseline and endpoint

```{r}
#| eval: true
#| label: tbl-emm-pairs
#| tbl-cap: Comparison of abundance marginal mean estimates (EMMs) betweeen intervention arm at baseline and endpoint

emm_zi_pairs <- lapply(emm_zi, function(x) {
  contrast(
    x,
    by = c("time"),
    method = "revpairwise",
    type = "scale",
    weights = "proportional",
    adjust = "fdr"
  )
})

emm_zi_pairs_df <- emm_bind(emm_zi_pairs, kable = F)

ZIsigArms <- emm_zi_pairs_df %>%
  dplyr::filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model") %>%
  mutate(
    change = ifelse(estimate < 0, "Decrease", "Increase"),
    timepoint = time
  )
```

##### Effect Size

> [!quote]
> When a logistic regression is calculated, the regression coefficient (b1) is the estimated increase in the log odds of the outcome per unit increase in the value of the exposure. In other words, the exponential function of the regression coefficient (eb1) is the odds ratio associated with a one-unit increase in the exposure. [@Szumilas2010]

i.e. if $\beta_x = log(odds)$, then the odds ratio is $OR = \exp(\beta_x) = e^{\beta_x}$

The "nonexposure" group in this context would be the reference age group of 12mths

```{r}
#| eval: false
#| label: zi-odds
#| tbl-cap: Effect size calculated as odds-ratio and converted to Cohen's d with respect to the rate of the desired outcome for zero-inflation component

oddsEff <- Map(summarize_zi_contrasts, emm_zi, emm_zi_pairs, fits)

oddsEff <- data.table::rbindlist(oddsEff, idcol = "Model") %>%
  select(
    Model,
    contrast,
    time,
    OR,
    OR_lower,
    OR_upper,
    P0,
    cohens_d,
    magnitude
  )

kable(head(oddsEff))
```

```{r}
#| eval: false
#| label: cohens-d
#| fig-cap: NA
#| tbl-cap: NA
ZIemmEff <- lapply(1:length(emm_zi_pairs), function(i) {
  eff_size(
    emm_zi_pairs[[i]],
    sigma = attr(VarCorr(fits[[i]])$cond$subject, "stddev") * sqrt(2),
    edf = df.residual(fits[[i]]),
    method = "identity",
    parens = FALSE,
    ratios = TRUE
  )
})

names(ZIemmEff) <- names(emm_zi_pairs)

ZIemmEff <- emm_bind(ZIemmEff, kable = F)

ZIemmEff <- ZIemmEff %>%
  dplyr::filter(Model %in% ZIsigArms$Model) %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    )
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, contrast, time, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

ZIsigArmsEff <- left_join(
  ZIsigArms,
  ZIemmEff,
  by = c("Model", "time", "contrast")
)

kable(
  ZIsigArmsEff <- left_join(
    ZIsigArmsEff,
    oddsEff,
    by = c("Model", "time", "contrast")
  ) %>%
    select(!c(Kingdom, Class, Order, Family, time))
)

if (!all(ZIsigArmsEff$magnitude == ZIsigArmsEff$OR_magnitude)) {
  echo("Warning: Effect size and odds ratio magnitudes do not match")
}

mth12v15 <- ZIsigArmsEff %>%
  dplyr::filter(str_detect(contrast, "12mths")) %>%
  dplyr::filter(str_detect(contrast, "15mths"))

mth12v18 <- ZIsigArmsEff %>%
  dplyr::filter(str_detect(contrast, "12mths")) %>%
  dplyr::filter(str_detect(contrast, "18mths"))
```

```{r}
#| eval: false
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between infants who started intervention at 12 months vs those who were 15 months old at baseline and endpoint

plotPhylaEMMs(mth12v15, highlight_common = FALSE)
```

```{r}
#| eval: false
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between infants who started intervention at 12 months vs those who were 18 months old at baseline and endpoint

plotPhylaEMMs(mth12v18, highlight_common = FALSE)
```

#### Rates of OGU abundace changes over time with respect to age at baseline

```{r}
#| eval: true
#| label: tbl-emm-arm-trend
#| tbl-cap: Rate of change per unit increase of abundance marginal mean estimates over time. The time.trend column represents the slope of the relationship with time.

trend_zi <- lapply(fits, function(x) {
  emm <- emtrends(
    x,
    ~ age_B,
    var = "time",
    at = list(age_B = c(0.0529, 0.454, 0.859)),
    adjust = "fdr",
    weights = "proportional",
    type = "scale",
    data = predictors
  )

  levels(emm)$age_B <- c(
    "12mths at Baseline",
    "15mths at Baseline",
    "18mths at Baseline"
  )

  return(emm)
})
```

```{r}
#| eval: true
#| label: tbl-emm-arm-trend-compare
#| tbl-cap: Comparison of the rates of change from baseline to endpoint in abundance EMMs between intervention arms.

slopes_zi <- lapply(trend_zi, function(x) {
  contrast(
    x,
    method = "revpairwise",
    type = "scale",
    weights = "proportional",
    adjust = "fdr"
  )
})

slopes_zi_df <- emm_bind(slopes_zi, kable = F)

ZIsigArmsTime <- slopes_zi_df %>%
  dplyr::filter(p.value <= 0.05) %>%
  left_join(rownames_to_column(data$taxa_data, "Model"), by = "Model")
```

##### Effect Size

The default value used to standardise the effect size is the model's residual standard deviation (sigma; within-group variation). However, this may underestimate effect sizes in mixed models where combined variance may be more accurate. Here, the random effect variance is used to standardise effect sizes.

```{r}
#| eval: true
#| label: sigArm-eff
#| fig-cap: NA
#| tbl-cap: Effect sizes calculated as a standardized mean difference similar to Cohen's d.

emmEff_zi <- lapply(1:length(slopes_zi), function(i) {
  eff_size(
    slopes_zi[[i]],
    sigma = attr(VarCorr(fits[[i]])$cond$subject, "stddev") * sqrt(2),
    edf = df.residual(fits[[i]]),
    method = "identity",
    parens = FALSE
  )
})

names(emmEff_zi) <- names(slopes_zi)

emmEff_zi_df <- emm_bind(emmEff_zi, kable = F)

emmEff_zi_df <- emmEff_zi_df %>%
  dplyr::filter(Model %in% ZIsigArmsTime$Model) %>%
  mutate(
    magnitude = case_when(
      abs(effect.size) < 0.2 ~ "negligible",
      abs(effect.size) >= 0.2 & abs(effect.size) < 0.5 ~ "small",
      abs(effect.size) >= 0.5 & abs(effect.size) < 0.8 ~ "medium",
      abs(effect.size) >= 0.8 ~ "large",
    )
  ) %>%
  arrange(abs(effect.size)) %>%
  select(Model, contrast, effect.size, magnitude) %>%
  mutate_at(vars(everything()), type.convert, as.is = T)

ZIsigArmsTimeEff <- left_join(
  ZIsigArmsTime,
  emmEff_zi_df,
  by = c("Model", "contrast")
)

mth12v15 <- sigArmsTimeEff %>%
  dplyr::filter(str_detect(contrast, "12mths")) %>%
  dplyr::filter(str_detect(contrast, "15mths"))

mth12v18 <- sigArmsTimeEff %>%
  dplyr::filter(str_detect(contrast, "12mths")) %>%
  dplyr::filter(str_detect(contrast, "18mths"))

```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between infants who started intervention at 12 months vs those who were 15 months old at baseline and endpoint

plotPhylaEMMs(dplyr::filter(mth12v15, abs(effect.size) >= 0.5), highlight_common = FALSE)
```

```{r}
#| eval: true
#| label: fig-phyla-emms
#| fig-cap: Estimated marginal mean differences between infants who started intervention at 12 months vs those who were 18 months old at baseline and endpoint

plotPhylaEMMs(dplyr::filter(mth12v18, abs(effect.size) >= 0.5), highlight_common = FALSE)
```

## Results Summary

```{r}
#| eval: true
#| echo: false
#| results: asis
#| cache: false

list_all_results()
```

```{r}
#| results: asis
#| cache: false
generate_results_section(directory_naming = "both")
```

## Appendix

```{r}
#| eval: true
#| label: info

print(sessionInfo(), locale = FALSE)
```
